### 2022-10-29 paper/paper.tex
package builds.

\documentclass[11pt, fleqn]{article}
\usepackage[english]{babel}
\usepackage{amsmath}

\usepackage[linesnumbered]{algorithm2e}

\usepackage{natbib}
\setlength{\parskip}{\baselineskip}%
\setlength{\parindent}{0pt}%

\title{Spit-Apply-Combine with Collapsing Groups}
\author{Mark P.J. van der Loo}

\begin{document}
\maketitle

Let $U$ be a finite population, and let $A$ be a finite set with $|A|\leq |U|$.
Any function 
\begin{equation}
f: U\to A,
\end{equation}
defines a partition on $U$, in the sense that two elements are equivalent if they
map to the same element of $A$. In the rest of the discussion it will be useful to 
also define the formal inverse $f^{-1}:A\to 2^U$, such that $f^{-1}(a)$ is the set of 
elements in $U$ that maps to $a$. In notation:
\begin{equation}
f^{-1}(a) \equiv \{u\in U: f(u)=a\}.
\end{equation}

\begin{algorithm}[H]
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{$U$, $\phi: 2^U\to X$, a partition $(f,A)$}
\Output{The value of $\phi$ for every part of $U$ induced by $(f,A)$, represented
       as a set pairs $(a,x)\in A\times X$. }

$R = \{\}$\;
\While{$U\not=\{\}$}{
Choose any $u\in U$\;
$d = (f^{-1}\circ f) (u)$\;
$U = U - d$\;
$ R= R\cup (f(u), \phi(d)$)\;
}

\end{algorithm}

\bibliographystyle{plainnat}
\bibliography{vanderloo}
\end{document}

### 2022-10-29 pkg/R/collapse.R

@@ -1,0 +1,76 @@
+#' Split-Apply-Combine with Collapsing Groups
+#'
+#' Attempt to compute a grouped aggregate. If a group does not satisfy certain
+#' user-defined conditions (such as too many missings, or not enough records)
+#' then the group is expanded according to a user-defined 'collapsing' scheme.
+#' This happens recursively until either the group satisfies all conditions and
+#' the aggregate is computed, or we run out of collapsing possibilities and the 
+#' \code{NA} is returned for that group.
+#'
+#'
+#' @param data \code{[data.frame]} The data to aggregate by (collapsing) groups.
+#' @param collapse \code{[data.frame]} representing a group collapsing sequence (see details).
+#'        All columns must be of the same type.
+#' @param test \code{[function]} A function that takes a subset of \code{data} and returns
+#'             \code{TRUE} if it is suitable for computing the desired aggregates and 
+#'             \code{FALSE} if a collapsing step is necessary.
+#' @param ... A comma-separated list of \code{name=expression}, where \code{expression}
+#'            defines the aggregating operation.
+#'
+#' @return
+#' A data frame with \code{nrow(collapse)} rows. The first column equals the first column
+#' of collapse and indicates the group for which an aggregate was computed. 
+#' The second column indicates the group used to compute the aggregate. The subsequent
+#' columns contain the results of the expressins listed in the \code{...} argument.
+#'
+#' @examples
+#' # some example data
+#' dat <- data.frame(A0 = c("11","12","11","22"), Y = c(2,4,6,8))
+#' # collapsing scheme
+#' csh <- data.frame(
+#'    A0 = c("11","12","22")
+#'  , A1 = c("1" ,"1", "2") 
+#' )
+#' collapse(data = dat
+#'    , collapse = csh
+#'    , test     = function(d) if (nrow(d)<2) FALSE else TRUE
+#'    , mn = mean(Y, na.rm=TRUE)
+#'    , md = median(Y, na.rm=TRUE)
+#' )
+#'
+#' @export
+collapse <- function(data, collapse, test, ...){
+  exprs <- as.list(substitute(list(...))[-1])
+
+
+  collapse <- as.data.frame(lapply(collapse, as.character))
+  rownames(collapse) <- collapse[,1]
+
+  # set up output set
[…]

### 2022-10-29 pkg/inst/tinytest/test_collapse.R

@@ -1,0 +1,18 @@
+
+#source("../../R/rcollapse.R")
+
+d <- data.frame(A0 = as.character(c(11,11,12)), Y = c(4,8,6))
+A <- data.frame(A0 = as.character(c(11,12)), A1 = as.character(c(1,1)))
+
+out <- collapse(data = d
+  , collapse  = A
+  , test      = function(d) if (nrow(d)>=2) TRUE else FALSE
+  , mn = mean(Y, na.rm=TRUE)
+  , md = median(Y,na.rm=TRUE) )
+
+expect_equal(out[,1],c("11","12"))
+expect_equal(out[,2],c("11","1"))
+expect_equal(out[,3],c(6,6))
+expect_equal(out[,4],c(6,6))
+
+

### 2022-10-29 pkg/tests/tinytest.R

@@ -1,0 +1,5 @@
+
+if ( requireNamespace("tinytest", quietly=TRUE) ){
+  tinytest::test_package("rcollapse")
+}
+

### 2022-10-29 pkg/R/helpers.R
added some helper functions to facilitate specifying demands

@@ -1,0 +1,149 @@
+#' Demand minimal number of functons
+#' 
+#' @param n Minimal number of records in a group.
+#'
+#' @examples
+#'
+#' min_records(5)(women)
+#' min_records(200)(women)
+#'
+#' @family helpers
+#'
+#' @export
+min_records <- function(n){
+  min_rec <- n
+  function(d) if (nrow(d) < n) FALSE else TRUE
+}
+
+#' Demand maximal number of records
+#'
+#' @param n Minimal number of records in a group.
+#'
+#' @examples
+#'
+#' max_records(5)(women)
+#' max_records(200)(women)
+#'
+#'
+#' @family helpers
+#'
+#' @export
+max_records <- function(n){
+  max_rec <- n
+  function(d) if (nrow(d) >n) FALSE else TRUE
+}
+
+#' Demand any number of expressions to be TRUE
+#'
+#' @param ... comma-separated list of expressions that must all evaluate to \code{TRUE}
+#'            in order for a data set to be acceptable.
+#'
+#' @return A data-checking function.
+#' @family helpers
+#' @export
+demand <- function(...){
+  expr <- as.list(substitute(list(...))[-1])  
+  function(d){
+    L <- lapply(expr, function(e) with(d, eval(e)))
+    all(sapply(L,all))
+  }
+}
+
+#' Number of incomplete combinations
+#'
+#' A number of vectors is pointwise tested for missing values.  For each
+#' position, if any vector has a missing \code{FALSE} is computed otherwise
+#' \code{TRUE}. The total number of \code{FALSE}s is returned.
+#'
+#' @param ... comma-separated list of vectors of equal length
+#'
+#' @note This function is intended to facilitate the use of \code{\link{demand}}
+#' @return A non-negative integer.
+#'
+#'
+#' @examples
+#' n_incomplete(c(1,NA,3), 1:3)
+#'
+#' 
+#' @family helpers
+#'
+#' @export
+n_incomplete <- function(...){
+  L <- list(...)
+  sum( Reduce(`|`, lapply(L,  is.na)) )
+}
+
+#' Number of complete combinations
+#'
+#' A number of vectors is pointwise tested for missing values.  For each
+#' position, if any vector has a missing \code{FALSE} is computed otherwise
+#' \code{TRUE}. The total number of \code{TRUE}s is returned.
+#'
+#' @param ... comma-separated list of vectors of equal length
+#'
[…]

### 2022-10-29 pkg/inst/tinytest/test_helpers.R

@@ -1,0 +1,7 @@
+
+expect_equal(n_complete(c(1,NA,2), 1:3), 2)
+expect_equal(fraction_complete(c(1,NA,2), 1:3), 2/3)
+expect_equal(n_incomplete(c(1,NA,2), 1:3), 1)
+expect_equal(fraction_incomplete(c(1,NA,2), 1:3), 1/3)
+
+

### 2022-10-31 paper/paper.tex
work on paper

@@
 \usepackage[english]{babel}
-\usepackage{amsmath}
+\usepackage{amsmath, amssymb}
 
-\usepackage[linesnumbered]{algorithm2e}
+\usepackage[ruled, linesnumbered]{algorithm2e}
 
@@
 \author{Mark P.J. van der Loo}
+\date{CONCEPT \today{}}
 
@@
 
-Let $U$ be a finite population, and let $A$ be a finite set with $|A|\leq |U|$.
-Any function 
+\section{Introduction}
+The operation of splitting a data set into non-overlapping groups, computing an
+aggregate for each group, and combining the results into a new datasets is one
+of the most common operations in data analyses. The idea behind this so-called
+split-apply-combine approach is that the calculation for each groups only uses
+data available within the group. However, in some cases the quality of data in
+a group may be insufficient for producing a reliable group aggregate. Common
+quality issues include insufficient records, or too many records with values
+for certain variables missing. In such cases it is not uncommon to use data
+from outside the target group  to estimate the group aggregate. One example is
+Small Area Estimation, where the lack of observations in one group is
+compensated by using data from similar but other groups. Another area where
+such approaches occur is group-wise imputation. For example, one may wish to
+impute a group estimate (say, mean, or median) for missing values. If there are
+too little observations to provide an accurate estimate of the group mean, the
+group may be expanded, trading bias for variance.
+
+One approach that frequently occurs in practice is to use a so-called
+collapsing scheme. In this approach one attempts to estimate the aggregate for
+the most fine-grained grouping first.  If this is not possible, two or more
+groups are combined to provide input for estimating a value for the
+fine-grained group. The collapsing scheme defines which groups may be combined,
+and it may consist of multiple levels of collapsing; in principle to the point
+where the whole data set is used to estimate an aggregate value for a single
+group.
+
+Collapsing schemes can be the result of an optimisation, but they frequently
+result from subject matter as well. For example, in the case of hierarchical
+classifications, there is a natural ordering of groups that are similar from a
+subject matter point of view (whether that means that they can be considered
+representative for each other is another, statistical, question).  Examples
+include the NACE classification of economic activities or the ISCO
+classification of professions. Both have a tree-like structure that can be used
+to combine groups with a common ancestor in the hierarchy.
+
+In this paper we formalise the concept of `collapsing scheme', and give an
+algorithm for split-apply-combine operations based on it. Next, we demonstrate
+the R package \texttt{rcollapse} that implements this algorithm together with
+some utility functions that make working with collapsing schemes easier.
+
+\section{Formal descriptions and algorithms}
+In this section we give a formal description of the algorithms. In order to be
+unambiguous and to represent the algorithms compactly, we introduce some
+notation. The notation also makes the pseudocode independent of the
+implementation language used, or how the data is represented.
+
+We start by giving the algorithm for ordinary split-apply-combine, so it is
+easy to see where the algorithm must be generalized to allow for a collapsing
+scheme.
+
+\subsection{Split-apply-combine}
+In order to analyse a data set group by group we need to specify a data set, a
+way to split it into groups, and a function that takes a subset of data and
+returns an aggregate. Let us introduce some notation for that.
+
+Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
+$U$ and returns a single value in some domain $X$. Here, $U$ is our data set
+and $\phi$ an aggregating function. We split $U$ into groups using the
+following notation. Let $A$ be finite set that has no more elements than $U$,
+and let $f$ be a function that takes an element of $U$ and returns a value in
+$A$. We can think of $A$ as a set of group labels, and $f$ as the function that
+assigns a label to each element of $U$. This way, $f$ divides $U$ into
+non-overlapping groups, since every element in $U$ can only get one label (that
+is just how mathematical functions work). In formal terminology we say that
+$f:U\to A$ is a \emph{partition} of $U$.
+
+In this notation, any split-apply-combine operation can be computed with the following
+algorithm.
+
+\begin{algorithm}[H]
+\caption{Split-Apply-Combine}
+\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
+\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
+\Output{$R$: the value of $\phi$ for every part of $U$ as a set pairs $(a,x)\in
+A\times X$ }
+
+$R = \{\}$\;
+\For{$a\in A$}{
+  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
+  $R = R\cup \{(a,\phi(a))\}$\tcp*{aggregate and add to result}
+}
+\end{algorithm}
+Here, the output is collected in a set $R$. We also introduced some new
+notation: the notation $2^U$ stands for `the set of all subsets of $U$'. The
+notation $f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you
+give it a set of labels from $A$ and it returns all the elements of $U$ that
+have any of those labels. This definition is slightly too general for the
+current algorithm, but it will come in handy when we move to collapsing groups.
+
+Let us see how the elements $U$, $f\to A$ and $\phi$ are implemented
+in practice. Consider the signature of the R \texttt{aggregate} function for data
+frames (we skip arguments that are not important for the discussion).
+\begin{verbatim}
+  aggregate(x, by, FUN)
+\end{verbatim}
+Here, \texttt{x} is a data frame where each column is a variable to be
+aggregated.  The argument \texttt{by} represents the label(s) for each record
+that are to be used for grouping. So the function $f\to A$ is implemented by
+forcing the user to make sure that the position of each label in \texttt{by}
+corresponds to the row number in the data frame \texttt{x}. The argument
+\texttt{FUN} is the function that aggregates each subset of \texttt{x}.  Here is
+an example of how a user might use this function on the R prompt.
+%
+\begin{verbatim}
+> aggregate(iris["Sepal.Length"], by=iris["Species"], FUN=mean)
+     Species Sepal.Length
+1     setosa        5.006
+2 versicolor        5.936
+3  virginica        6.588
+\end{verbatim}
+%
+Note that the correspondence in position of the \texttt{Species} label and the
+record position is implemented by taking them from the same data frame.
+
+\subsection{Split-apply-combine with collapsing groups}
+
 \begin{equation}
-f: U\to A,
-\end{equation}
-defines a partition on $U$, in the sense that two elements are equivalent if they
-map to the same element of $A$. In the rest of the discussion it will be useful to 
-also define the formal inverse $f^{-1}:A\to 2^U$, such that $f^{-1}(a)$ is the set of 
-elements in $U$ that maps to $a$. In notation:
-\begin{equation}
-f^{-1}(a) \equiv \{u\in U: f(u)=a\}.
+U\xrightarrow{f_0}A_0\xrightarrow{f_1}\cdots\xrightarrow{f_n}A_n
+\label{eq:collapsingsequence}
 \end{equation}
 
+We also generalize the labeling function $f$ of the previous section
+to the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
+returns the corresponding label in $A_k$. This can formally be defined 
+as consecutive application of $f_1, f_2\cdots, f_k$ to an element of 
+$A_0$. In notation
+\begin{equation*}
+F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
+\end{equation*}
+Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
+\begin{equation*}
+F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
+\end{equation*}
+This function asks for a set of labels in $A_k$ and returns all the records in
+$U$ that are mapped to that label via the collapsing sequence of
+Equation~\eqref{eq:collapsingsequence}.
+
 \begin{algorithm}[H]
+\caption{Split-Apply-Combine with Collapsing Groups}
 \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
-\Input{$U$, $\phi: 2^U\to X$, a partition $(f,A)$}
-\Output{The value of $\phi$ for every part of $U$ induced by $(f,A)$, represented
-       as a set pairs $(a,x)\in A\times X$. }
+\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
+      and a collapsing sequence $U\xrightarrow{f_0}A_0\cdots \xrightarrow{f_n}  A_n$}
+
+\Output{$R$: the value of $\phi$ for every part of $U$, for which a suitable
+collapsing group can be found, as a set of triples $(a,k,x)\in A_0\times
+A_k\times X $ with $0\leq k\leq n$.}
 
 $R = \{\}$\;
-\While{$U\not=\{\}$}{
-Choose any $u\in U$\;
-$d = (f^{-1}\circ f) (u)$\;
-$U = U - d$\;
-$ R= R\cup (f(u), \phi(d)$)\;
+\For{$a\in A_0$}{
+  $i=0$\;
+  $d = f_0^{-1}(\{a\})$\tcp*{get subset of $U$}
+  \While{$i<n \land \lnot\beta(d)$}{
+    $i = i+1$ \tcp*{Increase collapse level} 
+    $d = (f_0^{-1}\circ F_i^{-1}\circ F_i)(a)$ \tcp*{Collapse and get subset}
+  }
+  \If{$\beta(d)$}{
+    $R = R\cup \{(a, i,\phi(d))\}$\;
+  }
 }
@@
 
+\subsection{Caveat: unbalanced hierarchical classifications}
+\begin{itemize}
+\item tree with different levels of details
+\end{itemize}
+
+\section{The \texttt{rcollapse} package}
+
 \bibliographystyle{plainnat}

### 2022-10-31 paper/paper.tex
few fixes, ideas, and added csh_from_digits

@@
 \texttt{FUN} is the function that aggregates each subset of \texttt{x}.  Here is
-an example of how a user might use this function on the R prompt.
+an example of how a user might call this function from the R prompt.
 %
@@
 Note that the correspondence in position of the \texttt{Species} label and the
-record position is implemented by taking them from the same data frame.
+record position is implemented by taking them from the same data frame. The output
+also reveals in the first column the set $A$: each row corresponds to a unique
+value in \texttt{Species} column.
 
@@
 
+\begin{itemize}
+\item{Specification of collapsing sequence with multiple collapsing variables (e.g. first size class, then NACE code)}
+\end{itemize}
+
 \bibliographystyle{plainnat}

### 2022-10-31 pkg/R/collapse.R

@@ -19,9 +19,13 @@
 #'
 #' @return
 #' A data frame with \code{nrow(collapse)} rows. The first column equals the first column
-#' of collapse and indicates the group for which an aggregate was computed. 
-#' The second column indicates the group used to compute the aggregate. The subsequent
+#' of \code{collapse} and indicates the group for which an aggregate was computed. 
+#' The second column indicates the number of collapses that took place. The subsequent
 #' columns contain the results of the expressins listed in the \code{...} argument.
+#' If no amount of collapsing yields a data set that passes the test, then for that
+#' row, the second and all subsequent columns are \code{NA}.
+#' 
+#'
 #'
 #' @examples
 #' # some example data
@@ -48,7 +52,7 @@ collapse <- function(data, collapse, test, ...){
 
   # set up output set
   out <- data.frame(collapse[,1], NA, row.names=collapse[,1])
-  colnames(out) <- c(names(collapse)[1], "input")
+  colnames(out) <- c(names(collapse)[1], "collapse_level")
   vals <- as.data.frame(matrix(NA, nrow=nrow(collapse), ncol=length(exprs)))
   colnames(vals) <- names(exprs)
   out <- cbind(out, vals)
@@ -65,7 +69,8 @@ collapse <- function(data, collapse, test, ...){
       grps <- collapse[collapse[,j] == collapse[a, j],1]
       d <- data[data[,g] %in% grps, ]
     }
-    out[a,2] <- collapse[a,j]
+    if (!test(d)) next
+    out[a,2] <- j-1
     out[a,3:ncol(out)] <- lapply(exprs, function(e) with(d, eval(e)))
   }
   rownames(out) <- NULL

### 2022-10-31 pkg/R/helpers.R

@@ -147,3 +147,28 @@ fraction_complete <- function(...){
 }
 
 
+#' Derive collapsing scheme from hierarchical classification
+#'
+#' Derive a collapsing scheme where group labels collapse to their
+#' parents in the hierarchy.
+#'
+#' @param x \code{[character|integer]} labels in a hierarchical classification (lowest level)
+#' @param pad \code{[character]} padding character used for codes shorter than the maximum length.
+#'
+#' 
+#' @export
+csh_from_digits <- function(x,pad="0"){
+  x <- as.character(x)
+  nlevels <- max(nchar(x))
+
+  A <- matrix(NA_character_,nrow=length(x), ncol=nlevels)
+  for ( i in seq_len(nlevels)){
+    A[,i] <- substr(x,1,nlevels+1-i)
+  }
+  rownames(A) <- x
+  colnames(A) <- sprintf("A%d",seq_len(nlevels)-1)
+  A
+}
+
+
+

### 2022-10-31 pkg/inst/tinytest/test_collapse.R

@@ -11,7 +11,7 @@ out <- collapse(data = d
   , md = median(Y,na.rm=TRUE) )
 
 expect_equal(out[,1],c("11","12"))
-expect_equal(out[,2],c("11","1"))
+expect_equal(out[,2],c(0,1))
 expect_equal(out[,3],c(6,6))
 expect_equal(out[,4],c(6,6))
 

### 2022-11-01 paper/paper.tex
paper updates

@@
 \caption{Split-Apply-Combine}
+\label{alg:sac}
 \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
@@
 \subsection{Split-apply-combine with collapsing groups}
-
+The goal of the algorithm is to compute a value for each 
+subset of a dataset, possibly using values from a larger
+subset. The input of the algorithm consists again of a
+finite set $U$, and aggregation function $\phi$ that takes
+a subset of $U$ and returns a value in some domain $X$.
+Compared to Algorithm~\ref{alg:sac} we need two extra
+inputs. First, we need a user-defined function that checks whether
+a given subset $d$ of $U$ is suitable for computing $\phi(d)$.
+We will call this function $\beta: 2^U\to \mathbb{B}$, where
+$\mathbb{B}=\{\texttt{True},\texttt{False}\}$. 
+Typical tests are checking whether there are sufficient records
+available, or whether certain variables have a low enough fraction
+of missing values. Second, we need
+a \emph{collapsing scheme}
 \begin{equation}
-U\xrightarrow{f_0}A_0\xrightarrow{f_1}\cdots\xrightarrow{f_n}A_n
+U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
 \label{eq:collapsingsequence}
 \end{equation}
+A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
+partitions $A_{i-1}$ into $|A_i|$ groups.
 
-We also generalize the labeling function $f$ of the previous section
-to the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
+Collapsing schemes can be represented as tables. For example, the European NACE
+classification is a hierarchical classification of economic activity. Each
+group is encoded as a 4-digit code where the first digit indicates the highest
+grouping level, and the last digit the most detailed level. This induces a
+collapsing sequence by combining groups that have the same parent in the
+hierarchy. Below is an example, based on a small piece of the NACE code list.
+%
+\begin{center}
+\begin{tabular}{rrr}
+\hline
+ $A_0$& $A_1$ & $A_2$\\
+\hline
+0111&   011 & 01\\
+0112&   011 & 01\\
+0113&   011 & 01\\
+0121&   012 & 01\\
+0122&   012 & 01\\
+0123&   012 & 01\\
+0124&   012 & 01\\
+\hline
+\end{tabular}
+\end{center}
+Here, $A_0$ is the most detailed partition, splitting a dataset into seven
+groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
+The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
+$0113$ and the second, labeled $01.2$ is constructed by combining
+$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.
+
+To compute an aggregate for records labeled with $0111$, the algorithm will
+select those records and test whether the subset is suitable. If not, it will
+move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
+follows.
+\begin{displaymath}
+f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
+\end{displaymath}
+It will then move on and find all records labeled with any of those three
+group labels, and test again. If this new data set is suitable, the result is
+computed, if not, another collapse takes place. The whole procedure is repeated 
+for each element in $A_0$.
+
+To generalize this algorithm to generic collapsing schemes, we introduce
+the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
 returns the corresponding label in $A_k$. This can formally be defined 
@@
 \end{equation*}
-This function asks for a set of labels in $A_k$ and returns all the records in
-$U$ that are mapped to that label via the collapsing sequence of
-Equation~\eqref{eq:collapsingsequence}.
+This function asks for a set of labels in $A_k$ and returns all the labels in
+$A_0$ that are mapped to that label via the collapsing sequence of
+Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally
+write down our main Algorithm~\ref{alg:saccg}.
 
+%
 \begin{algorithm}[H]
 \caption{Split-Apply-Combine with Collapsing Groups}
+\label{alg:saccg}
 \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
@@
   }
-  \If{$\beta(d)$}{
+  \If{$\beta(d)$}{ \label{line:cond}
     $R = R\cup \{(a, i,\phi(d))\}$\;
@@
 \end{algorithm}
+%
+Since the used level of collapsing is determined dynamically by data
+circumstances in $U$, the algorithm also reports the collapsing level used to
+compute a value foe each member of $A_0$. For each member of $A_0$ there is a
+triple $(a,i,\phi(d))$, where $a\in A_0$ is the label to which the value
+$\phi(d)$ pertains, and $i$ is the number of collapses applied to reach a
+suitable dataset. The condition in Line~\ref{line:cond} ensures that if
+no suitable dataset is found after the whole collapsing sequence has been
+executed, then no answer is returned.
 

### 2022-11-02 paper/paper.tex
some ideas

@@
 \begin{itemize}
-\item{Specification of collapsing sequence with multiple collapsing variables (e.g. first size class, then NACE code)}
+\item{Specification of collapsing sequence with multiple collapsing variables (e.g. first size class, then NACE code). Idea: \texttt{collapse(data, by, formula, FUN,...)} where \texttt{formula} is
+of the form \texttt{A0*B0 \~. A0*B1 + A1*B1 + A2*B1}. lhs is target level, rhs is collapsing sequence. }
+\item{Specification of collapsing sequence with different aggregators per collapsing step.}
+\item{Record-wise results}
+\item{Record-wise imputation}
+\item{Representation of hierarchical classifications: parent-child lists vs tabular}
 \end{itemize}

### 2022-11-15 pkg/R/collapse.R
some work

@@ -1,29 +1,68 @@
-#' Split-Apply-Combine with Collapsing Groups
+#' Split-Apply-Combine with Collapsing Groups: Multiple Aggregators
 #'
-#' Attempt to compute a grouped aggregate. If a group does not satisfy certain
-#' user-defined conditions (such as too many missings, or not enough records)
-#' then the group is expanded according to a user-defined 'collapsing' scheme.
-#' This happens recursively until either the group satisfies all conditions and
-#' the aggregate is computed, or we run out of collapsing possibilities and the 
-#' \code{NA} is returned for that group.
+#' Attempt to compute one or more grouped aggregates. If a group does not
+#' satisfy certain user-defined conditions (such as too many missings, or not
+#' enough records) then the group is expanded according to a user-defined
+#' 'collapsing' scheme.  This happens recursively until either the group
+#' satisfies all conditions and the aggregate is computed, or we run out of
+#' collapsing possibilities and the \code{NA} is returned for that group.
 #'
 #'
 #' @param data \code{[data.frame]} The data to aggregate by (collapsing) groups.
-#' @param collapse \code{[data.frame]} representing a group collapsing sequence (see details).
-#'        All columns must be of the same type.
+#' @param collapse \code{[formula|data.frame]} representing a group collapsing sequence.
+#'        See below for details on how to specify each option.
 #' @param test \code{[function]} A function that takes a subset of \code{data} and returns
 #'             \code{TRUE} if it is suitable for computing the desired aggregates and 
 #'             \code{FALSE} if a collapsing step is necessary.
 #' @param ... A comma-separated list of \code{name=expression}, where \code{expression}
 #'            defines the aggregating operation.
 #'
+#' @section Using a formula to define the collapsing sequence:
+#'
+#' If all combinations of collapsing options are stored as columns in
+#' \code{data}, the \code{formula} interface can be used. An example is the
+#' easiest way to see how it works is to illustrate the idea with an example.
+#' Suppose that \code{collapse = A*B ~ A1*B + B} This means:
+#' \itemize{
+#'  \item{Compute output for groups defined by variables A and B}
[…]

### 2022-11-18 paper/paper.tex
some work
updates and improvements

--- 2022-11-01 paper/paper.tex
@@
 \usepackage{amsmath, amssymb}
+
+\usepackage{colortbl, xcolor}
 
@@
 group may be expanded, trading bias for variance.
+
+\begin{itemize}
+\item Add SDC collapsing as application
+\end{itemize}
 
@@
   $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
-  $R = R\cup \{(a,\phi(a))\}$\tcp*{aggregate and add to result}
+  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
 }
@@
 \subsection{Split-apply-combine with collapsing groups}
+\label{sect:saccg}
 The goal of the algorithm is to compute a value for each 
@@
 returns the corresponding label in $A_k$. This can formally be defined 
-as consecutive application of $f_1, f_2\cdots, f_k$ to an element of 
+as consecutive application of $f_1, f_2,\cdots, f_k$ to an element of 
 $A_0$. In notation
@@
 collapsing group can be found, as a set of triples $(a,k,x)\in A_0\times
-A_k\times X $ with $0\leq k\leq n$.}
+\underline{n}\times X $ where $\underline{n}=\{0,1,\ldots,n\}$.}
 
@@
   }
-  \If{$\beta(d)$}{ \label{line:cond}
+  \If{$i<n \lor \beta(d)$}{ \label{line:cond}
     $R = R\cup \{(a, i,\phi(d))\}$\;
@@
 $\phi(d)$ pertains, and $i$ is the number of collapses applied to reach a
-suitable dataset. The condition in Line~\ref{line:cond} ensures that if
-no suitable dataset is found after the whole collapsing sequence has been
+suitable dataset. The condition in Line~\ref{line:cond} ensures that if no
+suitable dataset is found after the whole collapsing sequence has been
 executed, then no answer is returned.
@@
 \section{The \texttt{rcollapse} package}
+Grouped aggregation with a fall-though scenario based on a collapsing scheme
+requires a lot of parameters to be specified by the user. Besides the data, one
+needs to specify the method(s) of aggregation, the collapsing scheme, and the
+condition to decide whether a subset is fit for aggregation or a next collapse
+is necessary. The main function of the package is caled \texttt{collapse}
+and has the following signature.
+\begin{verbatim}
+  collapse(data, collapse, test, ...)
+\end{verbatim}
+where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
+sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
+and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
+of \texttt{label = expression} pairs that represent aggregation
+functions, in a similar style as \texttt{dplyr::summarize}. The output
+is a data frame that can schematically be represented as follows.
+\begin{verbatim}
+  [Grouping Variables, Collapse level, Output aggregates]
+\end{verbatim}
+The first columns represent the variables that define the output grouping, the
+next column is an integer that indicates the level of collapsing used to
+compute the aggregate, and the last set of columns store the aggregates.
+
+To accommodate the different use cases, the \texttt{rcollapse} package supports
+two different interfaces for specifying collapse sequences.  The first and most
+general is the formula interface, that requires that the collapsing sequence is
+represented as variables in the data set to be aggregated. The second is the
+tabular interface discussed before. This only supports a single grouping
+variable. For hierarchical classifications there is a helper function to
+translate parent-child representations of such classifications into a
+collapsing sequence.
+
+Besides complicated collapsing schemes, users must have the ability to specify
+a possibly complex condition for determining when a dataset is suitable for
+aggregation or not. The package offers some convenience functions that support common
+use cases as well as support for specifying generic conditions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The formula interface}
+It is not uncommon for users to use multiple variables for defining collapsing
+schemes. For example when grouping by two categorical variables \texttt{A} and
+\texttt{B}, one may first want to collapse variable $\texttt{B}$ into a
+variable \texttt{B1} and if that is not sufficient, in the next step grouping
+in the second variable is dropped so grouping is only defined by \texttt{A}.
+In \texttt{rcollapse} such a scheme is denoted with the formula \texttt{A*B \~{}
+A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
+that is attempted first, and the right-hand-side denotes consecutive collapse
+levels separated by \texttt{+}.
+
+The following tables illustrate the idea. The left table represents the data
+set to be aggregated by \texttt{A} and \texttt{B}. In this example we wish to
+compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
+least three records in a group, otherwise we collapse one step.
+%
+\begin{center}
+\begin{tabular}{cccl}
+\multicolumn{4}{l}{\texttt{Input data}}\\
+\hline
+\texttt{A} & \texttt{B} & \texttt{B1} & $Y$\\
+\hline
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
+\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
+\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
+\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
+\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
+\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
+\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
+\hline
+\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
+\multicolumn{3}{l}{\texttt{Output}}\\
+\hline
+\texttt{A*B} & \texttt{level} & $t_Y$\\
+\hline
+\rowcolor{red!25}   1 11        & 0             & 7  \\
+\rowcolor{green!25} 2 12        & 1             & 56 \\
+\rowcolor{green!25} 2 13        & 1             & 56 \\
+\rowcolor{blue!25}  3 21        & 2             & 448\\
+\rowcolor{blue!25}  3 22        & 2             & 448\\
+\rowcolor{blue!25}  3 12        & 2             & 448\\
+\hline
+\end{tabular}
+\end{center}
+%
+The table on the right represents the output. Colors indicate which data was
+used. The First row in the output represents group
+$(\texttt{A}=1,\texttt{B}=11)$. The collapsing level is zero, which means
+that no collapsing was necessary. Indeed, in the data table we see that there
+are three rows with \texttt{A==1} and \texttt{B==11} with $Y$ values $1, 2$,
+and $7$, resulting in $t_Y=7$ for this group.
+
+Next, the algorithm tries to compute the total for group
+$(\texttt{A}=2,\texttt{B}=12)$ (in green) but finds that there are only two
+such rows. It now defines a new group $(\texttt{A}=2,\texttt{B1}=1)$ and finds
+that there are three records in that group so it computes $t_Y=8+16+32=56$.
+Similarly, it finds one record with $(\texttt{A}=2,\texttt{B}=13)$ and
+collapses to $(\texttt{A}=2,\texttt{B1}=1)$, again yielding $t_y=56$.
+
+Finally, for $(\texttt{A}=2,\texttt{B}=21)$ the algorithm finds only a single
+record. It collapses to $(\texttt{A}=2,\texttt{B1}=2)$ and finds two so it
+collapses further to $(\texttt{A}=2)$ to find three records.  This yields
+$t_Y=64+128+256=448$. Similarly, the groups $(\texttt{A}=3, \texttt{B}=22)$ and
+$(\texttt{A}=3,\texttt{B}=12)$ are collapsed to $(\texttt{A}=2)$.
+
+In R code the previous example can be executed as follows.
+\begin{verbatim}
+  > library(rcollapse)
+  > input <- data.frame(
+  +      A  = c(1,1,1,2,2,2,3,3,3)
+  +    , B  = c(11,11,11,12,12,13,21,22,12)
+  +    , B1 = c(1,1,1,1,1,1,2,2,1)
+  +    , Y  = 2^(0:8)
+  + )
+  >
+  > collapse(data     = input
+  +        , collapse = A*B ~ A*B1 + A
+  +        , test     = function(d) nrow(d)>=3
+  +        , tY       = sum(Y) )
+    A  B level  tY
+  1 1 11     0   7
+  2 2 12     1  56
+  3 2 13     1  56
+  4 3 21     2 448
+  5 3 22     2 448
+  6 3 12     2 448
+\end{verbatim}
+%
+The formula interface allows users to quickly experiment with different
+collapsing schemes, but only after the necessary variables have been added to
+the original data frame, which may require some extensive data manipulation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The data frame interface}
+
+\subsection{Specifying tests}
+
+\subsection{Programming over \texttt{collapse}}
+
+\section{Extensive example: imputing economic data}
+
+\section{Summary and conclusion}
 
 \begin{itemize}
-\item{Specification of collapsing sequence with multiple collapsing variables (e.g. first size class, then NACE code)}
+\item{Record-wise results}
+\item{Record-wise imputation}
+\item{Representation of hierarchical classifications: parent-child lists vs tabular}
 \end{itemize}

### 2022-11-18 pkg/R/collapse.R

@@ -67,7 +67,21 @@
 #'
 #'
 #' @examples
-#' # some example data
+#'
+#' # example of formula defining collapsing scheme
+#' input <- data.frame(
+#'    A  = c(1,1,1,2,2,2,3,3,3)
+#'  , B  = c(11,11,11,12,12,13,21,22,12)
+#'  , B1 = c(1,1,1,1,1,1,2,2,1)
+#'  , Y  = 2^(0:8)
+#' )
+#' collapse(input, collapse=A*B ~ A*B1 + A
+#'         , test = function(d) nrow(d) >= 3
+#'         , tY = sum(Y))
+#' 
+#'
+#'
+#' # example with data.frame defining collapsing scheme
 #' dat <- data.frame(A0 = c("11","12","11","22"), Y = c(2,4,6,8))
 #' # collapsing scheme
 #' csh <- data.frame(
@@ -140,23 +154,26 @@ collapse2 <- function(data, formula, test,...){
   # set up output set
   lhs_vars <- all.vars(formula[[2]])
   out <- cbind(unique(data[lhs_vars]), level=NA_integer_)
-  vals <- as.data.frame(matrix(NA, nrow=nrow(collapse), ncol=length(expres)))
+  vals <- as.data.frame(matrix(NA, nrow=nrow(out), ncol=length(exprs)))
   colnames(vals) <- names(exprs)
   out <- cbind(out, vals)
   
   for (ia in seq_len(nrow(out))){
     d <- merge(data, out[ia,lhs_vars])
     j = 0
+    rhs_vars <- ""
     while(j < length(collapse) && !test(d)){
       j <- j + 1
       rhs_vars <- all.vars(collapse[[j]])
-      d <- merge(data, d[1, rhs_vars])
+      d <- merge(data, d[1, rhs_vars,drop=FALSE])
     }
+
     if ( j < length(collapse) || test(d)){
       out[ia,length(lhs_vars) + 1] <- j
-      out[ia,(length(lhs_vars)+1):ncol(out)] <- lapply(exprs, function(e) with(d, eval(e)))
+      out[ia,(length(lhs_vars)+2):ncol(out)] <- lapply(exprs, function(e) with(d, eval(e)))
     }
   }
+  rownames(out) <- NULL
   out
 
 }

### 2022-11-18 pkg/inst/tinytest/test_collapse.R

@@ -15,4 +15,24 @@ expect_equal(out[,2],c(0,1))
 expect_equal(out[,3],c(6,6))
 expect_equal(out[,4],c(6,6))
 
+input <- data.frame(
+    A  = c(1,1,1,2,2,2,3,3,3)
+  , B  = c(11,11,11,12,12,13,21,22,12)
+  , B1 = c(1,1,1,1,1,1,2,2,1)
+  , Y  = 2^(0:8)
+)
+
+output <- data.frame(
+    A = c(1,2,2,3,3,3)
+  , B = c(11,12,13,21,22,12)
+  , level = c(0,1,1,2,2,2)
+  , tY    = c(7,56,56,448,448,448)
+)
+
+out <- collapse(data=input
+        , collapse=A*B ~ A*B1 + A
+        , test=function(d) nrow(d)>=3
+        , tY = sum(Y) )
+
+expect_equivalent(out, output)
 

### 2022-11-23 paper/paper.tex
typos

@@
 Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
-$U$ and returns a single value in some domain $X$. Here, $U$ is our data set
-and $\phi$ an aggregating function. We split $U$ into groups using the
-following notation. Let $A$ be finite set that has no more elements than $U$,
-and let $f$ be a function that takes an element of $U$ and returns a value in
-$A$. We can think of $A$ as a set of group labels, and $f$ as the function that
-assigns a label to each element of $U$. This way, $f$ divides $U$ into
-non-overlapping groups, since every element in $U$ can only get one label (that
-is just how mathematical functions work). In formal terminology we say that
-$f:U\to A$ is a \emph{partition} of $U$.
+$U$ and returns a value in some domain $X$. Here, $U$ represents a data set and
+$\phi$ an aggregating function. We split $U$ into groups using the following
+notation. Let $A$ be finite set that has no more elements than $U$, and let $f$
+be a function that takes an element of $U$ and returns a value in $A$. We can
+think of $A$ as a set of group labels, and $f$ as the function that assigns a
+label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
+groups, since every element in $U$ can only get one label.  In formal
+terminology we say that $f:U\to A$ is a \emph{partition} of $U$.
 
-In this notation, any split-apply-combine operation can be computed with the following
-algorithm.
+In this notation, any split-apply-combine operation can be computed with the
+following algorithm.
 
@@
 \Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
-\Output{$R$: the value of $\phi$ for every part of $U$ as a set pairs $(a,x)\in
+\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs $(a,x)\in
 A\times X$ }
@@
 \end{algorithm}
-Here, the output is collected in a set $R$. We also introduced some new
-notation: the notation $2^U$ stands for `the set of all subsets of $U$'. The
-notation $f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you
-give it a set of labels from $A$ and it returns all the elements of $U$ that
-have any of those labels. This definition is slightly too general for the
-current algorithm, but it will come in handy when we move to collapsing groups.
+Here, the output is collected in a set $R$ (this set is technically a function
+$A\to X$ since it pairs for each $a\in A$ with one element of $X$). We also
+introduced some new notation: the notation $2^U$ stands for `the set of all
+subsets of $U$'. The notation $f^{-1}$ stands for the formal \emph{inverse} of
+$f$. That is: you give it a set of labels from $A$ and it returns all the
+elements of $U$ that have any of those labels. This definition is slightly too
+general for the current algorithm, but it will come in handy when we move to
+collapsing groups.
 
-Let us see how the elements $U$, $f\to A$ and $\phi$ are implemented
+Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented
 in practice. Consider the signature of the R \texttt{aggregate} function for data
@@
 \end{verbatim}
-Here, \texttt{x} is a data frame where each column is a variable to be
-aggregated.  The argument \texttt{by} represents the label(s) for each record
-that are to be used for grouping. So the function $f\to A$ is implemented by
-forcing the user to make sure that the position of each label in \texttt{by}
-corresponds to the row number in the data frame \texttt{x}. The argument
-\texttt{FUN} is the function that aggregates each subset of \texttt{x}.  Here is
-an example of how a user might call this function from the R prompt.
+Here, \texttt{x} is a data frame where each column is a variable $U$ to be
+aggregated.  The argument \texttt{by} represents the record-wise label(s) that
+are used for grouping. So the function $f\to A$ is implemented by forcing the
+user to make sure that the position of each label in \texttt{by} corresponds to
+the row number in the data frame \texttt{x}. The argument \texttt{FUN} is the
+function $\phi$ that aggregates each subset of \texttt{x}.  Here is an example of how
+a user might call this function from the R prompt.
 %
@@
 \label{sect:saccg}
-The goal of the algorithm is to compute a value for each 
-subset of a dataset, possibly using values from a larger
-subset. The input of the algorithm consists again of a
-finite set $U$, and aggregation function $\phi$ that takes
-a subset of $U$ and returns a value in some domain $X$.
-Compared to Algorithm~\ref{alg:sac} we need two extra
-inputs. First, we need a user-defined function that checks whether
-a given subset $d$ of $U$ is suitable for computing $\phi(d)$.
-We will call this function $\beta: 2^U\to \mathbb{B}$, where
-$\mathbb{B}=\{\texttt{True},\texttt{False}\}$. 
-Typical tests are checking whether there are sufficient records
-available, or whether certain variables have a low enough fraction
-of missing values. Second, we need
-a \emph{collapsing scheme}
+The goal of the algorithm is to compute a value for each part of a dataset,
+possibly using values from a larger subset. The input of the algorithm consists
+again of a finite set $U$ and an aggregation function $\phi$ that takes a subset
+of $U$ and returns a value in some domain $X$.  Compared to
+Algorithm~\ref{alg:sac} we need two extra inputs. First, we need a user-defined
+function that checks whether a given subset $d$ of $U$ is suitable for
+computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
+where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
+checking whether there are sufficient records available, or whether certain
+variables have a low enough fraction of missing values. Second, we need a
+\emph{collapsing scheme}
 \begin{equation}
@@
 %
-\begin{center}
+\begin{equation}
 \begin{tabular}{rrr}
@@
 \end{tabular}
-\end{center}
+\label{eq:tablerep}
+\end{equation}
 Here, $A_0$ is the most detailed partition, splitting a dataset into seven
@@
 
-To generalize this algorithm to generic collapsing schemes, we introduce
-the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
-returns the corresponding label in $A_k$. This can formally be defined 
-as consecutive application of $f_1, f_2,\cdots, f_k$ to an element of 
-$A_0$. In notation
+In order to generalize this algorithm towards generic collapsing schemes, we
+introduce the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and
+returns the corresponding label in $A_k$. This can formally be defined as
+consecutive application of $f_1, f_2,\cdots, f_k$ to an element of $A_0$. In
+notation
 \begin{equation*}
@@
 This function asks for a set of labels in $A_k$ and returns all the labels in
-$A_0$ that are mapped to that label via the collapsing sequence of
-Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally
-write down our main Algorithm~\ref{alg:saccg}.
+$A_0$ that are mapped to those labels via the collapsing sequence of
+Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally write
+down our main Algorithm~\ref{alg:saccg}.
 
@@
 \subsection{The data frame interface}
+\eqref{eq:tablerep}
 

### 2022-12-27 pkg/R/collapse.R
renamed

@@ -75,7 +75,7 @@
 #'  , B1 = c(1,1,1,1,1,1,2,2,1)
 #'  , Y  = 2^(0:8)
 #' )
-#' collapse(input, collapse=A*B ~ A*B1 + A
+#' accumulate(input, collapse=A*B ~ A*B1 + A
 #'         , test = function(d) nrow(d) >= 3
 #'         , tY = sum(Y))
 #' 
@@ -88,7 +88,7 @@
 #'    A0 = c("11","12","22")
 #'  , A1 = c("1" ,"1", "2") 
 #' )
-#' collapse(data = dat
+#' accumulate(data = dat
 #'    , collapse = csh
 #'    , test     = function(d) if (nrow(d)<2) FALSE else TRUE
 #'    , mn = mean(Y, na.rm=TRUE)
@@ -96,7 +96,7 @@
 #' )
 #'
 #' @export
-collapse <- function(data, collapse, test, ...){
+accumulate <- function(data, collapse, test, fun=NULL, ...){
 
   if (inherits(collapse,"data.frame")){
     collapse1(as.data.frame(data), as.data.frame(collapse), test, ...)

### 2022-12-27 pkg/inst/tinytest/test_accumulate.R

@@ -4,7 +4,7 @@
 d <- data.frame(A0 = as.character(c(11,11,12)), Y = c(4,8,6))
 A <- data.frame(A0 = as.character(c(11,12)), A1 = as.character(c(1,1)))
 
-out <- collapse(data = d
+out <- accumulate(data = d
   , collapse  = A
   , test      = function(d) if (nrow(d)>=2) TRUE else FALSE
   , mn = mean(Y, na.rm=TRUE)
@@ -29,7 +29,7 @@ output <- data.frame(
   , tY    = c(7,56,56,448,448,448)
 )
 
-out <- collapse(data=input
+out <- accumulate(data=input
         , collapse=A*B ~ A*B1 + A
         , test=function(d) nrow(d)>=3
         , tY = sum(Y) )

### 2022-12-27 pkg/R/accumulate.R
pkg functions. Needs documentation

@@ -1,0 +1,340 @@
+#' Split-Apply-Combine with Collapsing Groups
+#'
+#' Compute grouped aggregates. If a group does not satisfy certain user-defined
+#' conditions (such as too many missings, or not enough records) then the group
+#' is expanded according to a user-defined 'collapsing' scheme.  This happens
+#' recursively until either the group satisfies all conditions and the
+#' aggregate is computed, or we run out of collapsing possibilities and the
+#' \code{NA} is returned for that group.
+#'
+#'
+#' @param data \code{[data.frame]} The data to aggregate by (collapsing) groups.
+#' @param collapse \code{[formula|data.frame]} representing a group collapsing sequence.
+#'        See below for details on how to specify each option.
+#' @param test \code{[function]} A function that takes a subset of \code{data} and returns
+#'             \code{TRUE} if it is suitable for computing the desired aggregates and 
+#'             \code{FALSE} if a collapsing step is necessary.
+#' @param fun \code{[function]} A scalar function that will be applied to all columns
+#'            of \code{data}.
+#' @param ... For \code{accumulate}, extra arguments to be passed to \code{fun}. For
+#'            \code{cumulate}, a comma-separated list of \code{name=expression}, 
+#'            where \code{expression} defines the aggregating operation.
+#'
+#' @section Using a formula to define the collapsing sequence:
+#'
+#' If all combinations of collapsing options are stored as columns in
+#' \code{data}, the \code{formula} interface can be used. An example is the
+#' easiest way to see how it works is to illustrate the idea with an example.
+#' Suppose that \code{collapse = A*B ~ A1*B + B} This means:
+#' \itemize{
+#'  \item{Compute output for groups defined by variables A and B}
+#'  \item{If for a certain combination \code{(a,b)} in \code{AxB} the data does not
[…]

### 2022-12-27 pkg/R/collapse.R

@@ -1,193 +1,0 @@
-#' Split-Apply-Combine with Collapsing Groups: Multiple Aggregators
-#'
-#' Attempt to compute one or more grouped aggregates. If a group does not
-#' satisfy certain user-defined conditions (such as too many missings, or not
-#' enough records) then the group is expanded according to a user-defined
-#' 'collapsing' scheme.  This happens recursively until either the group
-#' satisfies all conditions and the aggregate is computed, or we run out of
-#' collapsing possibilities and the \code{NA} is returned for that group.
-#'
-#'
-#' @param data \code{[data.frame]} The data to aggregate by (collapsing) groups.
-#' @param collapse \code{[formula|data.frame]} representing a group collapsing sequence.
-#'        See below for details on how to specify each option.
-#' @param test \code{[function]} A function that takes a subset of \code{data} and returns
-#'             \code{TRUE} if it is suitable for computing the desired aggregates and 
-#'             \code{FALSE} if a collapsing step is necessary.
-#' @param ... A comma-separated list of \code{name=expression}, where \code{expression}
-#'            defines the aggregating operation.
-#'
-#' @section Using a formula to define the collapsing sequence:
-#'
-#' If all combinations of collapsing options are stored as columns in
-#' \code{data}, the \code{formula} interface can be used. An example is the
-#' easiest way to see how it works is to illustrate the idea with an example.
-#' Suppose that \code{collapse = A*B ~ A1*B + B} This means:
-#' \itemize{
-#'  \item{Compute output for groups defined by variables A and B}
-#'  \item{If for a certain combination \code{(a,b)} in \code{AxB} the data does not
-#'        pass the \code{test}, use \code{(a1,b)} in \code{A1xB} as alternative combination to compute
-#'        a value for \code{(a,b)} (\code{A1xB} must yield larger groups than \code{AxB}).}
-#'  \item{If that does not work, use only \code{B} as a grouping variable to compute
-#'        a value for \code{(a,b)}}.
[…]

### 2022-12-27 pkg/R/helpers.R

@@ -1,3 +1,8 @@
+stopf <- function(fmt,...) stop(sprintf(fmt,...), call.=FALSE)
+
+last <- function(x) x[length(x)]
+
+
 #' Demand minimal number of functons
 #' 
 #' @param n Minimal number of records in a group.

### 2022-12-27 pkg/inst/tinytest/test_accumulate.R

@@ -1,10 +1,57 @@
 
 #source("../../R/rcollapse.R")
 
+## Accumulate with collapsing sequence as data frame
+input    <- data.frame(Y1 = 2^(0:8), Y2 = 2^(0:8))
+input$Y2[c(1,4,7)] <- NA
+
+input$A0 <- c(123,123,123,135,136,137,212,213,225)
+
+
+collapse <- data.frame(
+      A0   = c(123, 135, 136, 137, 212, 213, 225)
+    , A1   = c(12 , 13 , 13 , 13 , 21 , 21 , 22 )
+    , A2   = c(1  , 1  , 1  , 1  , 2  , 2  , 2  )
+)
+
+out <- accumulate(input
+         , collapse
+         , test = function(d) nrow(d)>=3
+         , fun  = sum, na.rm=TRUE)
+
+expect_equal(out[,1], unique(collapse[,1]))
+expect_equal(out[,2], c(0, 1 ,  1,  1,   2, 2  , 2))
+expect_equal(out[,3], c(7, 56, 56, 56, 448, 448, 448))
+expect_equal(out[,4], c(6, 48, 48, 48, 384, 384, 384))
+
+## Accumulate with collapsing scheme as formula
+input <- data.frame(
+    A  = c(1,1,1,2,2,2,3,3,3)
+  , B  = c(11,11,11,12,12,13,21,22,12)
+  , B1 = c(1,1,1,1,1,1,2,2,1)
+  , Y1 = 2^(0:8)
+  , Y2 = 2^(0:8)
+)
+
+input$Y2[c(1,4,7)] <- NA
+
+out <- accumulate(input
+  , collapse = A*B ~ A*B1 + A
+  , test=function(a) nrow(a)>=3
+  , fun = sum, na.rm=TRUE)
+
+
+expect_equal( out[,1], c( 1,  2,  2,   3,   3,   3) )
+expect_equal( out[,2], c(11, 12, 13,  21,  22,  12) )
+expect_equal( out[,3], c( 0,  1,  1,   2,   2,   2) )
+expect_equal( out[,4], c( 7, 56, 56, 448, 448, 448) )
+expect_equal( out[,5], c( 6, 48, 48, 384, 384, 384) )
+
+
 d <- data.frame(A0 = as.character(c(11,11,12)), Y = c(4,8,6))
 A <- data.frame(A0 = as.character(c(11,12)), A1 = as.character(c(1,1)))
 
-out <- accumulate(data = d
+out <- cumulate(data = d
[…]

### 2022-12-27 pkg/tests/tinytest.R

@@ -1,5 +1,5 @@
 
 if ( requireNamespace("tinytest", quietly=TRUE) ){
-  tinytest::test_package("rcollapse")
+  tinytest::test_package("accumulate")
 }
 

### 2022-12-27 pkg/R/accumulate.R
updated docs

@@ -5,7 +5,12 @@
 #' is expanded according to a user-defined 'collapsing' scheme.  This happens
 #' recursively until either the group satisfies all conditions and the
 #' aggregate is computed, or we run out of collapsing possibilities and the
-#' \code{NA} is returned for that group.
+#' \code{NA} is returned for that group. 
+#' \itemize{
+#'  \item{\code{accumulate} aggregates over all non-grouping variables defined in 
+#'  \code{collapse}} 
+#' \item{\code{cumulate} uses a syntax akin to \code{dplyr::summarise}}
+#' }
 #'
 #'
 #' @param data \code{[data.frame]} The data to aggregate by (collapsing) groups.
@@ -24,8 +29,8 @@
 #'
 #' If all combinations of collapsing options are stored as columns in
 #' \code{data}, the \code{formula} interface can be used. An example is the
-#' easiest way to see how it works is to illustrate the idea with an example.
-#' Suppose that \code{collapse = A*B ~ A1*B + B} This means:
+#' easiest way to see how it works.  Suppose that \code{collapse = A*B ~ A1*B +
+#' B} This means:
 #' \itemize{
 #'  \item{Compute output for groups defined by variables A and B}
 #'  \item{If for a certain combination \code{(a,b)} in \code{AxB} the data does not
@@ -36,14 +41,14 @@
 #'  \item{If that does not work, return \code{NA} for that particular combination \code{(a,b)}.}
 #' }
 #' Generally, the \code{formula} must be of the form \code{X0 ~ X1 + X2 + ... +
-#' Xn} where each \code{Xi} is a product of grouping variables in the data set.
+#' Xn} where each \code{Xi} is a (product of) grouping variable(s) in the data set.
 #'
-#' @section Using a data frame to define the collapsing sequence:
+#' @section Using a data frame to define the collapsing scheme:
 #'
-#' In this case \code{collapse} is a data frame with columns
-#' \code{[A0, A1, ..., An]}. The variable \code{A0} must also be
-#' present in \code{data}, and should represent a collapsed version
-#' of \code{A0}. This means:
+#' In this case \code{collapse} is a data frame with columns \code{[A0, A1,
+#' ..., An]}.  The variable \code{A0} represents the most fine-grained
+#' grouping and must also be present in \code{data}. Aggregation works
+#' as follows.
 #' \itemize{
[…]

### 2022-12-27 pkg/inst/tinytest/test_helpers.R

@@ -4,4 +4,6 @@ expect_equal(fraction_complete(c(1,NA,2), 1:3), 2/3)
 expect_equal(n_incomplete(c(1,NA,2), 1:3), 1)
 expect_equal(fraction_incomplete(c(1,NA,2), 1:3), 1/3)
 
+expect_error(accumulate:::stopf("foo %g",bar))
+expect_equal(accumulate:::last(1:3),3)
 

### 2022-12-28 pkg/R/helpers.R
added example data and vignette

@@ -170,9 +170,8 @@ csh_from_digits <- function(x,pad="0"){
   for ( i in seq_len(nlevels)){
     A[,i] <- substr(x,1,nlevels+1-i)
   }
-  rownames(A) <- x
   colnames(A) <- sprintf("A%d",seq_len(nlevels)-1)
-  A
+  as.data.frame(A)
 }
 
 

### 2022-12-28 pkg/R/producers.R

@@ -1,0 +1,26 @@
+#' @name producers
+#' @title synthetic data on producers
+#' @description 
+#'   A synthetic dataset listing several sources of turnover
+#'   and other income for producers. The producers are classified
+#'   in size classes and SBI (a refinement of NACE). Load with \code{data(producers)}.
+#'
+#'
+#' \itemize{
+#'  \item  sbi: Classification of economic activity (refinement of NACE2008)
+#'  \item  size: Size class in 0 (smallest) to 9.
+#'  \item  industrial: Turnover from industrial activities.
+#'  \item  trade: Turnover from trade
+#'  \item  other: Turnover from other activities
+#'  \item  other_income: Income not from turnover (e.g. from financial transactions)
+#'  \item  total: Rowwise sum of indistrial, trade, and other turnover and other income.
+#'}
+#'
+#' @family datasets
+#'
+#' @docType data
+#' @format A \code{.rda} file, one producer per row.
+NULL
+
+
+

### 2022-12-29 paper/article.Rnw
some writing, moved to jss format

--- 2022-11-23 paper/paper.tex
@@
-\documentclass[11pt, fleqn]{article}
-\usepackage[english]{babel}
+\documentclass[article,table]{jss}
+
+%% -- LaTeX packages and custom commands ---------------------------------------
+%% recommended packages
+\usepackage{orcidlink,thumbpdf,lmodern}
+
 \usepackage{amsmath, amssymb}
-
-\usepackage{colortbl, xcolor}
-
 \usepackage[ruled, linesnumbered]{algorithm2e}
 
-\usepackage{natbib}
-\setlength{\parskip}{\baselineskip}%
-\setlength{\parindent}{0pt}%
+%% new custom commands
+\newcommand{\class}[1]{`\code{#1}'}
+\newcommand{\fct}[1]{\code{#1()}}
 
-\title{Spit-Apply-Combine with Collapsing Groups}
-\author{Mark P.J. van der Loo}
-\date{CONCEPT \today{}}
+%% For Sweave-based articles about R packages:
+%% need no \usepackage{Sweave}
+\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
+<<preliminaries, echo=FALSE, results=hide>>=
+options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+@
+
+%% -- Article metainformation (author, title, ...) -----------------------------
+
+%% - \author{} with primary affiliation (and optionally ORCID link)
+%% - \Plainauthor{} without affiliations
+%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
+%% - \AND starts a new line, \And does not.
+\author{Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
+        Statistics Netherlands and University of Leiden}
+
+\Plainauthor{Mark P.J. van der Loo}
+
+%% - \title{} in title case
+%% - \Plaintitle{} without LaTeX markup (if any)
+%% - \Shorttitle{} with LaTeX markup (if any), used as running title
+\title{Split-Apply-Combine with Collapsing Groups}
+\Plaintitle{Split-Apply-Combine with Collapsing Groups}
+\Shorttitle{Split-Apply-Combine with Collapsing Groups}
+
+\Abstract{
+foo
+}
+
+%% - \Keywords{} with LaTeX markup, at least one required
+%% - \Plainkeywords{} without LaTeX markup (if necessary)
+%% - Should be comma-separated and in sentence case.
+\Keywords{data analysis, \proglang{R}}
+\Plainkeywords{data analysis, R}
+
+%% - \Address{} of at least one author
+%% - May contain multiple affiliations for each author
+%%   (in extra lines, separated by \emph{and}\\).
+%% - May contain multiple authors for the same affiliation
+%%   (in the same first line, separated by comma).
+\Address{
+  Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
+  Research and Development\\
+  Statistics Netherlands\\
+  Henri Faasdreef 312\\
+  2492JP Den Haag, the Netherlands\\
+  E-mail: \email{mpj.vanderloo@cbs.nl}\\
+  URL: \url{https://www.markvanderloo.eu}\\
+  \emph{and}\\
+  Leiden Institute of Advanced Computer Science (LIACS)\\
+  P.O. Box 9512\\
+  2300 RA Leiden, The Netherlands\\
+}
 
 \begin{document}
-\maketitle
+
+%% -- Introduction -------------------------------------------------------------
+
+%% - In principle "as usual".
+%% - But should typically have some discussion of both _software_ and _methods_.
+%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
 
@@
 The operation of splitting a data set into non-overlapping groups, computing an
-aggregate for each group, and combining the results into a new datasets is one
-of the most common operations in data analyses. The idea behind this so-called
-split-apply-combine approach is that the calculation for each groups only uses
-data available within the group. However, in some cases the quality of data in
-a group may be insufficient for producing a reliable group aggregate. Common
-quality issues include insufficient records, or too many records with values
-for certain variables missing. In such cases it is not uncommon to use data
-from outside the target group  to estimate the group aggregate. One example is
-Small Area Estimation, where the lack of observations in one group is
-compensated by using data from similar but other groups. Another area where
-such approaches occur is group-wise imputation. For example, one may wish to
-impute a group estimate (say, mean, or median) for missing values. If there are
-too little observations to provide an accurate estimate of the group mean, the
-group may be expanded, trading bias for variance.
+aggregate for each group, and combining the results into a new dataset is one
+of the most common operations in data analyses. Indeed, any software for data
+analyses includes some functionality for this. For example, the combination of
+\code{split}/\code{lapply}/\code{unsplit} as well as \code{aggregate} have been
+a part of the \proglang{S} \citep{becker1988new} and \proglang{R} \citep{rcore}
+languages for a long time. For \proglang{R} there are several packages that
+implement functionality for this, including \pkg{plyr}
+\citep{wickham2011split}, it's successor \pkg{dplyr} \citep{wickham2022dplyr},
+it's drop-in replacement \pkg{poorman} \citep{eastwood2022poorman}, and
+\proglang{R} packages \pkg{collapse} \citep{krantz2022collapse} and
+\pkg{data.table} \citep{dowle2022datatable}. In \proglang{Python} the
+\code{pandas} package implements several methods for grouping records and
+aggregating over one or more columns in data frame objects. Similarly, the
+\pkg{DataFrames} package for \proglang{julia} implements such functionality
+\citep{kaminski2022dataframes}.
+
+In all packages mentioned, the calculation for each group uses data available
+within the group. However, there are valid use cases where a group aggregate is
+determined using attributes from out-of-group entities. One example where this
+occurs is in the area of small area estimation (see e.g.
+\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
+for example a geographical region, or a detailed population subset, where the
+number of (sampled) observations so small that the variance of the estimate
+would be unacceptably large. Small area estimation is a collection of
+bias-variance trade-offs where one `borrows statistical strength' from
+out-of-group records to obtain a stable estimate for groups with few
+observations. The out-of-group records can be obtained, for example by
+combining the original group with a group of records that are deemed similar in
+certain respects. A second area where out-of-group records play a role is in
+certain hot-deck imputation methods \cite{andridge2010review}.  In
+$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
+records that are preferably in the same group, but this condition may be
+relaxed if there are not enough records in the group. In the \pkg{VIM} package
+for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
+combination of the Gower distance and setting conditions on minimal number of
+donors. A third example comes from the field of statistical disclosure control
+\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule states
+that the value in a grouped aggregation may not be determined for more than $p$
+percent by a single entity. This occurs for example, when a certain economic
+activity in a geographic region is dominated by a single company. In such cases
+the value can either not be published, or one publishes an aggregate over
+larger groups.
+
+Common quality issues
+include insufficient records, or too many records with values for certain
+variables missing. In such cases it is not uncommon to use data from outside
+the target group  to estimate the group aggregate. One example is Small Area
+Estimation, where the lack of observations in one group is compensated by using
+data from similar but other groups. Another area where such approaches occur is
+group-wise imputation. For example, one may wish to impute a group estimate
+(say, mean, or median) for missing values. If there are too little observations
+to provide an accurate estimate of the group mean, the group may be expanded,
+trading bias for variance.
 
@@
 algorithm for split-apply-combine operations based on it. Next, we demonstrate
-the R package \texttt{rcollapse} that implements this algorithm together with
+the R package \texttt{accumulate} that implements this algorithm together with
 some utility functions that make working with collapsing schemes easier.
@@
 \end{algorithm}
+
 Here, the output is collected in a set $R$ (this set is technically a function
@@
 %
-\begin{equation}
+\begin{table}[h!]
 \begin{tabular}{rrr}
@@
 \label{eq:tablerep}
-\end{equation}
+\end{table}
+
 Here, $A_0$ is the most detailed partition, splitting a dataset into seven
@@
 
-\section{The \texttt{rcollapse} package}
+\section{The accumulate R package}
 Grouped aggregation with a fall-though scenario based on a collapsing scheme
@@
 \begin{verbatim}
-  collapse(data, collapse, test, ...)
+  accumulate(data, collapse, test, ...)
 \end{verbatim}
@@
 
-To accommodate the different use cases, the \texttt{rcollapse} package supports
+To accommodate the different use cases, the \pkg{accumulate} package supports
 two different interfaces for specifying collapse sequences.  The first and most
@@
 in the second variable is dropped so grouping is only defined by \texttt{A}.
-In \texttt{rcollapse} such a scheme is denoted with the formula \texttt{A*B \~{}
+In \pkg{accumulate} such a scheme is denoted with the formula \texttt{A*B \~{}
 A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
@@
 \begin{verbatim}
-  > library(rcollapse)
+  > library(accumulate)
   > input <- data.frame(
@@
 
-\subsection{Programming over \texttt{collapse}}
+\subsection{Programming over accumulate }
 
@@
 
-\bibliographystyle{plainnat}
-\bibliography{vanderloo}
+\bibliography{refs}
+
 \end{document}

### 2022-12-29 paper/article.Rnw
writing in progress

@@
 certain respects. A second area where out-of-group records play a role is in
-certain hot-deck imputation methods \cite{andridge2010review}.  In
+certain hot-deck imputation methods \citep{andridge2010review}.  In
 $k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
@@
 combination of the Gower distance and setting conditions on minimal number of
-donors. A third example comes from the field of statistical disclosure control
-\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule states
-that the value in a grouped aggregation may not be determined for more than $p$
-percent by a single entity. This occurs for example, when a certain economic
-activity in a geographic region is dominated by a single company. In such cases
-the value can either not be published, or one publishes an aggregate over
-larger groups.
+donors. In practice, imputation is often performed via a fall-though scenario,
+where one first tries to estimate a model within a group, but if the group is
+too small for a reliable estimate of model parameters, the group is enlarged by
+combining similar groups, similar to the small-area estimation scenario. A
+third example comes from the field of statistical disclosure control
+\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
+states that the value in a grouped aggregation may not be determined for more
+than $p$ percent by a single entity. This occurs for example, when a certain
+economic activity in a geographic region is dominated by a single company. In
+such cases the value can either not be published (suppression), or one
+publishes an aggregate over larger groups.
 

### 2022-12-30 paper/article.Rnw
writing

@@
 number of (sampled) observations so small that the variance of the estimate
-would be unacceptably large. Small area estimation is a collection of
-bias-variance trade-offs where one `borrows statistical strength' from
-out-of-group records to obtain a stable estimate for groups with few
-observations. The out-of-group records can be obtained, for example by
-combining the original group with a group of records that are deemed similar in
-certain respects. A second area where out-of-group records play a role is in
-certain hot-deck imputation methods \citep{andridge2010review}.  In
+would be unacceptably large. In small area estimation (SAE) one trades bias for
+variance by `borrowing statistical strength' from out-of-group records. The
+out-of-group records can be obtained, for example by combining the original
+small group with a group of records that are deemed similar in certain
+respects. A second area where out-of-group records play a role is in certain
+hot-deck imputation methods \citep{andridge2010review}.  In
 $k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
@@
 combining similar groups, similar to the small-area estimation scenario. A
-third example comes from the field of statistical disclosure control
+third example comes from the field of statistical disclosure control (SDC)
 \citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
@@
 such cases the value can either not be published (suppression), or one
-publishes an aggregate over larger groups.
+publishes an aggregate over larger groups. Similarly for microdata there is the
+concept of $k$-anonymity, which states that each combination of properties must
+occur at least $k$ times in a data set so as to prevent reidentification or
+disclosure of individual properties. One way to achieve this is to combine
+detailed groupings based on categorical variables into larger groups. For
+example by combining groups that in a hierarchical classification have the same
+parent.
 
-Common quality issues
-include insufficient records, or too many records with values for certain
-variables missing. In such cases it is not uncommon to use data from outside
-the target group  to estimate the group aggregate. One example is Small Area
-Estimation, where the lack of observations in one group is compensated by using
-data from similar but other groups. Another area where such approaches occur is
-group-wise imputation. For example, one may wish to impute a group estimate
-(say, mean, or median) for missing values. If there are too little observations
-to provide an accurate estimate of the group mean, the group may be expanded,
-trading bias for variance.
+Both SAE and SDC are supported by \proglang{R} and other free software. Small
+area estimation methodology has for example been implemented in the \pkg{sae}
+package of \cite{molina2015sae} and the \pkg{hbsae} package of
+\cite{boonstra2022hbsae}.  Disclosure control methodology is implemented in the
+free and open source standalone software $\tau-$Argus for tabular data and
+$\mu$-Argus for microdata \citep{argus2022}, and is also supported by
+\proglang{R} packages \pkg{sdcMicro} and \pkg{sdcTable}
+\citep{templ2015statistical, meindl2022sdctable}. Regarding imputation
+methodology, the CRAN task view on missing
+data\footnote{https://cran.r-project.org/web/views/MissingData.html} currently
+lists 203 \pkg{R} packages that support some form of estimating missing data.
+The \pkg{simputation} package \citep{loo2022simputation} seems to be the only
+one that allows for some kind of fall-through scenario for selecting methods,
+but it does not allow for dynamic grouping.
 
-\begin{itemize}
-\item Add SDC collapsing as application
-\end{itemize}
+Summarizing, we see that on one hand there are many implementations available
+for generic aggregation based on fixed groups. On the other hand there are
+domain-specific implementation for methods where dynamic grouping of a set of
+records plays a role. This paper presents a generic solution to
+split-apply-combine aggregation where groups can be collapsed dynamically in
+the form of an \pkg{R} package \pkg{accumulate} \cite{loo2022accumulate}.
 
-One approach that frequently occurs in practice is to use a so-called
-collapsing scheme. In this approach one attempts to estimate the aggregate for
-the most fine-grained grouping first.  If this is not possible, two or more
-groups are combined to provide input for estimating a value for the
-fine-grained group. The collapsing scheme defines which groups may be combined,
-and it may consist of multiple levels of collapsing; in principle to the point
-where the whole data set is used to estimate an aggregate value for a single
-group.
+The \code{accumulate} package serves the use case where a user wishes to
+compute aggregates for a certain grouping of records. However, if a certain
+instance of a group does not meet user-defined quality requirements, the set of
+records is expanded by (recursively) collapsing the grouping according to a
+user-defined scheme. For example, given some financial data on companies, one
+wishes to compute the average profit to turnover ratio for each combination of
+economic activity and size class. If for a certain combination of economic
+activity and size class there are too few records for a reliable estimate, one
+could drop size class and compute the average ratio over all records within a
+certain economic activity. Or, one could choose to course-grain economic
+activities by collapsing groups with activities that are deemed similar enough.
 
-Collapsing schemes can be the result of an optimisation, but they frequently
-result from subject matter as well. For example, in the case of hierarchical
-classifications, there is a natural ordering of groups that are similar from a
-subject matter point of view (whether that means that they can be considered
-representative for each other is another, statistical, question).  Examples
-include the NACE classification of economic activities or the ISCO
-classification of professions. Both have a tree-like structure that can be used
-to combine groups with a common ancestor in the hierarchy.
+The package has been developed with the following design choices in mind.
+First, the interface should be easy to learn for \proglang{R} users, and thus
+should resemble existing popular interfaces where possible. Second, users
+should be free to define any possibly multi-step, collapsing scheme.  Here, we
+keep in mind that collapsing schemes may be constructed manually based on
+domain knowledge and that users may want to experiment with several schemes
+before deciding on a final solution. This calls for a certain separation of
+concerns between defining collapsing schemes and applying them to data.  The
+package should also support collapsing schemes that follow naturally from
+hierarchical classification systems. Third, users should have the flexibility
+to define any quality requirement on the grouping while common quality
+requiremtns are supported out of the box. Common quality requirements include a
+minimum number of records, or a minum fraction of records without missing
+values, or a minimum number of records with non-zero values.
 
-In this paper we formalise the concept of `collapsing scheme', and give an
-algorithm for split-apply-combine operations based on it. Next, we demonstrate
-the R package \texttt{accumulate} that implements this algorithm together with
-some utility functions that make working with collapsing schemes easier.
+The rest of this paper is organized as follows. In the next section we formally
+define the concepts of the quality requirements and a collapsing scheme. We
+describe the fundamental algorithm in pseudocode and compare it with the
+standard grouping scenario. We also discuss the time and space complexity and
+some caveats with implementation and performance. This section is mainly of
+interest for developers who wish to include such functionality in their own
+package. In Section~\ref{sect:accumulate} we describe \pkg{accumulate} package
+and demonstrate how the design choices have determined its implementation.  We
+give some toy examples to extensively discuss the input and output of the main
+functions, discuss how to define tests in some detail, and also explain how one
+can program over the functions in the package.  Section~\ref{sect:example}
+works through a realistic example based on a synthetic dataset of economic
+microdata. We conclude with a summary and outlook in
+Section~\ref{sect:conclusion}.
 
@@
 
-\section{The accumulate R package}
+\section{The accumulate R package}\label{sect:accumulate}
 Grouped aggregation with a fall-though scenario based on a collapsing scheme
@@
 
-\section{Extensive example: imputing economic data}
+\section{Extensive example: imputing economic data} \label{sect:example}
 
-\section{Summary and conclusion}
+\section{Summary and conclusion}\label{sect:conclusion}
 

### 2022-12-30 paper/article.Rnw
writing

@@
 Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented
-in practice. Consider the signature of the R \texttt{aggregate} function for data
+in practice. Consider the signature of the R \code{aggregate()} function for data
 frames (we skip arguments that are not important for the discussion).
@@
 \end{verbatim}
-Here, \texttt{x} is a data frame where each column is a variable $U$ to be
-aggregated.  The argument \texttt{by} represents the record-wise label(s) that
+Here, \code{x} is a data frame where each column is a variable $U$ to be
+aggregated.  The parameter \code{by} represents the record-wise label(s) that
 are used for grouping. So the function $f\to A$ is implemented by forcing the
-user to make sure that the position of each label in \texttt{by} corresponds to
-the row number in the data frame \texttt{x}. The argument \texttt{FUN} is the
-function $\phi$ that aggregates each subset of \texttt{x}.  Here is an example of how
+user to make sure that the position of each label in \code{by} corresponds to
+the row number in the data frame \code{x}. The argument \code{FUN} is the
+function $\phi$ that aggregates each subset of \code{x}.  Here is an example of how
 a user might call this function from the R prompt.
 %
-\begin{verbatim}
-> aggregate(iris["Sepal.Length"], by=iris["Species"], FUN=mean)
-     Species Sepal.Length
-1     setosa        5.006
-2 versicolor        5.936
-3  virginica        6.588
-\end{verbatim}
+<<>>=
+aggregate(iris[1:2], by=iris["Species"], FUN=mean)
+@
 %
@@
 %
-\begin{table}[h!]
+\begin{equation}
 \begin{tabular}{rrr}
@@
 \label{eq:tablerep}
-\end{table}
+\end{equation}
 
@@
 
-\subsection{Caveat: unbalanced hierarchical classifications}
-\begin{itemize}
-\item tree with different levels of details
-\end{itemize}
+\subsection{Performance, memory, paralellization}
+Both Algorithms~\ref{alg:sac} and \ref{alg:saccg} need only $\mathcal{O}(|U|)$
+memory. The time complexity of Algorithm~\ref{alg:sac} is $\mathcal{O}(|A||U|)$
+($|A|$ the number of groups occurring in the data, and $|U|$ the number of
+records) but the actual running time is also determined by the form of the
+aggregating function $\phi$.  For simple aggregating functions such as
+summation or counting, the output can be computed with a single pass through
+$|U|$. \proglang{R} packages such as \pkg{collapse} and \pkg{data.table} have
+implemented specialized low level (\proglang{C}/\proglang{C++}) code for fast
+groupwise aggregation for some commonly occurring aggregating functions.
+Generic paralellization of Algorithm~\ref{alg:sac} can be done by paralellising
+over the grouping values $a\in A$: each worker node gets access to a set of
+records corresponding to a single group and computes the local aggregate.
+
+Algorithm~\ref{alg:saccg} has time complexity of $\mathcal{O}((n+1)|A_0||U|)$,
+Namely, in the worst case, for each $a\in A_0$ we need to collapse to the
+maximum level $A_n$. This means that for each $a\in A_0$ we need to collapse
+$n$ times, plus the initial attempt of computing aggregate without collapsing.
+The fact that we cannot generically exclude records from a calculation means
+that a generic paralellisation scheme would need to ensure that each worker
+node has access to the full dataset $|U|$. There are two types of options to do
+this: either mutliple workers gain read-only access to $U$ or each worker gets
+a copy of $U$. In the first case the workers may suffer from colliding
+read-operations which hampers linear performance improvement as a function of
+the number of workers. In the second case, one needs as many copies of the data
+as there are threads. It is possible to compute beforehand whether the dataset
+can be split into subsets that can be treated independently by determining all
+the collapsed subsets. However, this is almost as expensive as running the
+algorithm completely since it only skips the calculation of aggregates after
+collapsing.
+
+Summarizing, we see that split-apply-combine with collapsing groups has a
+higher worst-case complexity than regular split-apply-combine. Moreover,
+generic paralellization needs to either consider a specific use case (e.g.
+large or small datasets) or require user input that guides the choice of
+paralellisation scheme.
 
@@
 condition to decide whether a subset is fit for aggregation or a next collapse
-is necessary. The main function of the package is caled \texttt{collapse}
-and has the following signature.
+is necessary. There are two main functions in \pkg{accumulate} that offer
+slightly different interfaces.
 \begin{verbatim}
-  accumulate(data, collapse, test, ...)
+  accumulate(data, collapse, test, fun, ...)
+  cumulate(data, collapse, test, ...)
 \end{verbatim}
-where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
-sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
-and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
-of \texttt{label = expression} pairs that represent aggregation
-functions, in a similar style as \texttt{dplyr::summarize}. The output
-is a data frame that can schematically be represented as follows.
+Here \texttt{data} is a data frame, \texttt{collapse} represents the collapse
+sequence (as a \class{formula} or a \class{data frame}), and \texttt{test} is a
+function that accepts a subset of \texttt{data} and returns a boolean that
+indicates whether a subset is suited for aggregation. In \code{accumulate()},
+the parameter \code{fun} represents an aggregation function that is applied to
+every column of \code{data}, and the ellepsis (\code{...}) is for arguments
+that are passed as extra argument to \code{fun}. The interface of
+\code{accumulate()} is modeled after the \code{aggregate()} function in
+\proglang{R}.  In \code{cumulate}, the ellepsis is a sequence of
+comma-separated \code{name=expression} pairs in the style of \code{summarise()}
+from the \code{dplyr} package.
+
+The output of both functions are of the same form.  The columns of the output
+data frame and can schematically be represented as follows. 
 \begin{verbatim}
@@
 next column is an integer that indicates the level of collapsing used to
-compute the aggregate, and the last set of columns store the aggregates.
+compute the aggregate (0 indicating no collapse), and the last set of columns
+store the aggregates.
 
-To accommodate the different use cases, the \pkg{accumulate} package supports
-two different interfaces for specifying collapse sequences.  The first and most
-general is the formula interface, that requires that the collapsing sequence is
+Both functions support two different interfaces for specifying collapsing
+schemes through the \code{collapse} parameter. The first and most general is
+the formula interface, which requires that the collapsing sequence is
 represented as variables in the data set to be aggregated. The second is the
-tabular interface discussed before. This only supports a single grouping
-variable. For hierarchical classifications there is a helper function to
-translate parent-child representations of such classifications into a
-collapsing sequence.
-
-Besides complicated collapsing schemes, users must have the ability to specify
-a possibly complex condition for determining when a dataset is suitable for
-aggregation or not. The package offers some convenience functions that support common
-use cases as well as support for specifying generic conditions.
+tabular interface, where a table similar to~\eqref{eq:tablerep} is expected.
 
@@
 It is not uncommon for users to use multiple variables for defining collapsing
-schemes. For example when grouping by two categorical variables \texttt{A} and
-\texttt{B}, one may first want to collapse variable $\texttt{B}$ into a
-variable \texttt{B1} and if that is not sufficient, in the next step grouping
-in the second variable is dropped so grouping is only defined by \texttt{A}.
-In \pkg{accumulate} such a scheme is denoted with the formula \texttt{A*B \~{}
+schemes. For example when grouping by two categorical variables $A$ and
+$B$, one may first want to collapse variable $B$ into a
+variable $B1$ and if that is not sufficient, grouping
+in the second variable is dropped so grouping is only defined by $A$.
+In \code{accumulate()} such a scheme is denoted with the formula \code{A*B \~{}
 A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
 that is attempted first, and the right-hand-side denotes consecutive collapse
-levels separated by \texttt{+}.
+levels separated by \code{+}.
 
 The following tables illustrate the idea. The left table represents the data
-set to be aggregated by \texttt{A} and \texttt{B}. In this example we wish to
+set to be aggregated by $A$ and $B$. In this example we wish to
 compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
-least three records in a group, otherwise we collapse one step.
+least three records in a group, otherwise we collapse one step (we set values
+of $Y$ to unique powers of two, so it is clear which records are used
+to compute each value of $t_Y$).
 %
@@
 \begin{tabular}{cccl}
-\multicolumn{4}{l}{\texttt{Input data}}\\
+\multicolumn{4}{l}{Input data}\\
 \hline
-\texttt{A} & \texttt{B} & \texttt{B1} & $Y$\\
+$A$ & $B$ & $B1$ & $Y$\\
 \hline
@@
 \end{tabular}\hspace{1cm}\begin{tabular}{ccl}
-\multicolumn{3}{l}{\texttt{Output}}\\
+\multicolumn{3}{l}{Output}\\
 \hline
-\texttt{A*B} & \texttt{level} & $t_Y$\\
+$A\times B$ & \code{level} & $t_Y$\\
 \hline
@@
 used. The First row in the output represents group
-$(\texttt{A}=1,\texttt{B}=11)$. The collapsing level is zero, which means
+$(A=1,B=11)$. The collapsing level is zero, which means
 that no collapsing was necessary. Indeed, in the data table we see that there
-are three rows with \texttt{A==1} and \texttt{B==11} with $Y$ values $1, 2$,
+are three rows with $A=1$ and $B==11$ with $Y$ values $1, 2$,
 and $7$, resulting in $t_Y=7$ for this group.
@@
 Next, the algorithm tries to compute the total for group
-$(\texttt{A}=2,\texttt{B}=12)$ (in green) but finds that there are only two
-such rows. It now defines a new group $(\texttt{A}=2,\texttt{B1}=1)$ and finds
+$(A=2,B=12)$ (in green) but finds that there are only two
+such rows. It now defines a new group $(A=2,B1=1)$ and finds
 that there are three records in that group so it computes $t_Y=8+16+32=56$.
-Similarly, it finds one record with $(\texttt{A}=2,\texttt{B}=13)$ and
-collapses to $(\texttt{A}=2,\texttt{B1}=1)$, again yielding $t_y=56$.
+Similarly, it finds one record with $(A=2,B=13)$ and
+collapses to $(A=2,B1=1)$, again yielding $t_y=56$.
 
-Finally, for $(\texttt{A}=2,\texttt{B}=21)$ the algorithm finds only a single
-record. It collapses to $(\texttt{A}=2,\texttt{B1}=2)$ and finds two so it
-collapses further to $(\texttt{A}=2)$ to find three records.  This yields
-$t_Y=64+128+256=448$. Similarly, the groups $(\texttt{A}=3, \texttt{B}=22)$ and
-$(\texttt{A}=3,\texttt{B}=12)$ are collapsed to $(\texttt{A}=2)$.
+Finally, for $(A=2,B=21)$ the algorithm finds only a single
+record. It collapses to $(A=2,B1=2)$ and finds two so it
+collapses further to $(A=2)$ to find three records.  This yields
+$t_Y=64+128+256=448$. Similarly, the groups $(A=3, B=22)$ and
+$(A=3,B=12)$ are collapsed to $(A=2)$.
 
 In R code the previous example can be executed as follows.
-\begin{verbatim}
-  > library(accumulate)
-  > input <- data.frame(
-  +      A  = c(1,1,1,2,2,2,3,3,3)
-  +    , B  = c(11,11,11,12,12,13,21,22,12)
-  +    , B1 = c(1,1,1,1,1,1,2,2,1)
-  +    , Y  = 2^(0:8)
-  + )
-  >
-  > collapse(data     = input
-  +        , collapse = A*B ~ A*B1 + A
-  +        , test     = function(d) nrow(d)>=3
-  +        , tY       = sum(Y) )
-    A  B level  tY
-  1 1 11     0   7
-  2 2 12     1  56
-  3 2 13     1  56
-  4 3 21     2 448
-  5 3 22     2 448
-  6 3 12     2 448
-\end{verbatim}
+<<>>=
+library(accumulate)
+input <- data.frame(
+     A  = c(1,1,1,2,2,2,3,3,3)
+   , B  = c(11,11,11,12,12,13,21,22,12)
+   , B1 = c(1,1,1,1,1,1,2,2,1)
+   , Y  = 2^(0:8)
+)
+
+cumulate(input, collapse = A*B ~ A*B1 + A
+        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
+@
 %
@@
 collapsing schemes, but only after the necessary variables have been added to
-the original data frame, which may require some extensive data manipulation.
+the original data frame.
+
+It is possible to get the same result with \code{accumulate()}. This will 
+cause summation over all variables that are not used in the formula object. In
+the below example we also introduce the helper function \code{min_records()}.
+<<>>=
+input$Y2 <- 3^(0:8)
+accumulate(input, collapse = A*B ~ A*B1 + A
+          , test = min_records(3), fun = sum)
+@
+This means that for experimentation, users must be careful to exclude categorical
+variables that are not used from the input data set. For example, if \code{B1}
+is not used, we get the following.
+<<>>=
+accumulate(input[-3], collapse = A*B ~ A
+          , test = min_records(3), fun = sum)
+@
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The data frame interface}
-\eqref{eq:tablerep}
+\subsection{The data frame interface} \label{sect:dfint}
+The data frame interface is somewhat limited because it only allows for a
+single grouping variable. The advantage however setting up a collapsing scheme
+in the form of a table closely connects to domain knowledge and allows
+fine-grained control on how groups are collapsed.
+
+In order to use the data frame interface, the input dataset needs to have the
+most fine-grained grouping variable. In the running example this is
+$A\times B$, so we need to combine that into a single variable, and remove
+the other ones.
+<<>>=
+input$AB <- paste(input$A, input$B, sep="-")
+input <- input[-(1:3)]
+input
+@
+We now define the collapsing scheme as follows.
+<<>>=
+csh <- data.frame(
+  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
+, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
+, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
+@
+The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
+<<>>=
+accumulate(input, collapse=csh, test = min_records(3), sum)
+cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
+@
+
+\subsection{Balanced and unbalanced hierarchical classifications}
 

### 2022-12-30 paper/article.Rnw
writing

@@
   Leiden Institute of Advanced Computer Science (LIACS)\\
+  University of Leiden\\
   P.O. Box 9512\\
@@
 @
+In this data frame, two consecutive columns should be read as a child-parent
+relation. For example, in the first collapsing step the groups defined by
+\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
+In this artificial example the codes do not mean anything, but in realistic
+cases where codes represent a (hierarchical) classification, domain experts
+usually have a good grasp of which codes can be combined.
+
 The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
@@
 
+\newpage
 \bibliography{refs}

### 2023-01-02 paper/article.Rnw
writing

@@
 \usepackage[ruled, linesnumbered]{algorithm2e}
+\usepackage{tikz}
 
@@
 \newcommand{\fct}[1]{\code{#1()}}
+
+% This allows referencing institutional authors. 
+% See: https://tex.stackexchange.com/questions/162659/formatting-of-institution-as-author-with-natbib-and-numbered-references
+\newcommand{\xfnm}[1][]{\ifx!#1!\else\unskip,\space#1\fi}
 
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{Split-Apply-Combine with Collapsing Groups}
-\Plaintitle{Split-Apply-Combine with Collapsing Groups}
-\Shorttitle{Split-Apply-Combine with Collapsing Groups}
+\title{Split-Apply-Combine with Dynamic Grouping}
+\Plaintitle{Split-Apply-Combine with Dynamic Grouping}
+\Shorttitle{Split-Apply-Combine with Dynamic Grouping}
 
@@
 Section~\ref{sect:conclusion}.
+
+\section{Dynamic grouping} \label{sect:dynamicgrouping}
+In this section the illustrate the concept of dynamic grouping with a minimal
+worked example. This example is not very realistic, but it is constructed
+to be simple enough so the whole procedure can be followed in detail.
+
+We consider a data set with three categorical variables $A$, $B$ and $B_1$, and
+one numerical variable $Y$.  Variable $A$ has levels $\{1, 2,3\}$ and variable
+$B$ is a hierarchical classification with levels $\{11,12,13,21,22\}$. Variable
+$B_1$ is a coursegraining of $B$: for each record the value for $B_1$ is the
+first digit of $B$. Hence, $B_1$ has levels $\{1,2\}$.
+
+Our goal is to compute the sum $t_Y$ over $Y$, grouped by $A\times B$.  We
+impose the condition that there must be at least three records in each group.
+If a certain group $(a\in A,b\in B)$ has less than three records, we attempt
+to compute the value for that group over records in $(a,b_1)$ where $b_1$ is
+obtained by taking the first digit of $b$. If we then still have less than
+three records, we take records of group $a$ to determine the value for $(a,b)$.
+
+The following tables illustrate the idea. The left table represents the data
+set to be aggregated by $A$ and $B$.  The table on the right represents the
+output. Colors indicate which data was used. 
+%
+\begin{center}
+\begin{tabular}{cccl}
+\multicolumn{4}{l}{Input data}\\
+\hline
+$A$ & $B$ & $B1$ & $Y$\\
+\hline
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
+\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
+\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
+\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
+\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
+\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
+\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
+\hline
+\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
+\multicolumn{3}{l}{Output}\\
+\hline
+$A\times B$ & \code{level} & $t_Y$\\
+\hline
+\rowcolor{red!25}   1 11        & 0             & 7  \\
+\rowcolor{green!25} 2 12        & 1             & 56 \\
+\rowcolor{green!25} 2 13        & 1             & 56 \\
+\rowcolor{blue!25}  3 21        & 2             & 448\\
+\rowcolor{blue!25}  3 22        & 2             & 448\\
+\rowcolor{blue!25}  3 12        & 2             & 448\\
+\hline
+\end{tabular}
+\label{eq:example}
+\end{center}
+%
+The First row in the output represents group $(A=1,B=11)$. The collapsing level
+is zero, which means that no collapsing was necessary. Indeed, in the data
+table we see that there are three rows with $A=1$ and $B==11$ with $Y$ values
+$1, 2$, and $7$, resulting in $t_Y=7$ for this group.
+
+Next, we try to compute the total for group $(A=2,B=12)$ (in green) but find
+that there are only two such rows. We now define a new group $(A=2,B_1=1)$ and
+find that there are three records in that group so we get 
+$t_Y=8+16+32=56$.  Similarly, there is only one record with $(A=2,B=13)$. 
+Collapsing groups to $(A=2,B1=1)$, yieds again $t_y=56$.
+
+Finally, for $(A=2,B=21)$ there is only a single record. Collapsing to
+$(A=2,B1=2)$ yields only two records, so we need to collapse further to $(A=2)$
+and finally obtain three records.  This yields $t_Y=64+128+256=448$. Similarly,
+the groups $(A=3, B=22)$ and $(A=3,B=12)$ are collapsed to $(A=2)$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5%%%%%%%
+\section{The accumulate R package}\label{sect:accumulate}
+Grouped aggregation with a fall-though scenario based on a collapsing scheme
+requires a fair amount of specification by the user. Besides the data to be
+aggregated, one needs to specify the method(s) of aggregation, the collapsing
+scheme, and the condition to decide whether a subset is fit for aggregation or
+a next collapse is necessary. There are two main functions in \pkg{accumulate}
+that offer slightly different interfaces.
+\begin{verbatim}
+  accumulate(data, collapse, test, fun, ...)
+  cumulate(data, collapse, test, ...)
+\end{verbatim}
+Here \texttt{data} is a data frame holding data to be aggregated;
+\texttt{collapse} represents the collapse sequence (as a \class{formula} or a
+\class{data frame}), and \texttt{test} is a function that accepts a subset of
+\texttt{data} and returns a boolean that indicates whether a subset is suited
+for aggregation or not. In \code{accumulate()}, the parameter \code{fun}
+represents an aggregation function that is applied to every column of
+\code{data}, and the ellepsis (\code{...}) is for arguments that are passed as
+extra argument to \code{fun}. The interface of \code{accumulate()} is modeled
+after the \code{aggregate()} function in \proglang{R}.  In \code{cumulate}, the
+ellepsis is a sequence of comma-separated \code{name=expression} pairs in the
+style of \code{summarise()} from the \code{dplyr} package.
+
+The output of both functions are of the same form. The columns of the output
+data frame and can schematically be represented as follows. 
+\begin{verbatim}
+  [Grouping Variables, Collapse level, Output aggregates]
+\end{verbatim}
+The first columns represent the variables that define the output grouping, the
+next column is an integer that indicates the level of collapsing used to
+compute the aggregate (0 indicating no collapse), and the last set of columns
+store the aggregates.
+
+Both functions support two different interfaces for specifying collapsing
+schemes through the \code{collapse} parameter. The first and most general is
+the formula interface, which requires that the collapsing sequence is
+represented as variables in the data set to be aggregated. The second is the
+tabular interface, where a table similar to~\eqref{eq:tablerep} is expected.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The formula interface}
+We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
+the formula interface.
+<<>>=
+library(accumulate)
+input <- data.frame(
+     A  = c(1,1,1,2,2,2,3,3,3)
+   , B  = c(11,11,11,12,12,13,21,22,12)
+   , B1 = c(1,1,1,1,1,1,2,2,1)
+   , Y  = 2^(0:8)
+)
+
+cumulate(input, collapse = A*B ~ A*B1 + A
+        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
+@
+%
+Consider the formula \code{A*B ~ A*B1 + A} in the call to \code{cumulate()}.
+The left-hand-side \code{A*B} is the target output grouping. The
+ritght-hand-side is to be interpreted as the collapsing sequnce: if an instance
+of \code{A*B} does not pass the test, then collapse to \code{A*B1}, and if that
+does not pass the test collapse to \code{A}. If this final grouping also does
+not pass the test, the result is \code{NA}.
+
+Summarizing, the formula interface is always of the follwing form.
+\begin{verbatim}
+  Target grouping ~ Alternative1 + Alternative2 + ... + AlternativeN
+\end{verbatim}
+
+It is possible to get the same result with \code{accumulate()}. This will cause
+summation over all variables that are not used in the formula object. In the
+below example we also introduce the helper function \code{min_records()}.
+<<>>=
+input$Y2 <- 3^(0:8)
+accumulate(input, collapse = A*B ~ A*B1 + A
+          , test = min_records(3), fun = sum)
+@
+This means that for experimentation, users must be careful to exclude categorical
+variables that are not used from the input data set. For example, if \code{B1}
+is not used, we get the following.
+<<>>=
+accumulate(input[-3], collapse = A*B ~ A
+          , test = min_records(3), fun = sum)
+@
+
+The formula interface allows users to quickly experiment with different
+collapsing schemes. It does require that all categorical variables have been
+added to the data. As an alternative, one can use the data frame specification,
+which allows for a further separation between defining the collapsing scheme
+and the actual data processing.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The data frame interface} \label{sect:dfint}
+The data frame interface is somewhat limited because it only allows for a
+single grouping variable. The advantage however setting up a collapsing scheme
+in the form of a table closely connects to domain knowledge and allows
+fine-grained control on how groups are collapsed.
+
+In order to use the data frame interface, the input dataset must include the
+most fine-grained grouping variable. In the running example this is $A\times
+B$, so we need to combine that into a single variable, and remove the other
+ones.
+<<>>=
+input$AB <- paste(input$A, input$B, sep="-")
+input <- input[-(1:3)]
+input
+@
+We now define the collapsing scheme as follows.
+<<>>=
+csh <- data.frame(
+  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
+, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
+, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
+csh
+@
+In this data frame, two consecutive columns should be read as a child-parent
+relation. For example, in the first collapsing step the groups defined by
+\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
+In this artificial example the codes do not mean anything, but in realistic
+cases where codes represent a (hierarchical) classification, domain experts
+usually have a good grasp of which codes can be combined.
+
+The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
+<<>>=
+accumulate(input, collapse=csh, test = min_records(3), sum)
+cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Balanced and unbalanced hierarchical classifications}
+Hierarchical classifications are abundant in (official) statistics.
+They represent a classification of entities into non-overlapping, nested
+groupings. Examples include the International Standard Industrial Classification
+of Economic Activities (ISIC, \citet{un2022isic}), the related 
+Statistical Classification of Economic Activities in Europe (NACE, \citet{eu2006nace})
+and the Euroean Skills, Competences, Qualifications and Occupations classification
+(ESCO, \citet{eu2022esco}).
+
+Hierarchical classifications offer a natural mechanism for collapsing fine-graind groupings
+into larger groups because of the parent-child relationships. As an example consider
+a small piece of the NACE classification.
+
+\begin{center}
+\begin{tikzpicture}
+\tikzstyle{level 1}=[level distance=10mm,sibling distance=40mm]
+\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
+%  every node/.style = {shape=rectangle, rounded corners,
+%    draw, align=center,
+%    top color=white, bottom color=blue!20}]]
+  \node {\code{01}}
+    child { node {\code{011}}
+      child {node {\code{0111}}}
+      child {node {\code{0112}}} 
+      child {node {\code{0113}}} }
+    child { node {\code{012}}
+      child {node {\code{0121}}}
+      child {node {\code{0122}}} 
+      child {node {\code{0123}}} 
+      child {node {\code{0124}}} };
+\end{tikzpicture}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Specifying tests}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Programming over accumulate }
+
+\section{Extensive example: economic data} \label{sect:example}
 
@@
 
-\section{The accumulate R package}\label{sect:accumulate}
-Grouped aggregation with a fall-though scenario based on a collapsing scheme
-requires a lot of parameters to be specified by the user. Besides the data, one
-needs to specify the method(s) of aggregation, the collapsing scheme, and the
-condition to decide whether a subset is fit for aggregation or a next collapse
-is necessary. There are two main functions in \pkg{accumulate} that offer
-slightly different interfaces.
-\begin{verbatim}
-  accumulate(data, collapse, test, fun, ...)
-  cumulate(data, collapse, test, ...)
-\end{verbatim}
-Here \texttt{data} is a data frame, \texttt{collapse} represents the collapse
-sequence (as a \class{formula} or a \class{data frame}), and \texttt{test} is a
-function that accepts a subset of \texttt{data} and returns a boolean that
-indicates whether a subset is suited for aggregation. In \code{accumulate()},
-the parameter \code{fun} represents an aggregation function that is applied to
-every column of \code{data}, and the ellepsis (\code{...}) is for arguments
-that are passed as extra argument to \code{fun}. The interface of
-\code{accumulate()} is modeled after the \code{aggregate()} function in
-\proglang{R}.  In \code{cumulate}, the ellepsis is a sequence of
-comma-separated \code{name=expression} pairs in the style of \code{summarise()}
-from the \code{dplyr} package.
-
-The output of both functions are of the same form.  The columns of the output
-data frame and can schematically be represented as follows. 
-\begin{verbatim}
-  [Grouping Variables, Collapse level, Output aggregates]
-\end{verbatim}
-The first columns represent the variables that define the output grouping, the
-next column is an integer that indicates the level of collapsing used to
-compute the aggregate (0 indicating no collapse), and the last set of columns
-store the aggregates.
-
-Both functions support two different interfaces for specifying collapsing
-schemes through the \code{collapse} parameter. The first and most general is
-the formula interface, which requires that the collapsing sequence is
-represented as variables in the data set to be aggregated. The second is the
-tabular interface, where a table similar to~\eqref{eq:tablerep} is expected.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The formula interface}
-It is not uncommon for users to use multiple variables for defining collapsing
-schemes. For example when grouping by two categorical variables $A$ and
-$B$, one may first want to collapse variable $B$ into a
-variable $B1$ and if that is not sufficient, grouping
-in the second variable is dropped so grouping is only defined by $A$.
-In \code{accumulate()} such a scheme is denoted with the formula \code{A*B \~{}
-A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
-that is attempted first, and the right-hand-side denotes consecutive collapse
-levels separated by \code{+}.
-
-The following tables illustrate the idea. The left table represents the data
-set to be aggregated by $A$ and $B$. In this example we wish to
-compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
-least three records in a group, otherwise we collapse one step (we set values
-of $Y$ to unique powers of two, so it is clear which records are used
-to compute each value of $t_Y$).
-%
-\begin{center}
-\begin{tabular}{cccl}
-\multicolumn{4}{l}{Input data}\\
-\hline
-$A$ & $B$ & $B1$ & $Y$\\
-\hline
-\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
-\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
-\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
-\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
-\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
-\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
-\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
-\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
-\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
-\hline
-\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
-\multicolumn{3}{l}{Output}\\
-\hline
-$A\times B$ & \code{level} & $t_Y$\\
-\hline
-\rowcolor{red!25}   1 11        & 0             & 7  \\
-\rowcolor{green!25} 2 12        & 1             & 56 \\
-\rowcolor{green!25} 2 13        & 1             & 56 \\
-\rowcolor{blue!25}  3 21        & 2             & 448\\
-\rowcolor{blue!25}  3 22        & 2             & 448\\
-\rowcolor{blue!25}  3 12        & 2             & 448\\
-\hline
-\end{tabular}
-\end{center}
-%
-The table on the right represents the output. Colors indicate which data was
-used. The First row in the output represents group
-$(A=1,B=11)$. The collapsing level is zero, which means
-that no collapsing was necessary. Indeed, in the data table we see that there
-are three rows with $A=1$ and $B==11$ with $Y$ values $1, 2$,
-and $7$, resulting in $t_Y=7$ for this group.
-
-Next, the algorithm tries to compute the total for group
-$(A=2,B=12)$ (in green) but finds that there are only two
-such rows. It now defines a new group $(A=2,B1=1)$ and finds
-that there are three records in that group so it computes $t_Y=8+16+32=56$.
-Similarly, it finds one record with $(A=2,B=13)$ and
-collapses to $(A=2,B1=1)$, again yielding $t_y=56$.
-
-Finally, for $(A=2,B=21)$ the algorithm finds only a single
-record. It collapses to $(A=2,B1=2)$ and finds two so it
-collapses further to $(A=2)$ to find three records.  This yields
-$t_Y=64+128+256=448$. Similarly, the groups $(A=3, B=22)$ and
-$(A=3,B=12)$ are collapsed to $(A=2)$.
-
-In R code the previous example can be executed as follows.
-<<>>=
-library(accumulate)
-input <- data.frame(
-     A  = c(1,1,1,2,2,2,3,3,3)
-   , B  = c(11,11,11,12,12,13,21,22,12)
-   , B1 = c(1,1,1,1,1,1,2,2,1)
-   , Y  = 2^(0:8)
-)
-
-cumulate(input, collapse = A*B ~ A*B1 + A
-        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
-@
-%
-The formula interface allows users to quickly experiment with different
-collapsing schemes, but only after the necessary variables have been added to
-the original data frame.
-
-It is possible to get the same result with \code{accumulate()}. This will 
-cause summation over all variables that are not used in the formula object. In
-the below example we also introduce the helper function \code{min_records()}.
-<<>>=
-input$Y2 <- 3^(0:8)
-accumulate(input, collapse = A*B ~ A*B1 + A
-          , test = min_records(3), fun = sum)
-@
-This means that for experimentation, users must be careful to exclude categorical
-variables that are not used from the input data set. For example, if \code{B1}
-is not used, we get the following.
-<<>>=
-accumulate(input[-3], collapse = A*B ~ A
-          , test = min_records(3), fun = sum)
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The data frame interface} \label{sect:dfint}
-The data frame interface is somewhat limited because it only allows for a
-single grouping variable. The advantage however setting up a collapsing scheme
-in the form of a table closely connects to domain knowledge and allows
-fine-grained control on how groups are collapsed.
-
-In order to use the data frame interface, the input dataset needs to have the
-most fine-grained grouping variable. In the running example this is
-$A\times B$, so we need to combine that into a single variable, and remove
-the other ones.
-<<>>=
-input$AB <- paste(input$A, input$B, sep="-")
-input <- input[-(1:3)]
-input
-@
-We now define the collapsing scheme as follows.
-<<>>=
-csh <- data.frame(
-  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
-, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
-, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
-@
-In this data frame, two consecutive columns should be read as a child-parent
-relation. For example, in the first collapsing step the groups defined by
-\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
-In this artificial example the codes do not mean anything, but in realistic
-cases where codes represent a (hierarchical) classification, domain experts
-usually have a good grasp of which codes can be combined.
-
-The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
-<<>>=
-accumulate(input, collapse=csh, test = min_records(3), sum)
-cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
-@
-
-\subsection{Balanced and unbalanced hierarchical classifications}
-
-\subsection{Specifying tests}
-
-\subsection{Programming over accumulate }
-
-\section{Extensive example: imputing economic data} \label{sect:example}
-
 \section{Summary and conclusion}\label{sect:conclusion}

### 2023-01-02 pkg/R/helpers.R
added levels parameter

@@ -158,11 +158,13 @@ fraction_complete <- function(...){
 #' parents in the hierarchy.
 #'
 #' @param x \code{[character|integer]} labels in a hierarchical classification (lowest level)
-#' @param pad \code{[character]} padding character used for codes shorter than the maximum length.
+#' @param levels \code{[integer]>=0} how many collapsing levels to include. Zero means
+#'        only include the original labels.
 #'
 #' 
 #' @export
-csh_from_digits <- function(x,pad="0"){
+csh_from_digits <- function(x, levels=max(nchar(x))-1){
+  stopifnot(levels>=0)
   x <- as.character(x)
   nlevels <- max(nchar(x))
 
@@ -171,7 +173,7 @@ csh_from_digits <- function(x,pad="0"){
     A[,i] <- substr(x,1,nlevels+1-i)
   }
   colnames(A) <- sprintf("A%d",seq_len(nlevels)-1)
-  as.data.frame(A)
+  as.data.frame(A)[1:(levels+1)]
 }
 
 

### 2023-01-02 paper/article.Rnw
writing

@@
 \tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
-%  every node/.style = {shape=rectangle, rounded corners,
-%    draw, align=center,
-%    top color=white, bottom color=blue!20}]]
   \node {\code{01}}
@@
 \end{center}
+Here, the hierarchy suggest to collapsing $\{0111,0112,0113\}$ into $\{011\}$
+(when needed) and similar for the right branch. The second level of collapsing
+would combine $\{012\}$ with $\{011\}$ into $\{01\}$. The \pkg{accumulate}
+package comes with a helper function that creates the collasping scheme from the
+lowest-level digits.
+<<>>=
+nace <- c("0111","0112","0113","0121","0121","0122","0123","0124")
+csh_from_digits(nace, levels=2)
+@ 
+Here, the parameter \code{levels} determines how many collapsing steps will be
+computed. Since all codes are prepended with zero, there is no need to collapse
+01 any further. The output can be used as argument to the \code{collapse}
+parameter of the \code{accumulate()} or \code{cumulate} functions.
+
+The situation becomes a little more involved when hierarchical classifcations
+form a tree such that the distance from leave to trunk is not the same for all
+leaves (unbalanced tree). This occurs in practice, for example when local
+organisations create an extra level of detail for some, but not all leaves.
+Below is an example of such a situation.
+%
+\begin{center}
+\begin{tikzpicture}
+\tikzstyle{level 1}=[level distance=10mm,sibling distance=50mm]
+\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
+\tikzstyle{level 2}=[level distance=10mm,sibling distance=15mm]
+  \node {\code{01}}
+    child { node {\code{011}}
+      child {node {\code{0111}}}
+      child {node {\code{0112}}} 
+      child {node {\code{0113}}} }
+    child { node {\code{012}}
+      child {node {\code{0121}}}
+      child {node {\code{0122}}} 
+      child {node {\code{0123}}} 
+      child {node {\code{0124}}
+       child {node {\code{01241}}}
+       child {node {\code{01242}}} } };
+\end{tikzpicture}
+\end{center}
+%
+In this case, not all leaves can be collapsed with the same number of
+collapsing levels. This provides an issue for specifying the collapsing
+sequence as it now depends on the leaf where you start how many collapsing
+levels are possible. It also complicates interpretability of the result as the
+collapsing level reported in the output, now means different things for
+different target groups. The solution chosen in \pkg{accumulate} is to extend
+the tree by making copies of the leaves that are not on the lowest level as
+follows.
+%
+\begin{center}
+\begin{tikzpicture}
+\tikzstyle{level 1}=[level distance=10mm,sibling distance=70mm]
+\tikzstyle{level 2}=[level distance=10mm,sibling distance=20mm]
+\tikzstyle{level 3}=[level distance=10mm,sibling distance=15mm]
+  \node {\code{01}}
+    child { node {\code{011}}
+      child {node {\code{0111}}
+       child {node{\code{0111}}}} 
+      child {node {\code{0112}} 
+       child {node{\code{0112}}}} 
+      child {node {\code{0113}}
+       child {node{\code{0113}}}} } 
+    child { node {\code{012}}
+      child {node {\code{0121}}
+       child {node {\code{0121}}}}
+      child {node {\code{0122}}
+       child {node{\code{0122}}}} 
+      child {node {\code{0123}}
+       child {node{\code{0123}}}} 
+      child {node {\code{0124}}
+       child {node {\code{01241}}}
+       child {node {\code{01242}}} } };
+\end{tikzpicture}
+\end{center}
+%
+The tradeoff is that although there may be some extra calculations in the case
+where a leaf is collapsed to itself. The gain is that the specification of the
+calculation as well as the interpretation of the results are now uniform
+accross all hierarchical classifications. Again, using \code{csh_from_digits()}
+deriving the collapsing scheme can be automated.
+<<>>=
+nace <- c("0111","0112","0113","0121","0122","0123","01241","01242")
+csh_from_digits(nace,levels=3)
+@
 

### 2023-01-03 pkg/R/accumulate.R
better helpers, integration with validate

@@ -71,8 +71,7 @@
 #' \code{test}, then for that row, the \code{level} and subsequent columns are
 #' \code{NA}.
 #' 
-#' @seealso \code{\link{demand}}
-#'
+#' 
 #' @examples
 #'
 #' ## Example of data frame defining collapsing scheme, using accumulate

### 2023-01-03 pkg/R/helpers.R

@@ -1,6 +1,10 @@
+
 stopf <- function(fmt,...) stop(sprintf(fmt,...), call.=FALSE)
 
 last <- function(x) x[length(x)]
+
+complete_cases <- function(d) !Reduce(`|`, lapply(d, is.na))
+
 
 
 #' Demand minimal number of functons
@@ -16,140 +20,93 @@ last <- function(x) x[length(x)]
 #'
 #' @export
 min_records <- function(n){
+  stopifnot(is.numeric(n), n>=0)
   min_rec <- n
   function(d) if (nrow(d) < n) FALSE else TRUE
 }
 
-#' Demand maximal number of records
+#' Demand minimal number of complete records
 #'
-#' @param n Minimal number of records in a group.
+#' @param n Minimal number of records that must be complete
+#' @param vars \code{[TRUE|column index]} Column index into the data to
+#'        be tested (e.g. a character vectod with variable names or
+#'        a numeric vector with column positions). The indexed columns
+#'        will be testsed for completeness (absence of \code{NA}). Be default
+#'        \code{vars=TRUE} meaning that all columns are taken into account.
+#'
+#' @family helpers
 #'
 #' @examples
 #'
-#' max_records(5)(women)
-#' max_records(200)(women)
-#'
-#'
-#' @family helpers
-#'
 #' @export
-max_records <- function(n){
-  max_rec <- n
-  function(d) if (nrow(d) >n) FALSE else TRUE
+min_complete <- function(n, vars=TRUE){
+  stopifnot(is.numeric(n), n>=0
+          , isTRUE(vars) || 
+            is.character(vars) || 
+            is.numeric(vars) || 
+            is.logical(vars))
+  min_compl <- n
+  vars <- vars
+  function(d) sum(complete_cases(d[vars])) >= min_compl
 }
 
-#' Demand any number of expressions to be TRUE
+#' Demand minimal fraction of complete records
 #'
-#' @param ... comma-separated list of expressions that must all evaluate to \code{TRUE}
-#'            in order for a data set to be acceptable.
+#' @param r Minimal fraction of records that must be complete.
+#' @param vars \code{[TRUE|column index]} Column index into the data to
+#'        be tested (e.g. a character vectod with variable names or
+#'        a numeric vector with column positions). The indexed columns
+#'        will be testsed for completeness (absence of \code{NA}). Be default
[…]

### 2023-01-03 pkg/inst/tinytest/test_helpers.R

@@ -1,9 +1,54 @@
 
-expect_equal(n_complete(c(1,NA,2), 1:3), 2)
-expect_equal(fraction_complete(c(1,NA,2), 1:3), 2/3)
-expect_equal(n_incomplete(c(1,NA,2), 1:3), 1)
-expect_equal(fraction_incomplete(c(1,NA,2), 1:3), 1/3)
-
 expect_error(accumulate:::stopf("foo %g",bar))
 expect_equal(accumulate:::last(1:3),3)
+
+
+d <- data.frame(x=c(1:5,rep(NA,5)), y=1:10)
+expect_equal(accumulate:::complete_cases(d)
+           , c(rep(TRUE,5), rep(FALSE,5)))
+
+
+
+expect_true( min_records(8)(d))
+expect_false(min_records(12)(d))
+expect_error(min_records(-1))
+expect_error(min_records("foo"))
+
+expect_true( min_complete(3)(d))
+expect_false(min_complete(8)(d))
+expect_true( min_complete(8,vars="y")(d))
+expect_false(min_complete(8,vars="x")(d))
+expect_error(min_complete(n="foo"))
+
+
+expect_true( frac_complete(0.4)(d))
+expect_false(frac_complete(0.8)(d))
+expect_true( frac_complete(0.8, vars="y")(d))
+expect_false(frac_complete(0.8, vars="x")(d))
+expect_error(min_complete(r="foo"))
+
+
+
+
+csh <- csh_from_digits(c("11","12","123"))
+expect_equal(csh_from_digits(c("11","12"))
+            , data.frame(A0=c("11","12"), A1=c("1","1")))
+expect_equal(csh_from_digits(c("11","12"), levels=0)
+            , data.frame(A0=c("11","12")))
+
+expect_error(csh_from_digits(c(11,12), levels=-1))
+expect_error(csh_from_digits(c(11,12), levels= 2))
+
+#from_validator
+
+if (!requireNamespace('validate', quietly=TRUE)) exit_file("validate not installed")
+
+expect_true(from_validator( validate::validator(sum(is.na(x)) < 8) )(d))
+expect_false(from_validator( validate::validator(sum(is.na(x)) < 4) )(d))
+
+
+
+
+
+
 

### 2023-01-03 paper/article.Rnw
Writing in progress

@@
 \subsection{Specifying tests}
+The \code{test} parameter of \code{cumulate()} and \code{accumulate} accepts a
+function that takes a subset of the data and must returns a boolean. In
+principle this gives users enough flexibility to specify any possible test to
+determine whether a collapsing step is necessary or not. However, the package
+exports a few utility functions that make common test conditions easier to
+express without the need for writing a function. Moreover, to support users
+that are not comfortable with function writing, there is a higher order
+function called \code{demand()} that makes expressing combinations of
+tests easier.
+
+Common test conditions include demanding a minimal number of records, or a
+minimal number or fraction of complete records. These cases can be expressed
+with helper functions.
+%
+\begin{center}
+\begin{tabular}{ll}
+\code{min_records(n,vars)}   & At least $n$ records for variables \code{vars}\\
+\code{min_complete(n,vars)}   & At least $n$ records complete records for variables \code{vars}\\
+\code{frac_complete(r,vars)} & At least $100r$\% complete records for variables \code{vars}\\
+\end{tabular}
+\end{center}
 

### 2023-01-04 pkg/R/helpers.R
added smoketest function

@@ -1,4 +1,5 @@
 
+catf  <- function(fmt,...) cat(sprintf(fmt,...))
 stopf <- function(fmt,...) stop(sprintf(fmt,...), call.=FALSE)
 
 last <- function(x) x[length(x)]
@@ -141,6 +142,104 @@ csh_from_digits <- function(x, levels=max(nchar(x))-1){
   colnames(A) <- sprintf("A%d",seq_len(nlevels)-1)
   as.data.frame(A)[1:(levels+1)]
 }
+
+#' Test your testing function for common edge cases
+#'
+#' Writing a testing function that works on any subset of records of a
+#' dataframe can be quite subtle. This function tries the testing function on a
+#' number of common (edge) cases that are easily overlooked.  It is \emph{not}
+#' a unit test: a smoke test will not tell you whether your output is correct.
+#' It only checks the output data type (mustbe \code{TRUE} or \code{FALSE} and
+#' reports if errors, warnings, or messages occur.
+#'
+#' @param dat an example dataset. For example the full dataset
+#' to be fed into \code{\link{accumulate}} or \code{\link{cumulate}}.
+#' @param test A testing function to be passed as argument to \code{\link{accumulate}}
+#' or \code{\link{cumulate}}.
+#' @param halt \code{[logical]} toggle stopping when an error is thrown
+#'
+#' @return \code{NULL}, invisibly. This function has as side-effect that test
+#' results are printed to screen.
+#'
+#'
+#' @examples
+#' dat <- data.frame(x = 1:5, y=(-2):2)
+#' smoke_test(dat, function(d) y > 0)   #error: Y not found
+#' smoke_test(dat, function(d) d$y > 0) # issue: output too long, not robust against NA
+#' smoke_test(dat, function(d) sum(d$y > 0) > 2) # issue: not robust against NA
+#' smoke_test(dat, function(d) sum(d$y > 0, na.rm=TRUE) > 2) # OK
+#'
+#' @export
+smoke_test <- function(dat, test, halt=TRUE){
+  try_this(dat, test, info="full dataset") || !halt || return(invisible())
+  try_this(dat[1,,drop=FALSE], test, info="first record") || !halt || return(invisible())
+  try_this(dat[0,,drop=FALSE], test, info="zero records") || !halt || return(invisible())
+
+  vars <- colnames(dat)
[…]

### 2023-01-04 pkg/inst/tinytest/test_helpers.R

@@ -47,6 +47,14 @@ expect_true(from_validator( validate::validator(sum(is.na(x)) < 8) )(d))
 expect_false(from_validator( validate::validator(sum(is.na(x)) < 4) )(d))
 
 
+dat <- data.frame(x = 1:5, y=(-2):2)
+expect_stdout(smoke_test(dat, function(d) y > 0), pattern="ERR")  
+expect_stdout(smoke_test(dat, function(d) d$y > 0),pattern="length") 
+expect_stdout(smoke_test(dat, function(d) sum(d$y > 0) > 2), pattern="NA detected")
+expect_stdout(smoke_test(dat, function(d) sum(d$y > 0, na.rm=TRUE) > 2),pattern="OK")
+
+
+
 
 
 

### 2023-01-04 paper/article.Rnw
writing in progress

--- 2023-01-02 paper/article.Rnw
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The formula interface}
+\subsection{The formula interface}\label{sect:formula}
 We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
@@
 <<>>=
-input$AB <- paste(input$A, input$B, sep="-")
-input <- input[-(1:3)]
-input
+input1 <- input
+input1$AB <- paste(input$A, input$B, sep="-")
+input1 <- input1[-(1:3)]
+input1
 @
@@
 <<>>=
-accumulate(input, collapse=csh, test = min_records(3), sum)
-cumulate(input, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
+accumulate(input1, collapse=csh, test = min_records(3), sum)
+cumulate(input1, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
 @
@@
 \subsection{Specifying tests}
+The \code{test} parameter of \code{cumulate()} and \code{accumulate} accepts a
+function that takes a subset of the data and must returns a boolean. For common
+test conditions, including requiring a minimal number of records, or a minimal
+number or fraction of complete records there are helper functions available.
+%
+\begin{center}
+\begin{tabular}{ll}
+\code{min_records(n)}        & At least $n$ records.\\ 
+\code{min_complete(n,vars)}  & At least $n$ records complete records for variables \code{vars}.\\
+\code{frac_complete(r,vars)} & At least $100r$\% complete records for variables \code{vars}.\\
+\end{tabular}
+\end{center}
+
+Second, for multiple, possibly complex requirements on variables users can
+express conditions with the \pkg{validate} package \cite{loo2021data}.  The
+\pkg{validate} packages offers a domain-specific language for expressing,
+manipulating, and investigating conditions on datasets. It's core concept is a
+list of `data validation rules' stored as a \class{validator} object. A
+\class{validator} object is constructed with the enopymous function
+\code{validator()}.  For example, to demand that there are at least 3 rows in a
+group, and that there are at least three records where $Y\geq 2$ we create the
+following ruleset.
+<<>>=
+library(validate)
+rules <- validator(nrow(.) >= 3, sum(Y >= 2) >= 3)
+rules
+@
+Here the \code{.} refers to the dataset as a whole, while rules that can be
+evaluated within the dataset can be written as boolean \proglang{R}
+expressions.
+
+We will apply these conditions to the \code{input} dataset that was 
+constructed in Section~\ref{sect:formula}. As a reminder we print
+the first few records.
+<<>>=
+head(input, 4)
+@
+We use \code{A*B ~ A*B1 + B1} as collapsing scheme. The function
+\code{from_validator} passes the requirements as a test function to
+\code{accumulate()} (or \code{cumulate()}.
+<<>>=
+accumulate(input, collapse = A*B ~ A*B1 + B1,
+    test = from_validator(rules), fun=sum)
+@
+Note that for target groups $(A=2,B=21)$ and $(A=2,B=22)$ none of the available
+collapsing levels lead to a group that satisfied all conditions. Therefore the
+collapsing level and output variables are all missing (\code{NA}).
+
+The third and most flexible way for users to express tests is to write a custom
+testing function. The requirements are that it must work on any subset of a
+data frame, including a dataset with zero rows. The previous example can
+thus also be expressed as follows.
+<<>>=
+my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >=3
+accumulate(input, collapse = A*B ~ A*B1 + B1, test=my_test, fun=sum)
+@
 

### 2023-01-04 pkg/R/helpers.R
fix

@@ -153,9 +153,11 @@ csh_from_digits <- function(x, levels=max(nchar(x))-1){
 #' reports if errors, warnings, or messages occur.
 #'
 #' @param dat an example dataset. For example the full dataset
-#' to be fed into \code{\link{accumulate}} or \code{\link{cumulate}}.
+#'        to be fed into \code{\link{accumulate}} or \code{\link{cumulate}}.
 #' @param test A testing function to be passed as argument to \code{\link{accumulate}}
-#' or \code{\link{cumulate}}.
+#'        or \code{\link{cumulate}}.
+#' @param verbose \code{[logical]} If \code{TRUE}, all results (including
+#'        passed tests) are printed.  If \code{FALSE} only failed tests are printed.
 #' @param halt \code{[logical]} toggle stopping when an error is thrown
 #'
 #' @return \code{NULL}, invisibly. This function has as side-effect that test
@@ -170,28 +172,39 @@ csh_from_digits <- function(x, levels=max(nchar(x))-1){
 #' smoke_test(dat, function(d) sum(d$y > 0, na.rm=TRUE) > 2) # OK
 #'
 #' @export
-smoke_test <- function(dat, test, halt=TRUE){
-  try_this(dat, test, info="full dataset") || !halt || return(invisible())
-  try_this(dat[1,,drop=FALSE], test, info="first record") || !halt || return(invisible())
-  try_this(dat[0,,drop=FALSE], test, info="zero records") || !halt || return(invisible())
+smoke_test <- function(dat, test, verbose=FALSE, halt=TRUE){
+  try_this(dat, test, verbose, info="full dataset") || !halt || return(invisible())
+  try_this(dat[1,,drop=FALSE], test, verbose, info="first record") || !halt || return(invisible())
+  try_this(dat[0,,drop=FALSE], test, verbose, info="zero records") || !halt || return(invisible())
 
   vars <- colnames(dat)
   for (var in vars){
     d <- dat[1,,drop=FALSE]
     d[1,var] <- NA
-    try_this(d, test, info=sprintf("first record and %s is NA",var)) || !halt || return(invisible())
+    try_this(d, test, verbose
+     , info=sprintf("first record and %s is NA",var)) || !halt || return(invisible())
   } 
 
   d <- dat[1,,drop=FALSE]
   d[1,] <- NA
[…]

### 2023-01-04 pkg/R/helpers.R
typos

@@ -188,14 +188,14 @@ smoke_test <- function(dat, test, verbose=FALSE, halt=TRUE){
   d <- dat[1,,drop=FALSE]
   d[1,] <- NA
   try_this(d, test, verbose
-    , info="first record and all values NA") || !halt || return(invisiblt()) 
+    , info="first record and all values NA") || !halt || return(invisible()) 
 
   for (var in vars){
     d <- dat
     d[,var] <- NA
     try_this(d, test, verbose
             , info=sprintf("full dataset and %s is NA for all records",var)) || 
-      !halt || return(invisible)
+      !halt || return(invisible())
   }
 
 

### 2023-01-04 pkg/inst/tinytest/test_helpers.R

@@ -51,7 +51,7 @@ dat <- data.frame(x = 1:5, y=(-2):2)
 expect_stdout(smoke_test(dat, function(d) y > 0), pattern="ERR")  
 expect_stdout(smoke_test(dat, function(d) d$y > 0),pattern="length") 
 expect_stdout(smoke_test(dat, function(d) sum(d$y > 0) > 2), pattern="NA detected")
-expect_stdout(smoke_test(dat, function(d) sum(d$y > 0, na.rm=TRUE) > 2),pattern="OK")
+expect_stdout(smoke_test(dat, function(d) sum(d$y > 0, na.rm=TRUE) > 2),pattern="")
 
 
 

### 2023-01-04 paper/article.Rnw
writing

@@
 
+It is easy to overlook some edge cases when specifying test functions.
+Recall that a test function is required to return \code{TRUE} or \code{FALSE},
+regardless of the data circumstances. The only thing that a test function
+can assume is that the received data set is a subset of records from the
+dataset to be aggregated. As a service to the user, \pkg{accumulate} exports
+a function that checks a test function against common edge cases, including
+the occurrence of missing values, a dataset with zero rows, and the full
+dataset. The smoke test checks whether the output is \code{TRUE} or \code{FALSE}
+under all circumstances and also reports erros, warnings, and messages.
+It accepts a (realistic) dataset and a testing function and prints
+test results to the console.
+<<>>=
+smoke_test(input, my_test)
+@
+By default only failing tests are printed. In this case our test function
+is not robust against missing values for $Y$. This can be remedied
+by adding \code{na.rm=TRUE} to the \code{sum()} in the test function.
+<<>>=
+my_test1 <- function(d) nrow(d) >= 3 && sum(d$Y>=2, na.rm=TRUE) >=3 
+smoke_test(input, my_test1)
+@
+The smoke test is aimed at prevening complicated stack traces when errors occur
+in a call to \code{accumulate()} or \code{cumulate()}.  Users should be aware
+that it does not guarantee correctness of the results, only robustness against
+certain edge cases.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Programming over accumulate }
-
 \section{Extensive example: economic data} \label{sect:example}

### 2023-01-05 pkg/R/helpers.R
documented output, separated internal utility functions

@@ -1,12 +1,4 @@
 
-catf  <- function(fmt,...) cat(sprintf(fmt,...))
-stopf <- function(fmt,...) stop(sprintf(fmt,...), call.=FALSE)
-
-last <- function(x) x[length(x)]
-
-complete_cases <- function(d) !Reduce(`|`, lapply(d, is.na))
-
-
 
 #' Demand minimal number of functons
 #' 
@@ -16,6 +8,10 @@ complete_cases <- function(d) !Reduce(`|`, lapply(d, is.na))
 #'
 #' min_records(5)(women)
 #' min_records(200)(women)
+#'
+#' @return a function that accepts a data frame and returns \code{TRUE}
+#'         when the number of records is larger than or equal to \code{n}
+#'         and otherwise \code{FALSE}.
 #'
 #' @family helpers
 #'
@@ -34,6 +30,10 @@ min_records <- function(n){
 #'        a numeric vector with column positions). The indexed columns
 #'        will be testsed for completeness (absence of \code{NA}). Be default
 #'        \code{vars=TRUE} meaning that all columns are taken into account.
+#'
+#' @return a function that accepts a data frame and returns \code{TRUE}
+#'         when the number of complete records is larger than or equal to \code{n}
+#'         and otherwise \code{FALSE}.
 #'
 #' @family helpers
 #'
@@ -59,6 +59,10 @@ min_complete <- function(n, vars=TRUE){
 #'        a numeric vector with column positions). The indexed columns
 #'        will be testsed for completeness (absence of \code{NA}). Be default
 #'        \code{vars=TRUE} meaning that all columns are taken into account.
+#'
+#' @return a function that accepts a data frame and returns \code{TRUE} when the
+#'         fraction of complete records is larger than or equal to \code{n} and
+#'         otherwise \code{FALSE}.
 #'
 #' @family helpers
 #'
@@ -87,6 +91,11 @@ frac_complete <- function(r, vars=TRUE){
 #'
 #' @note
 #' Requires the \code{validate} package to be installed.
+#'
+#' @return a function that accepts a data fram and returns \code{TRUE}
+#'         when the data passes all checks in \code{v} and otherwise
+#'         \code{FALSE}.
+#'
 #'
 #' @references
 #' Mark P. J. van der Loo, Edwin de Jonge (2021). Data Validation
@@ -118,6 +127,10 @@ from_validator <- function(v,...){
 #' @param x \code{[character|integer]} labels in a hierarchical classification (lowest level)
[…]

### 2023-01-05 pkg/R/utils.R

@@ -1,0 +1,9 @@
+
+catf  <- function(fmt,...) cat(sprintf(fmt,...))
+stopf <- function(fmt,...) stop(sprintf(fmt,...), call.=FALSE)
+
+last <- function(x) x[length(x)]
+
+
+complete_cases <- function(d) !Reduce(`|`, lapply(d, is.na))
+

### 2023-01-05 paper/article.Rnw
writing

@@
 \section{Extensive example: economic data} \label{sect:example}
+In this section we treat three practical examples using a synthetic
+dataset included with the package.
+<<>>=
+data(producers)
+head(producers)
+@
+This \code{producers} dataset contains synthetic data records of various income
+sources for \Sexpr{nrow(producers)} industrial produces. The records are
+classified into a local version of the NACE classification called \code{sbi}
+and into \code{size} classes with values in $\{5,6,7,8,9\}$.
+
+\subsection{Small area estimation} 
+Small area estimation (SAE) is a collection of methods that are aimed at
+estimating subpopulation parameters in cases where there the number of
+observations in a subpopulation is so mall that direct estimation leads to
+unacceptable estimation variance. Instead one resorts to indirect estimation,
+meaning that one estimates the parameter for a larger subpopulation that
+includes the target subpopulation. This estimate is then interpreted as an
+estimate for the target subpopulation. Here, we shall be interested in
+estimating the average turnover from industrial activities (\code{industrial})
+by SBI and size class.
+
+In the simplest case, where no auxiliary information is available or used, one
+replaces the estimator of the mean over a subpopulation with the estimator of
+the mean over a larger subpopulation. If we assume that all units in the
+dataset were sampled with equal probability based on a simple random sampling
+design, the mean is estimated with the sample mean. We will demand that there
+are at least ten records for which turnover has been measured. The collapsing
+scheme is given by \code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and
+\code{sbi1} are classification by respectively the first two SBI digits and the
+first SBI digit. We first add those variables to the dataset.
+<<>>=
+producers$sbi2 <- substr(producers$sbi,1,2)
+producers$sbi1 <- substr(producers$sbi,1,1)
+head(producers, 3)
+@
+Our test function looks as follows.
+<<>>=
+testfun <- function(d) sum(!is.na(d$industrial)) >= 10
+@
+Using \code{accumulate()} we obtain the means. 
+<<>>=
+a <- cumulate(producers
+             , collapse = sbi*size ~ sbi + sbi2 + sbi1
+             , test = testfun
+             , mean_industrial = mean(industrial, na.rm=TRUE))
+head(a,3)
+@
+
+\subsection{Imputing missing values using SAE and ratio imputation} 
+Our goal in this example is to impute missing values for the \code{industrial}
+variable based on ratio imputation with \code{total} as predictor.  Ratio
+imputation is a method where the imputed value $\hat{y}_i$ for variable $Y$ of
+record $i$ is estimated as $\hat{y_i}=\hat{R}_d x_i$, where $\hat{R}$ is an
+estimate of the ratio between $Y$ and an auxiliary variable $X$ in
+subpopulation $d$. An unbiased estimate for $R_d$ is given by
+$\hat{\bar{Y}}_d/\hat{\bar{X}}_d$ where $\hat{\bar{Y}}_d$ and $\hat{\bar{X}}_d$
+are estimated subpopulation means.
+
+We use SAE to estimate the subpopulation ratios, and then use the \code{simputation}
+package \citep{loo2022simputation} to impute the missing values. 
+<<>>=
+r <- cumulate(producers
+         , collapse = sbi*size ~ sbi + sbi2 + sbi1
+         , test = testfun
+         , R = mean(industrial, na.rm=TRUE)/mean(total, na.rm=TRUE))
+head(r,3)
+@
+To impute the values, using \code{impute_proxy()} we need to merge the ratios
+with the producers dataset (which automatically happens by SBI and size class).
+<<>>=
+library(simputation)
+dat <- merge(producers, r)
+dat <- impute_proxy(dat, industrial ~ R*total)
+@
+We can inspect the imputed values as follows.
+<<>>=
+iNA <- is.na(producers$industrial)
+head(dat[iNA, c("sbi","size","level","R","industrial","total")])
+@
+From this, we have all the information to interpret the imputed values. For
+example, we see that in record 664, the ratio was estimated after collapsing
+the group $(\texttt{sbi},\texttt{size})=(21122,7)$ to $\texttt{sbi2}=21$ since
+the collapse \code{level} equals 2.
 
 \section{Formal descriptions and algorithms}
-In this section we give a formal description of the algorithms. In order to be
-unambiguous and to represent the algorithms compactly, we introduce some
-notation. The notation also makes the pseudocode independent of the
-implementation language used, or how the data is represented.
-
-We start by giving the algorithm for ordinary split-apply-combine, so it is
-easy to see where the algorithm must be generalized to allow for a collapsing
-scheme.
+In this section we give a formal description of the algorithm for aggregation
+with dynamic grouping.  We start by giving an algorithm for ordinary
+split-apply-combine to show where the algorithm must be generalized to allow
+for a collapsing scheme.
 
@@
 label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
-groups, since every element in $U$ can only get one label.  In formal
-terminology we say that $f:U\to A$ is a \emph{partition} of $U$.
+groups. We say that $f:U\to A$ is a \emph{partition} of $U$.
 
@@
         A\times X$ }
-
 $R = \{\}$\;
@@
 Here, the output is collected in a set $R$ (this set is technically a function
-$A\to X$ since it pairs for each $a\in A$ with one element of $X$). We also
-introduced some new notation: the notation $2^U$ stands for `the set of all
-subsets of $U$'. The notation $f^{-1}$ stands for the formal \emph{inverse} of
-$f$. That is: you give it a set of labels from $A$ and it returns all the
-elements of $U$ that have any of those labels. This definition is slightly too
-general for the current algorithm, but it will come in handy when we move to
-collapsing groups.
+$A\to X$ since it pairs each $a\in A$ with one element of $X$). We also
+introduced some new notation: the notation $2^U$ is the power set of $U$ and
+$f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you give it a
+set of labels from $A$ and it returns all the elements of $U$ that have any of
+those labels. This definition is somewhat too general for the current
+algorithm, but it will be necessary when we move to collapsing groups.
 
-Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented
-in practice. Consider the signature of the R \code{aggregate()} function for data
+Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented in
+practice. Consider the signature of the R \code{aggregate()} function for data
 frames (we skip arguments that are not important for the discussion).
@@
 
-Collapsing schemes can be represented as tables. For example, the European NACE
-classification is a hierarchical classification of economic activity. Each
-group is encoded as a 4-digit code where the first digit indicates the highest
-grouping level, and the last digit the most detailed level. This induces a
-collapsing sequence by combining groups that have the same parent in the
-hierarchy. Below is an example, based on a small piece of the NACE code list.
+%Collapsing schemes can be represented as tables. For example, the European NACE
+%classification is a hierarchical classification of economic activity. Each
+%group is encoded as a 4-digit code where the first digit indicates the highest
+%grouping level, and the last digit the most detailed level. This induces a
+%collapsing sequence by combining groups that have the same parent in the
+%hierarchy. Below is an example, based on a small piece of the NACE code list.
+%%
+%\begin{equation}
+%\begin{tabular}{rrr}
+%\hline
+% $A_0$& $A_1$ & $A_2$\\
+%\hline
+%0111&   011 & 01\\
+%0112&   011 & 01\\
+%0113&   011 & 01\\
+%0121&   012 & 01\\
+%0122&   012 & 01\\
+%0123&   012 & 01\\
+%0124&   012 & 01\\
+%\hline
+%\end{tabular}
+%\label{eq:tablerep}
+%\end{equation}
 %
-\begin{equation}
-\begin{tabular}{rrr}
-\hline
- $A_0$& $A_1$ & $A_2$\\
-\hline
-0111&   011 & 01\\
-0112&   011 & 01\\
-0113&   011 & 01\\
-0121&   012 & 01\\
-0122&   012 & 01\\
-0123&   012 & 01\\
-0124&   012 & 01\\
-\hline
-\end{tabular}
-\label{eq:tablerep}
-\end{equation}
+%Here, $A_0$ is the most detailed partition, splitting a dataset into seven
+%groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
+%The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
+%$0113$ and the second, labeled $01.2$ is constructed by combining
+%$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.
+%
+%To compute an aggregate for records labeled with $0111$, the algorithm will
+%select those records and test whether the subset is suitable. If not, it will
+%move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
+%follows.
+%\begin{displaymath}
+%f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
+%\end{displaymath}
+%It will then move on and find all records labeled with any of those three
+%group labels, and test again. If this new data set is suitable, the result is
+%computed, if not, another collapse takes place. The whole procedure is repeated 
+%for each element in $A_0$.
 
-Here, $A_0$ is the most detailed partition, splitting a dataset into seven
-groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
-The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
-$0113$ and the second, labeled $01.2$ is constructed by combining
-$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.
-
-To compute an aggregate for records labeled with $0111$, the algorithm will
-select those records and test whether the subset is suitable. If not, it will
-move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
-follows.
-\begin{displaymath}
-f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
-\end{displaymath}
-It will then move on and find all records labeled with any of those three
-group labels, and test again. If this new data set is suitable, the result is
-computed, if not, another collapse takes place. The whole procedure is repeated 
-for each element in $A_0$.
-
-In order to generalize this algorithm towards generic collapsing schemes, we
-introduce the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and
+Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
 returns the corresponding label in $A_k$. This can formally be defined as

### 2023-02-09 paper/article.Rnw
start cleaning code. Using explicit pullback function now

@@
 to define any quality requirement on the grouping while common quality
-requiremtns are supported out of the box. Common quality requirements include a
-minimum number of records, or a minum fraction of records without missing
+requirements are supported out of the box. Common quality requirements include a
+minimum number of records, or a minimum fraction of records without missing
 values, or a minimum number of records with non-zero values.
@@
 $B$ is a hierarchical classification with levels $\{11,12,13,21,22\}$. Variable
-$B_1$ is a coursegraining of $B$: for each record the value for $B_1$ is the
+$B_1$ is a course graining of $B$: for each record the value for $B_1$ is the
 first digit of $B$. Hence, $B_1$ has levels $\{1,2\}$.
@@
 $t_Y=8+16+32=56$.  Similarly, there is only one record with $(A=2,B=13)$. 
-Collapsing groups to $(A=2,B1=1)$, yieds again $t_y=56$.
+Collapsing groups to $(A=2,B1=1)$, yields again $t_y=56$.
 
@@
 represents an aggregation function that is applied to every column of
-\code{data}, and the ellepsis (\code{...}) is for arguments that are passed as
+\code{data}, and the ellipsis (\code{...}) is for arguments that are passed as
 extra argument to \code{fun}. The interface of \code{accumulate()} is modeled
 after the \code{aggregate()} function in \proglang{R}.  In \code{cumulate}, the
-ellepsis is a sequence of comma-separated \code{name=expression} pairs in the
+ellipsis is a sequence of comma-separated \code{name=expression} pairs in the
 style of \code{summarise()} from the \code{dplyr} package.
@@
 The left-hand-side \code{A*B} is the target output grouping. The
-ritght-hand-side is to be interpreted as the collapsing sequnce: if an instance
+right-hand-side is to be interpreted as the collapsing sequence: if an instance
 of \code{A*B} does not pass the test, then collapse to \code{A*B1}, and if that
@@
 
-Summarizing, the formula interface is always of the follwing form.
+Summarizing, the formula interface is always of the following form.
 \begin{verbatim}
@@
 
-Hierarchical classifications offer a natural mechanism for collapsing fine-graind groupings
-into larger groups because of the parent-child relationships. As an example consider
-a small piece of the NACE classification.
+Hierarchical classifications offer a natural mechanism for collapsing
+fine-grained groupings into larger groups because of the parent-child
+relationships. As an example consider a small piece of the NACE classification.
 
@@
 would combine $\{012\}$ with $\{011\}$ into $\{01\}$. The \pkg{accumulate}
-package comes with a helper function that creates the collasping scheme from the
+package comes with a helper function that creates the collapsing scheme from the
 lowest-level digits.
@@
 
-The situation becomes a little more involved when hierarchical classifcations
+The situation becomes a little more involved when hierarchical classifications
 form a tree such that the distance from leave to trunk is not the same for all
@@
 calculation as well as the interpretation of the results are now uniform
-accross all hierarchical classifications. Again, using \code{csh_from_digits()}
+across all hierarchical classifications. Again, using \code{csh_from_digits()}
 deriving the collapsing scheme can be automated.
@@
 list of `data validation rules' stored as a \class{validator} object. A
-\class{validator} object is constructed with the enopymous function
+\class{validator} object is constructed with the eponymous function
 \code{validator()}.  For example, to demand that there are at least 3 rows in a
@@
 \code{from_validator} passes the requirements as a test function to
-\code{accumulate()} (or \code{cumulate()}.
+\code{accumulate()} (or \code{cumulate()}).
 <<>>=
@@
 dataset. The smoke test checks whether the output is \code{TRUE} or \code{FALSE}
-under all circumstances and also reports erros, warnings, and messages.
+under all circumstances and also reports errors, warnings, and messages.
 It accepts a (realistic) dataset and a testing function and prints
@@
 @
-The smoke test is aimed at prevening complicated stack traces when errors occur
+The smoke test is aimed at preventing complicated stack traces when errors occur
 in a call to \code{accumulate()} or \code{cumulate()}.  Users should be aware
@@
 
-\subsection{Performance, memory, paralellization}
+\subsection{Performance, memory, parallelization}
 Both Algorithms~\ref{alg:sac} and \ref{alg:saccg} need only $\mathcal{O}(|U|)$
@@
 groupwise aggregation for some commonly occurring aggregating functions.
-Generic paralellization of Algorithm~\ref{alg:sac} can be done by paralellising
+Generic parallelization of Algorithm~\ref{alg:sac} can be done by parallelizing
 over the grouping values $a\in A$: each worker node gets access to a set of
@@
 The fact that we cannot generically exclude records from a calculation means
-that a generic paralellisation scheme would need to ensure that each worker
+that a generic parallelization scheme would need to ensure that each worker
 node has access to the full dataset $|U|$. There are two types of options to do
-this: either mutliple workers gain read-only access to $U$ or each worker gets
+this: either multliple workers gain read-only access to $U$ or each worker gets
 a copy of $U$. In the first case the workers may suffer from colliding
@@
 higher worst-case complexity than regular split-apply-combine. Moreover,
-generic paralellization needs to either consider a specific use case (e.g.
+generic parallelization needs to either consider a specific use case (e.g.
 large or small datasets) or require user input that guides the choice of
-paralellisation scheme.
+parallelization scheme.
 

### 2023-02-09 pkg/R/accumulate.R

@@ -145,6 +145,7 @@
 #'
 #' @export
 accumulate <- function(data, collapse, test, fun, ...){
+  # poor man's dispatch, on second argument.
   if (inherits(collapse, "data.frame")){
     accumulate1(as.data.frame(data), as.data.frame(collapse), test, fun, ...)
   } else if (inherits(collapse, "formula")){
@@ -158,46 +159,38 @@ accumulate <- function(data, collapse, test, fun, ...){
 
 accumulate1 <- function(data, collapse, test, fun, ...){
 
+
   # Target agregation level
   g  <- colnames(collapse)[1]
   ig <- colnames(data) == g
-  g_type <- last(class(data[,g]))
-
 
   if (!any(ig) ){
     stopf("First column '%s' in 'collapse' is not present in 'data'",g)
   }
-
-
-  collapse <- as.data.frame(lapply(collapse, as.character))
-  data[,g] <- as.character(data[,g])
-  rownames(collapse) <- collapse[,1] # for lookup
-
+  
  
   out <- data.frame(unique(data[,g]), level=NA_integer_) 
   names(out)[1] <- g
-  
   vals <- as.data.frame( matrix(NA, nrow = nrow(out), ncol = sum(!ig)) )
   colnames(vals) <- colnames(data)[!ig]
   out <- cbind(out, vals)
 
+  pullback <- get_pb(collapse, data)
+
   for ( ia in seq_len(nrow(out)) ){
     j = 1
     out_level <- out[ia,1]
-    d <- data[data[,g] == out_level,!ig,drop=FALSE]
+    d <- data[pullback(out_level, j-1), !ig,drop=FALSE]
     while( j < ncol(collapse) && !test(d) ){
       j <- j + 1
-      backmap <- collapse[collapse[,j] == collapse[out_level,j], 1]
-      d <-  data[data[,g] %in% backmap, !ig, drop=FALSE]
+      d <- data[pullback(out_level, j-1), !ig, drop=FALSE]
     }
-
 
     if ( j < length(collapse) || test(d)){
       out[ia, 2]           <- j-1
       out[ia, 3:ncol(out)] <- lapply(d[!ig], fun, ...)
     }
   }
-  out[,1] <- do.call(paste0("as.",g_type), list(out[,1]))
   out
 
 }
@@ -217,15 +210,14 @@ accumulate2 <- function(data, formula, test, fun, ...){
[…]

### 2023-02-09 pkg/R/utils.R

@@ -15,3 +15,66 @@ format.listcol <- function(x,...){
   sapply(x, function(u) sprintf("<%s>",paste(class(u),collapse=",")))
 }
 
+
+# get_pb: create pullback function.
+# x  : formula or data.frame
+# dat: data to create inverse mapping for.
+# out: function accepting (a) group label(s) and a collapse level
+#      that returns an integer row index for 'dat'
+get_pb <- function(x, dat){
+  # Poor man's dispatch since we don't want to export this function
+  t <- last(class(x))
+  switch(t
+    , "data.frame" =  get_pb.data.frame(x,dat)
+    , "formula"    =  get_pb.formula(x, dat)
+    , stop("'x' must be a data frame or a formula")
+  )
+
+}
+
+
+get_pb.data.frame <- function(x, dat){
+  group_labels <- dat[,colnames(x)[1]]
+  rownames(x) <- x[,1]
+
+
+  function(group, level){ 
+    which(group_labels %in% x[x[,level+1] == x[as.character(group), level+1],1])
+  }
+
+}
+
+get_pb.formula <- function(x, dat){
+  
+  collapse <- c(x[[2]], get_collapse(x[[3]]))
+
+
+  labels <- dat[all.vars(x)]
+  # add column with row numbers; make sure column name does not exist already.
+  iname <- "__idx__"
+  n <- 0
+  while( iname %in% names(labels) ){ 
+      n <- n + 1
+      iname <- sprintf("%s%i",iname,n)
+  }
+  labels[,iname] <- seq_len(nrow(labels))
+ 
+  function(group, level){
+    d <- merge(group, labels, by=names(group))
+    collapse_value <- d[1, all.vars(collapse[[level+1]]),drop=FALSE]
+    merge(labels, collapse_value, by=names(collapse_value))[,iname]
+  }
+
+}
+
+
+
+get_collapse <- function(e, L = list()){
+  if (length(e) == 1 || e[[1]]=="*") return(append(L, e))
+  c(get_collapse(e[[2]],L), get_collapse(e[[3]],L))
+}
+
+
+
+
+

### 2023-02-09 pkg/inst/tinytest/test_internals.R
added test file

@@ -1,0 +1,45 @@
+#source("../../R/acc.R")
+#library(tinytest)
+# 
+#
+input    <- data.frame(Y1 = 2^(0:8), Y2 = 2^(0:8))
+input$Y2[c(1,4,7)] <- NA
+input$A0 <- c(123,123,123,135,136,137,212,213,225)
+
+
+collapse <- data.frame(
+      A0   = c(123, 135, 136, 137, 212, 213, 225)
+    , A1   = c(12 , 13 , 13 , 13 , 21 , 21 , 22 )
+    , A2   = c(1  , 1  , 1  , 1  , 2  , 2  , 2  )
+)
+
+pullback <- accumulate:::get_pb(collapse, input)
+
+expect_equal(pullback(123,0), 1:3)
+expect_equal(pullback(135, 1), 4:6)
+expect_equal(pullback(212, 2), 7:9)
+
+
+input <- data.frame(
+    A  = c(1,1,1,2,2,2,3,3,3)
+  , B  = c(11,11,11,12,12,13,21,22,12)
+  , B1 = c(1,1,1,1,1,1,2,2,1)
+  , Y1 = 2^(0:8)
+  , Y2 = 2^(0:8)
+)
+
+input$Y2[c(1,4,7)] <- NA
+
+pullback <- accumulate:::get_pb(A*B ~ A*B1 + A, input)
+
+expect_equal(sort(pullback(data.frame(A=1,B=11), level=0)), 1:3)
+expect_equal(sort(pullback(data.frame(A=3,B=21), level=1)), 7:8)
+expect_equal(sort(pullback(data.frame(A=2,B=13), level=2)), 4:6)
+
+
+
+
+
+
+
+

### 2023-02-10 pkg/R/accumulate.R
internal code improvements

@@ -179,11 +179,11 @@ accumulate1 <- function(data, collapse, test, fun, ...){
 
   for ( ia in seq_len(nrow(out)) ){
     j = 1
-    out_level <- out[ia,1]
-    d <- data[pullback(out_level, j-1), !ig,drop=FALSE]
+    out_level <- out[ia,1,drop=FALSE]
+    d <- pullback(out_level, j-1)
     while( j < ncol(collapse) && !test(d) ){
       j <- j + 1
-      d <- data[pullback(out_level, j-1), !ig, drop=FALSE]
+      d <- pullback(out_level, j-1)
     }
 
     if ( j < length(collapse) || test(d)){
@@ -214,10 +214,10 @@ accumulate2 <- function(data, formula, test, fun, ...){
  
   for (ia in seq_len(nrow(out))){
     j <- 0
-    d <- data[pullback(out[ia, lhs_vars, drop=FALSE], level = j), , drop=FALSE]
+    d <- pullback(out[ia, lhs_vars, drop=FALSE], level = j)
     while(j < length(collapse) && !test(d)){
       j <- j + 1
-      d <- data[pullback(out[ia,lhs_vars,drop=FALSE], level = j), , drop=FALSE]
+      d <- pullback(out[ia,lhs_vars,drop=FALSE], level = j)
     }
 
     if ( j < length(collapse) || test(d)){
@@ -258,10 +258,10 @@ cumulate1 <- function(data, collapse, test, ...){
 
   for ( a in collapse[,1]){
     j <- 1
-    d <- data[pullback(a,j-1), , drop=FALSE]
+    d <- pullback(a,j-1)
     while ( j < ncol(collapse) && !test(d) ){
       j <- j + 1
-      d <- data[pullback(a,j-1), , drop=FALSE]
+      d <- pullback(a,j-1)
     }
     if ( j < ncol(collapse) || test(d)){
       out[a,2] <- j-1
@@ -290,10 +290,10 @@ cumulate2 <- function(data, formula, test,...){
  
   for (ia in seq_len(nrow(out))){
     j = 0
-    d <- data[pullback(out[ia,lhs_vars,drop=FALSE],j), , drop=FALSE]
+    d <- pullback(out[ia,lhs_vars,drop=FALSE],j)
     while(j < length(collapse) && !test(d)){
       j <- j + 1
-      d <- data[pullback(out[ia,lhs_vars,drop=FALSE],j), , drop=FALSE]
[…]

### 2023-02-10 pkg/R/utils.R

@@ -36,10 +36,12 @@ get_pb <- function(x, dat){
 get_pb.data.frame <- function(x, dat){
   group_labels <- dat[,colnames(x)[1]]
   rownames(x) <- x[,1]
-
-
+  
+  # group: a 1x1 data frame with a single group label 
+  # level: collapse level (column index-1 in x)
   function(group, level){ 
-    which(group_labels %in% x[x[,level+1] == x[as.character(group), level+1],1])
+    i <- group_labels %in% x[x[,level+1] == x[as.character(unclass(group)), level+1],1]
+    dat[i, , drop=FALSE]
   }
 
 }
@@ -47,7 +49,6 @@ get_pb.data.frame <- function(x, dat){
 get_pb.formula <- function(x, dat){
   
   collapse <- c(x[[2]], get_collapse(x[[3]]))
-
 
   labels <- dat[all.vars(x)]
   # add column with row numbers; make sure column name does not exist already.
@@ -62,17 +63,29 @@ get_pb.formula <- function(x, dat){
   function(group, level){
     d <- merge(group, labels, by=names(group))
     collapse_value <- d[1, all.vars(collapse[[level+1]]),drop=FALSE]
-    merge(labels, collapse_value, by=names(collapse_value))[,iname]
+    i <- merge(labels, collapse_value, by=names(collapse_value))[,iname]
+    dat[sort(i),,drop=FALSE]
   }
 
 }
-
-
 
 get_collapse <- function(e, L = list()){
   if (length(e) == 1 || e[[1]]=="*") return(append(L, e))
   c(get_collapse(e[[2]],L), get_collapse(e[[3]],L))
 }
+
+# function that retrieves the 
+get_ng <- function(x, dat){
+  if (inherits(x,"formula")){
+    lhs <- all.vars(x[[2]])
+    lbs <- unique(dat[lhs])
+    function(n) lbs[n,,drop=FALSE]
+  } else if (inherits(x, "data.frame")){
+    lbs <- dat[,1]
+    function(n) lbs[n]
+  }
+}
+
 
 
 

### 2023-02-10 pkg/inst/tinytest/test_internals.R

@@ -15,9 +15,9 @@ collapse <- data.frame(
 
 pullback <- accumulate:::get_pb(collapse, input)
 
-expect_equal(pullback(123,0), 1:3)
-expect_equal(pullback(135, 1), 4:6)
-expect_equal(pullback(212, 2), 7:9)
+expect_equal(pullback(123, 0), input[1:3,])
+expect_equal(pullback(135, 1), input[4:6,])
+expect_equal(pullback(212, 2), input[7:9,])
 
 
 input <- data.frame(
@@ -32,9 +32,9 @@ input$Y2[c(1,4,7)] <- NA
 
 pullback <- accumulate:::get_pb(A*B ~ A*B1 + A, input)
 
-expect_equal(sort(pullback(data.frame(A=1,B=11), level=0)), 1:3)
-expect_equal(sort(pullback(data.frame(A=3,B=21), level=1)), 7:8)
-expect_equal(sort(pullback(data.frame(A=2,B=13), level=2)), 4:6)
+expect_equal(pullback(data.frame(A=1,B=11), level=0), input[1:3,])
+expect_equal(pullback(data.frame(A=3,B=21), level=1), input[7:8,])
+expect_equal(pullback(data.frame(A=2,B=13), level=2), input[4:6,])
 
 
 

### 2023-02-20 pkg/R/accumulate.R
fix

@@ -372,7 +372,7 @@ output_backbone <- function(cps, dat){
 
 # get maximum number of collapsing steps (base 0).
 max_collapse <- function(cps){
-  if (inherits(cps,"formula")) length(cps[[3]]) else ncol(cps) - 1
+  if (inherits(cps,"formula")) length(get_collapse(cps[[3]])) else ncol(cps) - 1
 }
 
 # Get the variable names for the desired grouping from the collapsing scheme.

### 2023-02-20 pkg/inst/tinytest/test_accumulate.R

@@ -104,6 +104,25 @@ out <- cumulate(data=input
 
 expect_equivalent(sapply(out$model, coef), out$mean)
 
+## test connection with 'validator' 
+if ( !requireNamespace("validate", quietly=TRUE) ){ 
+  exit_file("validate not installed")
+}
+
+rules <- validate::validator(nrow(.) >= 3, sum(Y >= 2) >= 3)
+
+input <- data.frame(
+     A  = c(1,1,1,2,2,2,3,3,3)
+   , B  = c(11,11,11,12,12,13,21,22,12)
+   , B1 = c(1,1,1,1,1,1,2,2,1)
+   , Y  = 2^(0:8)
+)
+
+expect_silent(accumulate(input, collapse = A*B ~ A*B1 + B1,
+    test = from_validator(rules), fun=sum))
+
+
+
 
 
 

### 2023-02-21 paper/article.Rnw
wip

@@
 \usepackage{tikz}
+
+% shortcut for surjection arrow, and surjection version of xrightarrow
+\newcommand{\onto}{\twoheadrightarrow}
+\newcommand\xonto[2][]{%
+  \mathrel{\ooalign{$\xrightarrow[#1\mkern4mu]{#2\mkern4mu}$\cr%
+  \hidewidth$\rightarrow\mkern4mu$}}
+}
 
@@
 
-\section{Formal descriptions and algorithms}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Formal description and algorithms}
 In this section we give a formal description of the algorithm for aggregation
@@
 \subsection{Split-apply-combine}
-In order to analyse a data set group by group we need to specify a data set, a
+To analyse a data set group by group we need to specify a data set, a
 way to split it into groups, and a function that takes a subset of data and
@@
 
-Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
-$U$ and returns a value in some domain $X$. Here, $U$ represents a data set and
-$\phi$ an aggregating function. We split $U$ into groups using the following
-notation. Let $A$ be finite set that has no more elements than $U$, and let $f$
-be a function that takes an element of $U$ and returns a value in $A$. We can
-think of $A$ as a set of group labels, and $f$ as the function that assigns a
-label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
-groups. We say that $f:U\to A$ is a \emph{partition} of $U$.
+Denote with $U$ be a finite set, and and let $\phi:2^U\to X$ be a function that
+accepts a subset of $U$ and returns a value in some domain $X$.  Here, $U$
+represents a data set, $2^U$ its power set, and $\phi$ an aggregating function.
+We split $U$ into groups using the following notation. Let $A$ be finite set
+that has no more elements than $U$, and let $f:\onto A$ be a
+surjective function that takes an element of $U$ and returns a value in $A$. We
+can think of $A$ as a set of group labels, and $f$ as the function that assigns
+a label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
+subsets. We say that $f:U\onto A$ is a \emph{partition} of $U$. We
+also introduce the \emph{pullback along $f$}, $f^*:2^A\to 2^U$ defined as
+%
+\begin{equation*}
+f^*(S) = \{u\in U|f(u)\in S\},
+\end{equation*}
+where $S$ is a subset of $A$ \citep[Section 1.4]{fong2019invitation}.
 
@@
 \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
-\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
-\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs $(a,x)\in
-        A\times X$ }
+\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition
+       $f:U\onto A$.}
+\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs
+        $(a,x)\in A\times X$. }
 $R = \{\}$\;
 \For{$a\in A$}{
-  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
+  $d=f^*(\{a\})$\tcp*{get subset of $U$}
   $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
@@
 \end{algorithm}
+%
+In this algorithm the output is collected in a set $R$ containing pairs from
+$A\times X$: one pair for each element of $A$. Incidentely, the algorithm can
+be summarized even shorter in this notation as $R=\cup_{a\in A}(a,(\phi\circ
+f^*)(\{a\}))$, where $\circ$ denotes function composition.  Since $f^*$ always
+receives a singleton set, its definition as a function of the powerset of the
+group labels $A$ is too general for this algorithm.  However, this
+generalization turns out to be necessary when generalizing the algorithm to the
+case of dynamic grouping.
 
-Here, the output is collected in a set $R$ (this set is technically a function
-$A\to X$ since it pairs each $a\in A$ with one element of $X$). We also
-introduced some new notation: the notation $2^U$ is the power set of $U$ and
-$f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you give it a
-set of labels from $A$ and it returns all the elements of $U$ that have any of
-those labels. This definition is somewhat too general for the current
-algorithm, but it will be necessary when we move to collapsing groups.
-
-Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented in
-practice. Consider the signature of the R \code{aggregate()} function for data
-frames (we skip arguments that are not important for the discussion).
+It is interesting to see how the elements $U$, $f:U\onto A$, and
+$\phi$ are implemented in practice. Consider the signature of the
+\proglang{R}'s \code{aggregate()} function (we skip arguments that are not
+important for the discussion).
 \begin{verbatim}
@@
 Here, \code{x} is a data frame where each column is a variable $U$ to be
-aggregated.  The parameter \code{by} represents the record-wise label(s) that
-are used for grouping. So the function $f\to A$ is implemented by forcing the
-user to make sure that the position of each label in \code{by} corresponds to
-the row number in the data frame \code{x}. The argument \code{FUN} is the
-function $\phi$ that aggregates each subset of \code{x}.  Here is an example of how
-a user might call this function from the R prompt.
+aggregated.  The parameter \code{by} is a list of vectors of group labels,
+where each vector has a with a length that equals the number of rows in
+\code{x}. So the function $f:U\onto A$ is implemented by asking
+the user to make sure that the position of each label corresponds to the
+correct row number in the data frame \code{x}. The argument \code{FUN} is the
+function $\phi$ that aggregates each subset of \code{x}.  Here is an example of
+how a user might call this function from the R prompt.
 %
@@
 Note that the correspondence in position of the \texttt{Species} label and the
-record position is implemented by taking them from the same data frame. The output
-also reveals in the first column the set $A$: each row corresponds to a unique
-value in \texttt{Species} column.
+record position is implemented by taking them from the same data frame. The
+output also reveals in the first column the set $A$: each row corresponds to a
+unique value in \texttt{Species} column.
 
@@
 The goal of the algorithm is to compute a value for each part of a dataset,
-possibly using values from a larger subset. The input of the algorithm consists
-again of a finite set $U$ and an aggregation function $\phi$ that takes a subset
-of $U$ and returns a value in some domain $X$.  Compared to
-Algorithm~\ref{alg:sac} we need two extra inputs. First, we need a user-defined
-function that checks whether a given subset $d$ of $U$ is suitable for
+possibly using values external to the part. The input of the algorithm consists
+again of a finite set $U$ and an aggregation function $\phi$ that takes a
+subset of $U$ and returns a value in some domain $X$.  Compared to
+Algorithm~\ref{alg:sac} two different inputs are needed. First, a function must
+be defined that that checks whether a given subset $d$ of $U$ is suitable for
 computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
@@
 \begin{equation}
-U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
+U\xonto{f_0}A_0\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
 \label{eq:collapsingsequence}
@@
 
-%Collapsing schemes can be represented as tables. For example, the European NACE
-%classification is a hierarchical classification of economic activity. Each
-%group is encoded as a 4-digit code where the first digit indicates the highest
-%grouping level, and the last digit the most detailed level. This induces a
-%collapsing sequence by combining groups that have the same parent in the
-%hierarchy. Below is an example, based on a small piece of the NACE code list.
-%%
-%\begin{equation}
-%\begin{tabular}{rrr}
-%\hline
-% $A_0$& $A_1$ & $A_2$\\
-%\hline
-%0111&   011 & 01\\
-%0112&   011 & 01\\
-%0113&   011 & 01\\
-%0121&   012 & 01\\
-%0122&   012 & 01\\
-%0123&   012 & 01\\
-%0124&   012 & 01\\
-%\hline
-%\end{tabular}
-%\label{eq:tablerep}
-%\end{equation}
-%
-%Here, $A_0$ is the most detailed partition, splitting a dataset into seven
-%groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
-%The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
-%$0113$ and the second, labeled $01.2$ is constructed by combining
-%$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.
-%
-%To compute an aggregate for records labeled with $0111$, the algorithm will
-%select those records and test whether the subset is suitable. If not, it will
-%move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
-%follows.
-%\begin{displaymath}
-%f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
-%\end{displaymath}
-%It will then move on and find all records labeled with any of those three
-%group labels, and test again. If this new data set is suitable, the result is
-%computed, if not, another collapse takes place. The whole procedure is repeated 
-%for each element in $A_0$.
+Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
+returns the corresponding label in $A_k$. In other words, $F_k$ is the
+composition $f_k\circ f_{k-1}\circ\cdots \circ f_1$. Similarly we define the
+pullback along $F_k$ as $F_k^* = f_1^*\circ f_2^*\circ\cdots\circ f_k^*$.  This
+function accepts a set of labels in $A_k$ and returns all the labels in $A_0$
+that are mapped to those labels via the collapsing sequence of
+Equation~\eqref{eq:collapsingsequence}.  With this notation we can define the
+Algorithm for aggregation with dynamic grouping as follows.
 
-Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
-returns the corresponding label in $A_k$. This can formally be defined as
-consecutive application of $f_1, f_2,\cdots, f_k$ to an element of $A_0$. In
-notation
-\begin{equation*}
-F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
-\end{equation*}
-Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
-\begin{equation*}
-F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
-\end{equation*}
-This function asks for a set of labels in $A_k$ and returns all the labels in
-$A_0$ that are mapped to those labels via the collapsing sequence of
-Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally write
-down our main Algorithm~\ref{alg:saccg}.
-
-%
 \begin{algorithm}[H]
@@
 \Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
-      and a collapsing sequence $U\xrightarrow{f_0}A_0\cdots \xrightarrow{f_n}  A_n$}
+      and a collapsing sequence $U\xonto{f_0}A_0\xonto{f_1}\cdots \xonto{f_n}  A_n$}
 
@@
 \For{$a\in A_0$}{
-  $i=0$\;
-  $d = f_0^{-1}(\{a\})$\tcp*{get subset of $U$}
-  \While{$i<n \land \lnot\beta(d)$}{
+  $i=0$ \tcp*{Initiate collapse level}
+  $d = f_0^*(\{a\})$ \tcp*{Get subset of $U$}
+  \While{$i<n \land \lnot\beta(d)$}{ \label{line:while}
     $i = i+1$ \tcp*{Increase collapse level} 
-    $d = (f_0^{-1}\circ F_i^{-1}\circ F_i)(a)$ \tcp*{Collapse and get subset}
-  }
+    $d = (f_0^*\circ F_i^*\circ F_i)(a)$ \tcp*{Collapse and get subset}
+  }\label{line:endwhile}
   \If{$i<n \lor \beta(d)$}{ \label{line:cond}
@@
 %
-Since the used level of collapsing is determined dynamically by data
-circumstances in $U$, the algorithm also reports the collapsing level used to
-compute a value foe each member of $A_0$. For each member of $A_0$ there is a
-triple $(a,i,\phi(d))$, where $a\in A_0$ is the label to which the value
-$\phi(d)$ pertains, and $i$ is the number of collapses applied to reach a
-suitable dataset. The condition in Line~\ref{line:cond} ensures that if no
-suitable dataset is found after the whole collapsing sequence has been
-executed, then no answer is returned.
+
+In this algorithm the collapsing level $i$ is increased until the test is
+passed or the maximum collapsing level $n$ is reached
+(Lines~\ref{line:while}-\ref{line:endwhile}) %Since the eventual level of
+collapsing is determined dynamically by data %circumstances in $U$, the
+algorithm also reports the collapsing level $i$ used.  The condition in
+Line~\ref{line:cond} ensures that if no suitable dataset is found after the
+whole collapsing sequence has been executed, then no answer is returned. This
+means that in contrast with Algorithm~\ref{alg:sac} there is no guarantee that
+a value for each member of $A_0$ will be found. For each member of $A_0$ where
+an aggregate is computed, there is a triple $(a,i,\phi(d))$, where $a\in A_0$
+is the label to which the value $\phi(d)$ pertains, and $i$ is the number of
+collapses applied to reach a suitable dataset.
 

### 2023-02-21 paper/article.Rnw
wip

@@
 $A\times X$: one pair for each element of $A$. Incidentely, the algorithm can
-be summarized even shorter in this notation as $R=\cup_{a\in A}(a,(\phi\circ
-f^*)(\{a\}))$, where $\circ$ denotes function composition.  Since $f^*$ always
-receives a singleton set, its definition as a function of the powerset of the
-group labels $A$ is too general for this algorithm.  However, this
+be summarized even shorter in this notation as $R=\cup_{a\in A}\{(a,(\phi\circ
+f^*)(\{a\}))\}$, where $\circ$ denotes function composition.  Since $f^*$
+always receives a singleton set, its definition as a function of the powerset
+of the group labels $A$ is too general for this algorithm.  However, this
 generalization turns out to be necessary when generalizing the algorithm to the
@@
 \begin{equation}
-U\xonto{f_0}A_0\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
+U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
 \label{eq:collapsingsequence}
 \end{equation}
-A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
-partitions $A_{i-1}$ into $|A_i|$ groups.
+A collapsing scheme is a sequence of partitions where each $f_i$ partitions its
+domain in $|A_i|$ groups while $f$ partitions $U$ in $|A|$ groups.
 
-Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
+Denote with $F_k:A\to A_k$, the function that  accepts a label in $A$ and
 returns the corresponding label in $A_k$. In other words, $F_k$ is the
@@
 pullback along $F_k$ as $F_k^* = f_1^*\circ f_2^*\circ\cdots\circ f_k^*$.  This
-function accepts a set of labels in $A_k$ and returns all the labels in $A_0$
+function accepts a set of labels in $A_k$ and returns all the labels in $A$
 that are mapped to those labels via the collapsing sequence of
@@
 \Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
-      and a collapsing sequence $U\xonto{f_0}A_0\xonto{f_1}\cdots \xonto{f_n}  A_n$}
+      and a collapsing sequence $U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots \xonto{f_n}  A_n$.}
 
 \Output{$R$: the value of $\phi$ for every part of $U$, for which a suitable
-collapsing group can be found, as a set of triples $(a,k,x)\in A_0\times
+collapsing group can be found, as a set of triples $(a,k,x)\in A\times
 \underline{n}\times X $ where $\underline{n}=\{0,1,\ldots,n\}$.}
@@
 $R = \{\}$\;
-\For{$a\in A_0$}{
+\For{$a\in A$}{
   $i=0$ \tcp*{Initiate collapse level}
-  $d = f_0^*(\{a\})$ \tcp*{Get subset of $U$}
+  $d = f^*(\{a\})$ \tcp*{Get subset of $U$}
   \While{$i<n \land \lnot\beta(d)$}{ \label{line:while}
     $i = i+1$ \tcp*{Increase collapse level} 
-    $d = (f_0^*\circ F_i^*\circ F_i)(a)$ \tcp*{Collapse and get subset}
+    $d = (f^*\circ F_i^*\circ F_i)(a)$ \tcp*{Collapse and get subset}
   }\label{line:endwhile}
@@
 means that in contrast with Algorithm~\ref{alg:sac} there is no guarantee that
-a value for each member of $A_0$ will be found. For each member of $A_0$ where
-an aggregate is computed, there is a triple $(a,i,\phi(d))$, where $a\in A_0$
+a value for each member of $A$ will be found. For each member of $A$ where
+an aggregate is computed, there is a triple $(a,i,\phi(d))$, where $a\in A$
 is the label to which the value $\phi(d)$ pertains, and $i$ is the number of
@@
 
-\subsection{Performance, memory, parallelization}
-Both Algorithms~\ref{alg:sac} and \ref{alg:saccg} need only $\mathcal{O}(|U|)$
-memory. The time complexity of Algorithm~\ref{alg:sac} is $\mathcal{O}(|A||U|)$
-($|A|$ the number of groups occurring in the data, and $|U|$ the number of
-records) but the actual running time is also determined by the form of the
-aggregating function $\phi$.  For simple aggregating functions such as
-summation or counting, the output can be computed with a single pass through
-$|U|$. \proglang{R} packages such as \pkg{collapse} and \pkg{data.table} have
-implemented specialized low level (\proglang{C}/\proglang{C++}) code for fast
-groupwise aggregation for some commonly occurring aggregating functions.
-Generic parallelization of Algorithm~\ref{alg:sac} can be done by parallelizing
-over the grouping values $a\in A$: each worker node gets access to a set of
-records corresponding to a single group and computes the local aggregate.
-
-Algorithm~\ref{alg:saccg} has time complexity of $\mathcal{O}((n+1)|A_0||U|)$,
-Namely, in the worst case, for each $a\in A_0$ we need to collapse to the
-maximum level $A_n$. This means that for each $a\in A_0$ we need to collapse
-$n$ times, plus the initial attempt of computing aggregate without collapsing.
-The fact that we cannot generically exclude records from a calculation means
-that a generic parallelization scheme would need to ensure that each worker
-node has access to the full dataset $|U|$. There are two types of options to do
-this: either multliple workers gain read-only access to $U$ or each worker gets
-a copy of $U$. In the first case the workers may suffer from colliding
-read-operations which hampers linear performance improvement as a function of
-the number of workers. In the second case, one needs as many copies of the data
-as there are threads. It is possible to compute beforehand whether the dataset
-can be split into subsets that can be treated independently by determining all
-the collapsed subsets. However, this is almost as expensive as running the
-algorithm completely since it only skips the calculation of aggregates after
-collapsing.
-
-Summarizing, we see that split-apply-combine with collapsing groups has a
-higher worst-case complexity than regular split-apply-combine. Moreover,
-generic parallelization needs to either consider a specific use case (e.g.
-large or small datasets) or require user input that guides the choice of
-parallelization scheme.
+Comparing Algorithms~\ref{alg:sac} and~\ref{alg:saccg}, we see that the
+standard split-apply-combine algorithm has worst-case runtime complexity
+$O(|A|)$, which is equal to the best case $\Omega(|A|)$. The
+split-apply-combine with collapsing groups algorithm has best case
+$\Omega(|A|)$ equal to standard split-apply-combine but worst case $O(n|A|)$.
+In fact, the best case for Algorithm~\ref{alg:saccg} is achieved by setting
+$\beta:d\mapsto\texttt{True}$. It is not hard to see that in this case
+Algorithm~\ref{alg:saccg} reduces to Algorithm~\ref{alg:sac}. The worst case is
+achieved by setting $\beta: d\mapsto \texttt{False}$.
 

### 2023-02-28 paper/notes/notes.tex
notes

\documentclass[fleqn,11pt]{article}
\usepackage{amsmath, amssymb}
\title{notes}

\begin{document}
\subsection*{Notation}
Given a finite population $U=\{1,2,\ldots, N\}$ partitioned in domains
$\{D_i\}_{i=1}^M$, of sizes $N_i$.  Draw a sample $S\subseteq U$ of size
$n$ uniformly without replacement from $U$. We have

$$
\mathbb{P}(j\in S) = \frac{n}{N}\textrm{ for all } j\in U.
$$

Assume we measure a variable $Y$ with realisations $\{y_j|j\in S\}$. We shall
be interested in estimating the population total $t_Y$, the population mean
$\mu_Y$ and the totals and means for each domain, denoted $t_{Y|i}$ and $\mu_{Y|i}$
respectively.

\subsection*{Horwitz-Thomson estimation}
The Horwitz-Thomson estimator for the population total is defined as
$$
\hat{t}^{\textrm{HT}}_Y = \sum_{j\in S} \frac{1}{\mathbb{P}(j\in S)}y_j
=\sum_{j\in S}\frac{N}{n}y_j = N\bar{y},
$$
where $\bar{y}$ is the sample mean of $Y$.  The HT estimator for a domain $D_i$
is given by
$$
\hat{t}_{Y|i}^{\textrm{HT}} = \sum_{j\in S_i}\frac{N}{n}y_j,\textrm{ with } S_i=D_i\cap S.
$$
This estimator has the property that
$\sum_{i=1}^M\hat{t}_{Y|i}^{\textrm{HT}}=\hat{t}_{Y}^{\textrm{HT}}$, or more
generally, for any $L\subseteq \{1,2,\ldots,M\}$ we have the additivity
property $\hat{t}^{\textrm{HT}}_{Y|L} = \sum_{l\in
L}\hat{t}^{\textrm{HT}}_{Y|l}$.

For the corresponding estimator of the population mean we get the following.
$$
\hat{\mu}_{Y}^{\textrm{HT}} = \frac{1}{N}\hat{t}^{\textrm{HT}}_Y = 
\frac{1}{n}\sum_{j\in S}y_j = \bar{y}.
$$
For the domain means we define
$$
\hat{\mu}_{Y|i}^{\textrm{HT}} = \frac{1}{N_i}\hat{t}^{\textrm{HT}}_{Y|i}
= \frac{N}{N_i n}\sum_{j\in S_i}y_j.
$$
We can rewrite this by observing that $\mathbb{E}(n_i)=\frac{N_in}{N}$ to get
$$
\hat{\mu}_{Y|i}^{\textrm{HT}} = \frac{1}{\mathbb{E}(n_i)}\sum_{j\in S_i}y_j
=\frac{n_i}{\mathbb{E}(n_i)}\bar{y}_i,
$$
where $\bar{y}_i$ is the sample mean of $Y$ observed in $S_i$.

\subsection*{The BARE estimator for undersampled domains}
In cases where $n_i$ is small, the estimation variance may be unacceptably
large.  A variance-bias tradeoff can be made by including units outside $S_i$
in the estimates for $S_i$. This is called \emph{Small Area Estimation} (SAE).
In SAE jargon, this is called `borrowing statistical strength' from other
domains.

One of the simplest estimators is the Broader Area Ratio Estimator (BARE). The
idea is to estimate the total for a broader domain that includes $D_i$ and then
disaggregate the total proportional to the domain size $N_i$. Choosing the
whole population $U$ for the broader we get the following.
$$
\hat{t}^{\textrm{BARE}}_Y = \frac{N_i}{N}\hat{t}^{\textrm{HT}}_Y = N_i\bar{y}.
$$
(Note that $\hat{y}$ is the mean, taken over the whole sample $S$). And for the
mean:
$$
\hat{\mu}^{\textrm{BARE}}_Y  = \frac{1}{N_i}\hat{t}^{\textrm{BARE}}_Y  = \bar{y}.
$$
Thus, the BARE estimator for $\mu_{Y|i}$ for a domain $i$ is given by the
estimated mean of the population.

More generally, suppose that $L\subseteq \{1,2,\ldots,M\}$; choose a $D_i$
with $i\in L$, and write $S_L = \cup_{l\in L}{S_l}$, then
%
$$
\hat{t}^{\textrm{BARE}}_{Y|i}(L) 
= \frac{N_i}{N_L}\hat{t}^{\textrm{HT}}_{Y|L}
=\frac{N_i}{N_L}\sum_{j\in S_L}\frac{N}{n}y_l 
= \frac{N}{N_L n}N_i n_L\bar{y}_L.
$$ 
Noting furthermore that $\mathbb{E}(n_L)=N_Ln/N$ we get
$$
\hat{t}^{\textrm{BARE}}_{Y|i}(L)= N_i\frac{n_L}{\mathbb{E}(n_L)}\bar{y}_L,
= N_i \hat{\mu}^{\textrm{HT}}_{Y|L}.
$$
Finally, we get for the mean
$$
\hat{\mu}^{\textrm{BARE}}_{Y|i}(L) 
= \frac{1}{N_i}\hat{t}^{\textrm{BARE}}_{Y|i}(L)
= \hat{\mu}^{\textrm{HT}}_{Y|L}
$$

If we assume the approximation $n_L/\frac{\mathbb{E}(n_L}\approx 1$ we 
get
$$
\hat{\mu}^{\textrm{BARE}}_{Y|i}(L)\approx \bar{y}_L.
$$

\end{document}

### 2023-02-28 paper/article.Rnw
witp
notes

@@
 unacceptable estimation variance. Instead one resorts to indirect estimation,
-meaning that one estimates the parameter for a larger subpopulation that
-includes the target subpopulation. This estimate is then interpreted as an
-estimate for the target subpopulation. Here, we shall be interested in
-estimating the average turnover from industrial activities (\code{industrial})
-by SBI and size class.
+meaning that one estimates parameters for a larger subpopulation which are then
+used in the estimate for the target subpopulation.  Here, we shall be
+interested in estimating the average turnover from industrial activities
+(\code{industrial}) by SBI and size class.
 
@@
   \If{$i<n \lor \beta(d)$}{ \label{line:cond}
-    $R = R\cup \{(a, i,\phi(d))\}$\;
+    $R = R\cup \{(a, i,\phi(d))\}$\; \label{line:R}
   } 
@@
 Algorithm~\ref{alg:saccg} reduces to Algorithm~\ref{alg:sac}. The worst case is
-achieved by setting $\beta: d\mapsto \texttt{False}$.
+achieved by setting $\beta: d\mapsto \texttt{False}$. In that case the while
+loop in Line~\ref{line:while} is iterated $n-1$ times (yielding total $n$
+executions of $f^*$) while Line~\ref{line:R} is never executed.
 

### 2023-02-28 paper/notes/notes.tex
notes

@@
 
-If we assume the approximation $n_L/\frac{\mathbb{E}(n_L}\approx 1$ we 
-get
+If we assume the approximation $n_L/\mathbb{E}(n_L)\approx 1$ we get
 $$
@@
 $$
+Note that in the limit where $L=\{1,2,\ldots M\}$, i.e. the case where $S_L$
+equals the full sample $S$, the approximation becomes an equality. It is not in
+true in general that the approximation improves when $L$ is increased.
 

### 2023-03-03 paper/article.Rnw
fix edge case with just one object_list output

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Specifying tests}
+The \code{test} parameter of \code{cumulate()} and \code{accumulate} accepts a
+function that takes a subset of the data and must returns a boolean. For common
+test conditions, including requiring a minimal number of records, or a minimal
+number or fraction of complete records there are helper functions available.
+%
+\begin{center}
+\begin{tabular}{lp{9cm}}
+\code{min_records(n)}        & At least $n$ records.\\ 
+\code{min_complete(n,vars)}  & At least $n$ records complete for variables \code{vars}.\\
+\code{frac_complete(r,vars)} & At least $100r$\% complete records for variables \code{vars}.\\
+\code{from_validator(v,...)} & Construct a testing function from a \code{validator} object
+                               of \proglang{R} package \pkg{validate}.\\
+\end{tabular}
+\end{center}
+%
+Second, for multiple, possibly complex requirements on variables users can
+express conditions with the \pkg{validate} package \cite{loo2021data}.  The
+\pkg{validate} packages offers a domain-specific language for expressing,
+manipulating, and investigating conditions on datasets. It's core concept is a
+list of `data validation rules' stored as a \class{validator} object. A
+\class{validator} object is constructed with the eponymous function
+\code{validator()}.  For example, to demand that there are at least 3 rows in a
+group, and that there are at least three records where $Y\geq 2$ we create the
+following ruleset.
+<<>>=
+library(validate)
+rules <- validator(nrow(.) >= 3, sum(Y >= 2) >= 3)
+rules
+@
+Here the \code{.} refers to the dataset as a whole, while rules that can be
+evaluated within the dataset can be written as boolean \proglang{R}
+expressions.
+
+We will apply these conditions to the \code{input} dataset that was 
+constructed in Section~\ref{sect:formula}. As a reminder we print
+the first few records.
+<<>>=
+head(input, 4)
+@
+We use \code{A*B ~ A*B1 + B1} as collapsing scheme. The function
+\code{from_validator} passes the requirements as a test function to
+\code{accumulate()} (or \code{cumulate()}).
+<<>>=
+accumulate(input, collapse = A*B ~ A*B1 + B1,
+    test = from_validator(rules), fun=sum)
+@
+Note that for target groups $(A=2,B=21)$ and $(A=2,B=22)$ none of the available
+collapsing levels lead to a group that satisfied all conditions. Therefore the
+collapsing level and output variables are all missing (\code{NA}).
+
+The third and most flexible way for users to express tests is to write a custom
+testing function. The requirements are that it must work on any subset of a
+data frame, including a dataset with zero rows. The previous example can
+thus also be expressed as follows.
+<<>>=
+my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >=3
+accumulate(input, collapse = A*B ~ A*B1 + B1, test=my_test, fun=sum)
+@
+
+It is easy to overlook some edge cases when specifying test functions.
+Recall that a test function is required to return \code{TRUE} or \code{FALSE},
+regardless of the data circumstances. The only thing that a test function
+can assume is that the received data set is a subset of records from the
+dataset to be aggregated. As a service to the user, \pkg{accumulate} exports
+a function that checks a test function against common edge cases, including
+the occurrence of missing values, a dataset with zero rows, and the full
+dataset. The smoke test checks whether the output is \code{TRUE} or \code{FALSE}
+under all circumstances and also reports errors, warnings, and messages.
+It accepts a (realistic) dataset and a testing function and prints
+test results to the console.
+<<>>=
+smoke_test(input, my_test)
+@
+By default only failing tests are printed. In this case our test function
+is not robust against missing values for $Y$. This can be remedied
+by adding \code{na.rm=TRUE} to the \code{sum()} in the test function.
+<<>>=
+my_test1 <- function(d) nrow(d) >= 3 && sum(d$Y>=2, na.rm=TRUE) >=3 
+smoke_test(input, my_test1)
+@
+The smoke test is aimed at preventing complicated stack traces when errors occur
+in a call to \code{accumulate()} or \code{cumulate()}.  Users should be aware
+that it does not guarantee correctness of the results, only robustness against
+certain edge cases.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Balanced and unbalanced hierarchical classifications}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Specifying tests}
-The \code{test} parameter of \code{cumulate()} and \code{accumulate} accepts a
-function that takes a subset of the data and must returns a boolean. For common
-test conditions, including requiring a minimal number of records, or a minimal
-number or fraction of complete records there are helper functions available.
-%
-\begin{center}
-\begin{tabular}{ll}
-\code{min_records(n)}        & At least $n$ records.\\ 
-\code{min_complete(n,vars)}  & At least $n$ records complete records for variables \code{vars}.\\
-\code{frac_complete(r,vars)} & At least $100r$\% complete records for variables \code{vars}.\\
-\end{tabular}
-\end{center}
-
-Second, for multiple, possibly complex requirements on variables users can
-express conditions with the \pkg{validate} package \cite{loo2021data}.  The
-\pkg{validate} packages offers a domain-specific language for expressing,
-manipulating, and investigating conditions on datasets. It's core concept is a
-list of `data validation rules' stored as a \class{validator} object. A
-\class{validator} object is constructed with the eponymous function
-\code{validator()}.  For example, to demand that there are at least 3 rows in a
-group, and that there are at least three records where $Y\geq 2$ we create the
-following ruleset.
-<<>>=
-library(validate)
-rules <- validator(nrow(.) >= 3, sum(Y >= 2) >= 3)
-rules
-@
-Here the \code{.} refers to the dataset as a whole, while rules that can be
-evaluated within the dataset can be written as boolean \proglang{R}
-expressions.
-
-We will apply these conditions to the \code{input} dataset that was 
-constructed in Section~\ref{sect:formula}. As a reminder we print
-the first few records.
-<<>>=
-head(input, 4)
-@
-We use \code{A*B ~ A*B1 + B1} as collapsing scheme. The function
-\code{from_validator} passes the requirements as a test function to
-\code{accumulate()} (or \code{cumulate()}).
-<<>>=
-accumulate(input, collapse = A*B ~ A*B1 + B1,
-    test = from_validator(rules), fun=sum)
-@
-Note that for target groups $(A=2,B=21)$ and $(A=2,B=22)$ none of the available
-collapsing levels lead to a group that satisfied all conditions. Therefore the
-collapsing level and output variables are all missing (\code{NA}).
-
-The third and most flexible way for users to express tests is to write a custom
-testing function. The requirements are that it must work on any subset of a
-data frame, including a dataset with zero rows. The previous example can
-thus also be expressed as follows.
-<<>>=
-my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >=3
-accumulate(input, collapse = A*B ~ A*B1 + B1, test=my_test, fun=sum)
-@
-
-It is easy to overlook some edge cases when specifying test functions.
-Recall that a test function is required to return \code{TRUE} or \code{FALSE},
-regardless of the data circumstances. The only thing that a test function
-can assume is that the received data set is a subset of records from the
-dataset to be aggregated. As a service to the user, \pkg{accumulate} exports
-a function that checks a test function against common edge cases, including
-the occurrence of missing values, a dataset with zero rows, and the full
-dataset. The smoke test checks whether the output is \code{TRUE} or \code{FALSE}
-under all circumstances and also reports errors, warnings, and messages.
-It accepts a (realistic) dataset and a testing function and prints
-test results to the console.
-<<>>=
-smoke_test(input, my_test)
-@
-By default only failing tests are printed. In this case our test function
-is not robust against missing values for $Y$. This can be remedied
-by adding \code{na.rm=TRUE} to the \code{sum()} in the test function.
-<<>>=
-my_test1 <- function(d) nrow(d) >= 3 && sum(d$Y>=2, na.rm=TRUE) >=3 
-smoke_test(input, my_test1)
-@
-The smoke test is aimed at preventing complicated stack traces when errors occur
-in a call to \code{accumulate()} or \code{cumulate()}.  Users should be aware
-that it does not guarantee correctness of the results, only robustness against
-certain edge cases.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Extensive example: economic data} \label{sect:example}
@@
 observations in a subpopulation is so mall that direct estimation leads to
-unacceptable estimation variance. Instead one resorts to indirect estimation,
-meaning that one estimates parameters for a larger subpopulation which are then
-used in the estimate for the target subpopulation.  Here, we shall be
-interested in estimating the average turnover from industrial activities
-(\code{industrial}) by SBI and size class.
+unacceptable estimation variance. Instead one may resort for example to
+indirect estimation, meaning that one estimates parameters for a larger
+subpopulation which are then used in the estimate for the target subpopulation.
+Here, we shall be interested in estimating the average turnover from industrial
+activities (\code{industrial}) by SBI and size class.
 
@@
 replaces the estimator of the mean over a subpopulation with the estimator of
-the mean over a larger subpopulation. If we assume that all units in the
-dataset were sampled with equal probability based on a simple random sampling
-design, the mean is estimated with the sample mean. We will demand that there
-are at least ten records for which turnover has been measured. The collapsing
-scheme is given by \code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and
-\code{sbi1} are classification by respectively the first two SBI digits and the
-first SBI digit. We first add those variables to the dataset.
+the mean over a larger subpopulation that includes the target subpopulation
+\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
+by simple random sampling from the population, the mean can be estimated with
+the sample mean. In this example we will demand that there are at least ten
+records for which turnover has been measured. The collapsing scheme is given by
+\code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
+classification by respectively the first two SBI digits and the first SBI
+digit. We first add those variables to the dataset.
 <<>>=
@@
 @
-Our test function looks as follows.
-<<>>=
-testfun <- function(d) sum(!is.na(d$industrial)) >= 10
-@
 Using \code{accumulate()} we obtain the means. 
@@
              , collapse = sbi*size ~ sbi + sbi2 + sbi1
-             , test = testfun
+             , test = min_complete(n=10, vars="industrial")
              , mean_industrial = mean(industrial, na.rm=TRUE))
@@
 @
+
+In terms of SAE, the collapsing scheme expresses the assumption that estimates
+on the level of respectively \code{sbi}, \code{sbi2} and \code{sbi1} introduce
+an acceptable bias.
 
@@
          , collapse = sbi*size ~ sbi + sbi2 + sbi1
-         , test = testfun
+         , test = min_complete(n=10, vars="industrial")
          , R = mean(industrial, na.rm=TRUE)/mean(total, na.rm=TRUE))
@@
 the collapse \code{level} equals 2.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Computing linear models}
+Generalizing further, one may be interested in estimating (statistical) models.
+With the \code{cumulate} function it is possible to specify complex aggregates
+that go beyond a simple number. Below, we estimate the following linear model
+\begin{displaymath}
+\code{total} = \beta_0 + \beta_i\code{industrial} + \varepsilon,
+\end{displaymath}
+demanding that there are at least 10 records where both predictor and predicted
+variable are available.
+<<>>=
+r <- cumulate(producers
+         , collapse = sbi*size ~ sbi + sbi2 + sbi1
+         , test = min_complete(n=5, vars=c("total","industrial"))
+         , model = lm(total ~ industrial)
+         , mean  = mean(total,na.rm=TRUE))
+head(r,3)
+@
 

### 2023-03-03 pkg/R/accumulate.R

@@ -222,7 +222,7 @@ is_scalar <- function(x){
 # Output:
 # A data frame, of the form [backbone, results].
 #
-combine <- function(backbone,results, ag){
+combine <- function(backbone, results, ag){
   # The simple case: aggregates are atomic; this probably covers
   # most cases.
   if (!any(sapply(results, is.list))){
@@ -239,7 +239,6 @@ combine <- function(backbone,results, ag){
       lapply(results, `[[`,var)
     }
   })
-
   names(L) <- names(results[[1]])
   # make listcol objects of columns that are lists, so they are 
   # printed better.
@@ -349,7 +348,7 @@ get_ag <- function(cps, x, dnames, ...){
     gv <- if(inherits(cps,"formula")) all.vars(cps) else colnames(cps)
     function(dat) sapply(dat[ ,!colnames(dat) %in% gv, drop=FALSE], f)
   } else {
-    function(dat) sapply(x, function(e) with(dat,eval(e)))
+    function(dat) lapply(x, function(e) with(dat,eval(e)))
   }
   outnames <- if (!is.function(x)) names(x)
               else if (inherits(cps,"formula")) dnames[!dnames %in% all.vars(cps)]

### 2023-03-03 pkg/inst/tinytest/test_accumulate.R

@@ -94,7 +94,15 @@ out <- cumulate(data=input
 expect_equivalent(out, output)
 
 # Case where the aggregate is an object (not a scalar)
+out <- cumulate(data=input
+        , collapse = A*B ~ A*B1 + A
+        , test = function(d) nrow(d) >= 3
+        , model = lm(Y ~ 1)
+       )
 
+expect_inherits(out$model, "object_list")
+
+# with extra columns
 out <- cumulate(data=input
         , collapse = A*B ~ A*B1 + A
         , test = function(d) nrow(d) >= 3
@@ -103,6 +111,9 @@ out <- cumulate(data=input
        )
 
 expect_equivalent(sapply(out$model, coef), out$mean)
+
+
+
 
 ## test connection with 'validator' 
 if ( !requireNamespace("validate", quietly=TRUE) ){ 

### 2023-03-03 paper/article.Rnw
wip

@@
 \Abstract{
-foo
+Partitioning a data set by one or more of its attributes and computing an
+aggregate for each part is one of the most common operations in data analyses.
+There are use cases where the partitioning is determined dynamically by
+collapsing smaller subsets into larger ones, to ensure sufficient support for
+the computed aggregate. These use cases are not supported by software
+implementing split-apply-combine types of operations. This paper presents the
+\proglang{R} package \code{accumulate} that offers convenient interfaces for
+defining grouped aggregation where the grouping itself is dynamically
+determined, based on user-defined conditions on subsets, and a user-defined
+subset collapsing scheme. Finally, we demonstrate and analyze the pseudocode
+underlying this method.
 }
@@
 it's drop-in replacement \pkg{poorman} \citep{eastwood2022poorman}, and
-\proglang{R} packages \pkg{collapse} \citep{krantz2022collapse} and
-\pkg{data.table} \citep{dowle2022datatable}. In \proglang{Python} the
-\code{pandas} package implements several methods for grouping records and
-aggregating over one or more columns in data frame objects. Similarly, the
-\pkg{DataFrames} package for \proglang{julia} implements such functionality
-\citep{kaminski2022dataframes}.
+performance-focused \proglang{R} packages \pkg{collapse}
+\citep{krantz2022collapse} and \pkg{data.table} \citep{dowle2022datatable}. In
+\proglang{Python} the \code{pandas} package implements several methods for
+grouping records and aggregating over one or more columns in data frame
+objects. Similarly, the \pkg{DataFrames} package for \proglang{julia}
+implements such functionality \citep{kaminski2022dataframes}.
 
@@
 determined using attributes from out-of-group entities. One example where this
-occurs is in the area of small area estimation (see e.g.
-\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
-for example a geographical region, or a detailed population subset, where the
-number of (sampled) observations so small that the variance of the estimate
-would be unacceptably large. In small area estimation (SAE) one trades bias for
-variance by `borrowing statistical strength' from out-of-group records. The
-out-of-group records can be obtained, for example by combining the original
-small group with a group of records that are deemed similar in certain
+occurs is in the area of Small Area Estimation (SAE, see \emph{e.g.}
+\citet{rao2015small, molina2015sae}). Here, one wishes to estimate an aggregate
+for a group, for example a geographical region, or a detailed population
+subset, where the number of (sampled) observations so small that the variance
+of the estimate would be unacceptably large. In small area estimation (SAE) one
+trades bias for variance by `borrowing statistical strength' from out-of-group
+records. The out-of-group records can be obtained, for example by combining the
+original small group with a group of records that are deemed similar in certain
 respects. A second area where out-of-group records play a role is in certain
@@
 too small for a reliable estimate of model parameters, the group is enlarged by
-combining similar groups, similar to the small-area estimation scenario. A
-third example comes from the field of statistical disclosure control (SDC)
-\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
-states that the value in a grouped aggregation may not be determined for more
-than $p$ percent by a single entity. This occurs for example, when a certain
-economic activity in a geographic region is dominated by a single company. In
-such cases the value can either not be published (suppression), or one
-publishes an aggregate over larger groups. Similarly for microdata there is the
-concept of $k$-anonymity, which states that each combination of properties must
-occur at least $k$ times in a data set so as to prevent reidentification or
-disclosure of individual properties. One way to achieve this is to combine
-detailed groupings based on categorical variables into larger groups. For
-example by combining groups that in a hierarchical classification have the same
-parent.
+combining similar groups, similar to the small-area estimation scenario.
 
-Both SAE and SDC are supported by \proglang{R} and other free software. Small
-area estimation methodology has for example been implemented in the \pkg{sae}
-package of \cite{molina2015sae} and the \pkg{hbsae} package of
-\cite{boonstra2022hbsae}.  Disclosure control methodology is implemented in the
-free and open source standalone software $\tau-$Argus for tabular data and
-$\mu$-Argus for microdata \citep{argus2022}, and is also supported by
-\proglang{R} packages \pkg{sdcMicro} and \pkg{sdcTable}
-\citep{templ2015statistical, meindl2022sdctable}. Regarding imputation
-methodology, the CRAN task view on missing
-data\footnote{https://cran.r-project.org/web/views/MissingData.html} currently
-lists 203 \pkg{R} packages that support some form of estimating missing data.
-The \pkg{simputation} package \citep{loo2022simputation} seems to be the only
-one that allows for some kind of fall-through scenario for selecting methods,
-but it does not allow for dynamic grouping.
+SAE as a special case is well supported by \proglang{R} and other free
+software. Methodology has for example been implemented in the \pkg{sae} package
+of \cite{molina2015sae} and the \pkg{hbsae} package of
+\cite{boonstra2022hbsae}.  Regarding imputation methodology, the CRAN task view
+on missing data\footnote{https://cran.r-project.org/web/views/MissingData.html}
+currently lists 203 \pkg{R} packages that support some form of estimating
+missing data.  The \pkg{simputation} package \citep{loo2022simputation} seems
+to be the only one that allows for some kind of fall-through scenario for
+selecting methods, but it does not allow for dynamic grouping.
 
@@
 compute aggregates for a certain grouping of records. However, if a certain
-instance of a group does not meet user-defined quality requirements, the set of
+instance of a group does not meet user-defined quality requirements the set of
 records is expanded by (recursively) collapsing the grouping according to a
@@
 to define any quality requirement on the grouping while common quality
-requirements are supported out of the box. Common quality requirements include a
-minimum number of records, or a minimum fraction of records without missing
-values, or a minimum number of records with non-zero values.
+requirements are supported out of the box. Common quality requirements include
+a minimum number of records, or a minimum fraction of records without missing
+values, or a minimum number of records with non-zero values. Finally, the
+package should support simple outputs such as counts or averages, but also
+compound objects such as the output of model estimates.
 
-The rest of this paper is organized as follows. In the next section we formally
-define the concepts of the quality requirements and a collapsing scheme. We
-describe the fundamental algorithm in pseudocode and compare it with the
+The rest of this paper is organized as follows.  In the next section we
+visually explain dynamic grouping via a collapsing scheme and introduce the
+running example that will be used in Section~\ref{sect:accumulate} which
+introduces the \pkg{accumulate} package, the main interfaces and helper
+functions. In Section~\ref{sect:accumulate} we also discuss the common case of
+collapsing via a predifined hiearchical classification scheme.
+Section~\ref{sect:example} demonstrates the package on a realistic synthetic
+dataset. The case of complex outputs is also demonstrated in this Section.  In
+Section~\ref{sect:algorithm} we derive the pseudocode that solves the problem
+of aggregating with dynamically collapsing groups. We show that it is a precise
+and straightforward generalization of the standard split-apply-combine problem
+and analyze its time complexity.
+
+We describe the fundamental algorithm in pseudocode and compare it with the
 standard grouping scenario. We also discuss the time and space complexity and
-some caveats with implementation and performance. This section is mainly of
-interest for developers who wish to include such functionality in their own
-package. In Section~\ref{sect:accumulate} we describe \pkg{accumulate} package
-and demonstrate how the design choices have determined its implementation.  We
-give some toy examples to extensively discuss the input and output of the main
-functions, discuss how to define tests in some detail, and also explain how one
-can program over the functions in the package.  Section~\ref{sect:example}
-works through a realistic example based on a synthetic dataset of economic
-microdata. We conclude with a summary and outlook in
-Section~\ref{sect:conclusion}.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Dynamic grouping} \label{sect:dynamicgrouping}
@@
 compute the aggregate (0 indicating no collapse), and the last set of columns
-store the aggregates.
+store the aggregates. Output aggregates may be of a simple data type
+(\code{numeric}, \code{character}, \code{logical},$\ldots$) or of a composed
+type such as the output of a linear model. The latter case is demonstrated in
+Section~\ref{sect:complex}.
 
@@
 the formula interface, which requires that the collapsing sequence is
-represented as variables in the data set to be aggregated. The second is the
-tabular interface, where a table similar to~\eqref{eq:tablerep} is expected.
+represented as variables in the data set to be aggregated. The second is a
+tabular interface where each row starts with the one of the lowest-level group
+labels, and subsequent columns contain labels of courser groups.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computing linear models}
+\subsection{Computing complex aggregates} \label{sect:complex}
 Generalizing further, one may be interested in estimating (statistical) models.
@@
 \begin{displaymath}
-\code{total} = \beta_0 + \beta_i\code{industrial} + \varepsilon,
+\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
 \end{displaymath}
@@
          , collapse = sbi*size ~ sbi + sbi2 + sbi1
-         , test = min_complete(n=5, vars=c("total","industrial"))
-         , model = lm(total ~ industrial)
-         , mean  = mean(total,na.rm=TRUE))
+         , test = min_complete(n=10, vars=c("total","industrial"))
+         , model = lm(total ~ industrial))
 head(r,3)
 @
+Here, the last column is a list of class \code{object_list}, where each
+element is either an object of class \code{lm} or \code{NA}. Variables of class
+\code{object_list} only differ from from a standard \proglang{R} list by their
+print method.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Formal description and algorithms}
+\section{Formal description and algorithms} \label{sect:algorithm}
 In this section we give a formal description of the algorithm for aggregation
@@
 standard split-apply-combine algorithm has worst-case runtime complexity
-$O(|A|)$, which is equal to the best case $\Omega(|A|)$. The
-split-apply-combine with collapsing groups algorithm has best case
-$\Omega(|A|)$ equal to standard split-apply-combine but worst case $O(n|A|)$.
-In fact, the best case for Algorithm~\ref{alg:saccg} is achieved by setting
-$\beta:d\mapsto\texttt{True}$. It is not hard to see that in this case
-Algorithm~\ref{alg:saccg} reduces to Algorithm~\ref{alg:sac}. The worst case is
-achieved by setting $\beta: d\mapsto \texttt{False}$. In that case the while
-loop in Line~\ref{line:while} is iterated $n-1$ times (yielding total $n$
-executions of $f^*$) while Line~\ref{line:R} is never executed.
+$O(|A|)$, as determined by counting applications of $f^*$. This is equal to the
+algorithm's best case $\Omega(|A|)$. The split-apply-combine with collapsing
+groups algorithm also has best case $\Omega(|A|)$ but has worst case $O(n|A|)$
+(with $n$ the number of collapsing steps). In fact, the best case for
+Algorithm~\ref{alg:saccg} is achieved by setting $\beta:d\mapsto\texttt{True}$.
+It is not hard to see that in this case Algorithm~\ref{alg:saccg} reduces to
+Algorithm~\ref{alg:sac}. The worst case is achieved by setting $\beta: d\mapsto
+\texttt{False}$. In that case the while loop in Line~\ref{line:while} is
+iterated $n-1$ times (yielding total $n$ executions of $f^*$) while
+Line~\ref{line:R} is never executed.
 
+The analyses above leaves open the question of how the runtime depends on
+application of the pullback $f^*$. First note that there is no difference in
+applying the $f^*$ or $f^*\circ F_i^*\circ F_i$: in practice a nested function
+call is not necessary as the collapsing scheme is fully available as input. The
+pullback performs a lookup of records based on matching one or more attributes
+with a set of attribute values. The time complexity of such operations are
+typically reduced by proper preparation of the dataset. For example, in
+databases one may define indices that speed up certain often-used lookup
+operations (that then may rely on those indices). In the implementation of the
+\code{accumulate()} package the pullback is (currently) implemented using
+standard join operations as implemented by \proglang{R}'s standard
+\code{merge()} function.  This implies that the time complexity of
+\code{accumulate()} and \code{cumulate()} should be multiplied with the time
+dependence of \code{merge()}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Summary and conclusion}\label{sect:conclusion}
+The \proglang{R} package \code{accumulate} introduced in this paper offers
+convenient interfaces for computing grouped aggregates where the grouping is
+dynamically determined based on a user-defined conditions and a user-defined
+collapsing scheme. We demonstrated that these interfaces can be used in
+situations including multiple grouping variables and complex collapsing
+schemes, as well as situations where multiple variables need to be aggregated
+over. Moreover, the package supports the collection of complex aggregates, such
+as model outputs and offers several service functions that aim to facilitate
+the definition of the collapsing conditions.
 
-\begin{itemize}
-\item{Record-wise results}
-\item{Record-wise imputation}
-\item{Representation of hierarchical classifications: parent-child lists vs tabular}
-\end{itemize}
+Finally, the pseudocode underlying the package's main functions has been
+formally analyzed and it is shown that aggregation with collapsing groups can
+be interpreted as a generalization of standard grouped aggregation in a very
+precise sense. We hope that this stimulates broader implementation of this
+algorithm in software offering grouped aggregation.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage

### 2023-03-03 paper/article.Rnw
wip

@@
 with dynamic grouping.  We start by giving an algorithm for ordinary
-split-apply-combine to show where the algorithm must be generalized to allow
-for a collapsing scheme.
+split-apply-combine to demonstrate how the algorithm must be generalized to
+allow for a collapsing scheme.
 
@@
 
-Denote with $U$ be a finite set, and and let $\phi:2^U\to X$ be a function that
+Denote with $U$ be a finite set, and let $\phi:2^U\to X$ be a function that
 accepts a subset of $U$ and returns a value in some domain $X$.  Here, $U$
@@
 We split $U$ into groups using the following notation. Let $A$ be finite set
-that has no more elements than $U$, and let $f:\onto A$ be a
+that has no more elements than $U$, and let $f:U\onto A$ be a
 surjective function that takes an element of $U$ and returns a value in $A$. We
@@
 In this algorithm the output is collected in a set $R$ containing pairs from
-$A\times X$: one pair for each element of $A$. Incidentely, the algorithm can
-be summarized even shorter in this notation as $R=\cup_{a\in A}\{(a,(\phi\circ
+$A\times X$: one pair for each element of $A$. Incidently, the algorithm can be
+summarized even shorter in this notation as $R=\cup_{a\in A}\{(a,(\phi\circ
 f^*)(\{a\}))\}$, where $\circ$ denotes function composition.  Since $f^*$
-always receives a singleton set, its definition as a function of the powerset
-of the group labels $A$ is too general for this algorithm.  However, this
-generalization turns out to be necessary when generalizing the algorithm to the
-case of dynamic grouping.
+always receives a singleton set, its definition as a pullback along $f$ more
+general then necessary for this algorithm.  However, this formulation turns out
+to be necessary when generalizing the algorithm to the case of dynamic
+grouping.
 

### 2023-03-14 paper/article.Rnw
wip

@@
 @
-Using \code{accumulate()} we obtain the means. 
+Using \code{cumulate()} we obtain the means. 
 <<>>=
@@
 
-\subsection{Imputing missing values using SAE and ratio imputation} 
+\subsection{Imputing missing values using SAE and ratio imputation} \label{sect:ratio}
 Our goal in this example is to impute missing values for the \code{industrial}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Random Nearest neighbours imputation with collapsing groups}
+Nearest neighbor (NN) imputation is a donor imputation method where the
+imputation value is copied from a (possibly random) donor record with
+attributes that are sufficiently similar \citep{andridge2010review}.
+One way to generate an imputation value is to select a random donor
+from the same group. To prevent the same donor from being used too
+often, it is not uncommon to demand a minimum number of records
+in the donor pool. A collapsing scheme is one way of guaranteeing
+this, and below we demonstrate how this problem can be expressed
+in \code{accumulate}.
+
+We wish to impute the variable \code{trade} in the \code{producers} dataset
+using donor imputation, where donors come from the same (\code{sbi},
+\code{size}) combination. We wish to sample donor values from a group of at
+least 5 donors. If this is not possible, we use the same fallback scenario as
+in the previous Section.
+
+We first define an `aggregation' function that takes a vector 
+of donors and returns a non-empty sample.
+<<>>=
+random_element <- function(x) sample(x[!is.na(x)], 1)
+@
+To make sure we obtain a donor for each record, we add an
+identifying column to use as grouping variable.
+\code{id} column.
+<<>>=
+producers <- cbind(id = sprintf("ID%03d",seq_len(nrow(producers)))
+                  , producers)
+set.seed(111)
+imputations <- cumulate(producers
+                 , collapse = id ~ sbi*size + sbi + sbi2 + sbi1
+                 , test = min_complete(5,"trade")
+                 , donor_trade = random_element(trade))
+
+head(imputations, 3)
+@
+To use the donor imputations, we merge the imputation candidates with the
+original dataset and use \code{impute_proxy()} of the \pkg{simputation} package
+for imputation.
+<<>>=
+imputed <- merge(producers, imputations) |>
+           impute_proxy(trade ~ donor_trade)
+cols <- c(1:3,9:10,5)
+head(producers[cols],3)
+head(imputed[c(cols,11)],3)
+@
+The merge operation automatically merges on the \code{id} column, which also
+adds the \code{level} column to the output.  The function \code{impute_proxy}
+copies values from \code{donor_trade} into trade where \code{trade} is missing.
+In the last expressions we only print the columns of interest.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Computing complex aggregates} \label{sect:complex}
-Generalizing further, one may be interested in estimating (statistical) models.
-With the \code{cumulate} function it is possible to specify complex aggregates
-that go beyond a simple number. Below, we estimate the following linear model
+Until now, the aggregates have been simple (scalar) values.  With the
+\code{cumulate()} function it is also possible to specify complex aggregates that
+go beyond simple aggregates. Below, we estimate the following linear model
 \begin{displaymath}

### 2023-03-14 paper/article.Rnw
wip

@@
 imputation is a method where the imputed value $\hat{y}_i$ for variable $Y$ of
-record $i$ is estimated as $\hat{y_i}=\hat{R}_d x_i$, where $\hat{R}$ is an
+record $i$ is estimated as $\hat{y_i}=\hat{R}_d x_i$, where $\hat{R}_d$ is an
 estimate of the ratio between $Y$ and an auxiliary variable $X$ in
@@
 Nearest neighbor (NN) imputation is a donor imputation method where the
-imputation value is copied from a (possibly random) donor record with
-attributes that are sufficiently similar \citep{andridge2010review}.
-One way to generate an imputation value is to select a random donor
-from the same group. To prevent the same donor from being used too
-often, it is not uncommon to demand a minimum number of records
-in the donor pool. A collapsing scheme is one way of guaranteeing
-this, and below we demonstrate how this problem can be expressed
-in \code{accumulate}.
+imputation value is copied from a record that is (randomly) chosen from a donor
+pool \citep{andridge2010review}. In this example we use the grouping variables
+in \code{producers} to define the donor pools. To prevent the same donor from
+being used too often, it is not uncommon to demand a minimum number of records
+in the donor pool. A collapsing scheme is one way of guaranteeing this, and
+below we demonstrate how this problem can be expressed in \code{accumulate}.
 
@@
 @
+We will use \code{cumulate()} to ensure that there are at least five non-empty
+values in \code{x} when \code{random_element()} is called.
+
 To make sure we obtain a donor for each record, we add an

### 2023-07-24 paper/article.Rnw
updated paper

@@
 determined, based on user-defined conditions on subsets, and a user-defined
-subset collapsing scheme. Finally, we demonstrate and analyze the pseudocode
-underlying this method.
+subset collapsing scheme. The formal underlaying algorithm is described and
+analyzed as well.
 }
@@
 grouping records and aggregating over one or more columns in data frame
-objects. Similarly, the \pkg{DataFrames} package for \proglang{julia}
-implements such functionality \citep{kaminski2022dataframes}.
+objects. The more recent \citet{polars2023} library for \proglang{Python} and
+\proglang{Rust} also implements such features. Similarly, the \pkg{DataFrames}
+package for \proglang{julia} implements split-apply-combine functionality
+\citep{kaminski2022dataframes}.
 
@@
 for a group, for example a geographical region, or a detailed population
-subset, where the number of (sampled) observations so small that the variance
+subset, where the number of (sampled) observations is so small that the variance
 of the estimate would be unacceptably large. In small area estimation (SAE) one
@@
 split-apply-combine aggregation where groups can be collapsed dynamically in
-the form of an \pkg{R} package \pkg{accumulate} \cite{loo2022accumulate}.
+the form of \pkg{R} package \pkg{accumulate} \citep{loo2022accumulate}.
 
@@
 should resemble existing popular interfaces where possible. Second, users
-should be free to define any possibly multi-step, collapsing scheme. Here, we
+should be free to define any, possibly multi-step, collapsing scheme. Here, we
 keep in mind that collapsing schemes may be constructed manually based on
@@
 and straightforward generalization of the standard split-apply-combine problem
-and analyze its time complexity.
-
-We describe the fundamental algorithm in pseudocode and compare it with the
-standard grouping scenario. We also discuss the time and space complexity and
+and analyze its time complexity. A summary and conlcusion follows in
+Section~\ref{sect:conclusion}.
 
@@
 
-Our goal is to compute the sum $t_Y$ over $Y$, grouped by $A\times B$.  We
+Our goal is to compute the mean $\mu_Y$ over $Y$, grouped by $A\times B$.  We
 impose the condition that there must be at least three records in each group.
@@
 \hline
-$A$ & $B$ & $B1$ & $Y$\\
+$A$ & $B$ & $B_1$ & $Y$\\
 \hline
@@
 \cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
-\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
-\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
-\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
-\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
-\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
-\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
-\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
+\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   3   \\
+\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 4   \\
+\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 5   \\
+\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 6   \\
+\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  7   \\
+\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  8   \\
+\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  9   \\
 \hline
@@
 \hline
-$A\times B$ & \code{level} & $t_Y$\\
+$A\times B$ & \code{level} & $\mu_Y$\\
 \hline
-\rowcolor{red!25}   1 11        & 0             & 7  \\
-\rowcolor{green!25} 2 12        & 1             & 56 \\
-\rowcolor{green!25} 2 13        & 1             & 56 \\
-\rowcolor{blue!25}  3 21        & 2             & 448\\
-\rowcolor{blue!25}  3 22        & 2             & 448\\
-\rowcolor{blue!25}  3 12        & 2             & 448\\
+\rowcolor{red!25}   1 11        & 0             & 2  \\
+\rowcolor{green!25} 2 12        & 1             & 5  \\
+\rowcolor{green!25} 2 13        & 1             & 5  \\
+\rowcolor{blue!25}  3 21        & 2             & 8  \\
+\rowcolor{blue!25}  3 22        & 2             & 8  \\
+\rowcolor{blue!25}  3 12        & 2             & 8  \\
 \hline
@@
 is zero, which means that no collapsing was necessary. Indeed, in the data
-table we see that there are three rows with $A=1$ and $B==11$ with $Y$ values
-$1, 2$, and $7$, resulting in $t_Y=7$ for this group.
+table we see that there are three rows with $A=1$ and $B=11$ with $Y$ values
+$1, 2$, and $3$, resulting in $\mu_Y=(1+2+3)/3=2$ for this group.
 
@@
 find that there are three records in that group so we get 
-$t_Y=8+16+32=56$.  Similarly, there is only one record with $(A=2,B=13)$. 
-Collapsing groups to $(A=2,B1=1)$, yields again $t_y=56$.
+$\mu_Y=(4+5+6)/3=5$.  Similarly, there is only one record with $(A=2,B=13)$. 
+Collapsing groups to $(A=2,B_1=1)$, yields again $\mu_y=5$.
 
 Finally, for $(A=2,B=21)$ there is only a single record. Collapsing to
-$(A=2,B1=2)$ yields only two records, so we need to collapse further to $(A=2)$
-and finally obtain three records.  This yields $t_Y=64+128+256=448$. Similarly,
+$(A=2,B_1=2)$ yields only two records, so we need to collapse further to $(A=2)$
+and finally obtain three records.  This yields $\mu_Y=(7+8+9)/3=8$. Similarly
 the groups $(A=3, B=22)$ and $(A=3,B=12)$ are collapsed to $(A=2)$.
@@
 \code{data}, and the ellipsis (\code{...}) is for arguments that are passed as
-extra argument to \code{fun}. The interface of \code{accumulate()} is modeled
-after the \code{aggregate()} function in \proglang{R}.  In \code{cumulate}, the
-ellipsis is a sequence of comma-separated \code{name=expression} pairs in the
-style of \code{summarise()} from the \code{dplyr} package.
+extra argument to \code{fun}. The interface of \code{accumulate()} is somewhat
+similar to that of the \code{aggregate()} function in \proglang{R}.  In
+\code{cumulate()}, the ellipsis is a sequence of comma-separated
+\code{name=expression} pairs in the style of \code{summarise()} from the
+\code{dplyr} package.
 
@@
 schemes through the \code{collapse} parameter. The first and most general is
-the formula interface, which requires that the collapsing sequence is
+the \class{formula} interface, which requires that the collapsing sequence is
 represented as variables in the data set to be aggregated. The second is a
@@
 We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
-the formula interface.
+the \class{formula} interface.
 <<>>=
@@
    , B1 = c(1,1,1,1,1,1,2,2,1)
-   , Y  = 2^(0:8)
+   , Y  = 1:9
 )
@@
 cumulate(input, collapse = A*B ~ A*B1 + A
-        , test = function(d) nrow(d) >= 3, tY = sum(Y) )
+        , test = function(d) nrow(d) >= 3, muY = mean(Y) )
 @
@@
 
-Summarizing, the formula interface is always of the following form.
+Summarizing, the \class{formula} interface is always of the following form.
 \begin{verbatim}
@@
 <<>>=
-input$Y2 <- 3^(0:8)
+input$Y2 <- 11:19
 accumulate(input, collapse = A*B ~ A*B1 + A
-          , test = min_records(3), fun = sum)
+          , test = min_records(3), fun = mean)
 @
@@
 accumulate(input[-3], collapse = A*B ~ A
-          , test = min_records(3), fun = sum)
+          , test = min_records(3), fun = mean)
 @
 
-The formula interface allows users to quickly experiment with different
+The \class{formula} interface allows users to quickly experiment with different
 collapsing schemes. It does require that all categorical variables have been
@@
 <<>>=
-accumulate(input1, collapse=csh, test = min_records(3), sum)
-cumulate(input1, collapse=csh, test=min_records(3), tY=sum(Y), tY2 = sum(Y2))
+accumulate(input1, collapse=csh, test = min_records(3), mean)
+cumulate(input1, collapse=csh, test=min_records(3), muY=mean(Y), muY2 = mean(Y2))
 @
@@
 \subsection{Specifying tests}
-The \code{test} parameter of \code{cumulate()} and \code{accumulate} accepts a
+The \code{test} parameter of \code{cumulate()} and \code{accumulate()} accepts a
 function that takes a subset of the data and must returns a boolean. For common
@@
 accumulate(input, collapse = A*B ~ A*B1 + B1,
-    test = from_validator(rules), fun=sum)
+    test = from_validator(rules), fun=mean)
 @
-Note that for target groups $(A=2,B=21)$ and $(A=2,B=22)$ none of the available
+Note that for target groups $(A=3,B=21)$ and $(A=3,B=22)$ none of the available
 collapsing levels lead to a group that satisfied all conditions. Therefore the
@@
 my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >=3
-accumulate(input, collapse = A*B ~ A*B1 + B1, test=my_test, fun=sum)
+accumulate(input, collapse = A*B ~ A*B1 + B1, test=my_test, fun=mean)
 @
@@
 the occurrence of missing values, a dataset with zero rows, and the full
-dataset. The smoke test checks whether the output is \code{TRUE} or \code{FALSE}
+dataset. The function \code{smoke_test()} checks whether the output is \code{TRUE} or \code{FALSE}
 under all circumstances and also reports errors, warnings, and messages.
@@
 @
-By default only failing tests are printed. In this case our test function
-is not robust against missing values for $Y$. This can be remedied
-by adding \code{na.rm=TRUE} to the \code{sum()} in the test function.
+By default only failing tests are printed. In this case our test function is
+not robust against missing values for $Y$. This can be remedied by passing
+\code{na.rm=TRUE} as parameter to \code{sum()} in the test function.
 <<>>=
@@
 \end{center}
-Here, the hierarchy suggest to collapsing $\{0111,0112,0113\}$ into $\{011\}$
-(when needed) and similar for the right branch. The second level of collapsing
+Here, the hierarchy suggests to collapse $\{0111,0112,0113\}$ into $\{011\}$
+when needed, and similarly for the right branch. The second level of collapsing
 would combine $\{012\}$ with $\{011\}$ into $\{01\}$. The \pkg{accumulate}
@@
 computed. Since all codes are prepended with zero, there is no need to collapse
-01 any further. The output can be used as argument to the \code{collapse}
-parameter of the \code{accumulate()} or \code{cumulate} functions.
+$01$ any further. The output can be used as argument to the \code{collapse}
+parameter of the \code{accumulate()} or \code{cumulate()} functions.
 
@@
 \section{Extensive example: economic data} \label{sect:example}
-In this section we treat three practical examples using a synthetic
+In this section we discuss three practical examples using a synthetic
 dataset included with the package.
@@
 We will use \code{cumulate()} to ensure that there are at least five non-empty
-values in \code{x} when \code{random_element()} is called.
-
-To make sure we obtain a donor for each record, we add an
-identifying column to use as grouping variable.
-\code{id} column.
+values in \code{x} when \code{random_element()} is called. To make sure we
+obtain a donor for each record, we add an identifying column \code{id} to use
+as grouping variable.  
 <<>>=
@@
 
-Denote with $U$ be a finite set, and let $\phi:2^U\to X$ be a function that
+Denote with $U$ a finite set, and let $\phi:2^U\to X$ be a function that
 accepts a subset of $U$ and returns a value in some domain $X$.  Here, $U$
@@
 \end{equation*}
-where $S$ is a subset of $A$ \citep[Section 1.4]{fong2019invitation}.
+where $S$ is a subset of $A$ (See e.g. \citet[Section 1.4]{fong2019invitation}).
 
@@
 \begin{algorithm}[H]
-\caption{Split-Apply-Combine}
+\caption{Split-Apply-Combine: $\textsc{SAC}(U,\phi,f)$}
 \label{alg:sac}
@@
 In this algorithm the output is collected in a set $R$ containing pairs from
-$A\times X$: one pair for each element of $A$. Incidently, the algorithm can be
+$A\times X$: one pair for each element of $A$. (Incidently, the algorithm can be
 summarized even shorter in this notation as $R=\cup_{a\in A}\{(a,(\phi\circ
-f^*)(\{a\}))\}$, where $\circ$ denotes function composition.  Since $f^*$
-always receives a singleton set, its definition as a pullback along $f$ more
-general then necessary for this algorithm.  However, this formulation turns out
-to be necessary when generalizing the algorithm to the case of dynamic
-grouping.
+f^*)(\{a\}))\}$, where $\circ$ denotes function composition).
 
-It is interesting to see how the elements $U$, $f:U\onto A$, and
-$\phi$ are implemented in practice. Consider the signature of the
-\proglang{R}'s \code{aggregate()} function (we skip arguments that are not
-important for the discussion).
+It is interesting to see how the elements $U$, $f$, and $\phi$ are implemented
+in practice. Consider the signature of the \proglang{R}'s \code{aggregate()}
+function (we skip arguments that are not important for the discussion).
 \begin{verbatim}
@@
 \end{verbatim}
-Here, \code{x} is a data frame where each column is a variable $U$ to be
-aggregated.  The parameter \code{by} is a list of vectors of group labels,
-where each vector has a with a length that equals the number of rows in
-\code{x}. So the function $f:U\onto A$ is implemented by asking
-the user to make sure that the position of each label corresponds to the
-correct row number in the data frame \code{x}. The argument \code{FUN} is the
-function $\phi$ that aggregates each subset of \code{x}.  Here is an example of
-how a user might call this function from the R prompt.
+Here, \code{x} is a data frame where each row represents an element of $U$. The
+parameter \code{by} is a list of vectors of group labels, where each vector has
+a length that equals the number of rows in \code{x}. So the function $f:U\onto
+A$ is implemented by asking the user to make sure that the position of each
+label in \code{by} corresponds to the correct row number in the data frame
+\code{x}. The argument \code{FUN} (of class \class{function}) represents the
+function $\phi$ that aggregates each subset of \code{x}. When the records in
+\code{x} contain more than one variable, the aggregator is applied to each one
+of them.  Here is an example of how a user might call this function from the
+\proglang{R} prompt.
 %
@@
 The goal of the algorithm is to compute a value for each part of a dataset,
-possibly using values external to the part. The input of the algorithm consists
-again of a finite set $U$ and an aggregation function $\phi$ that takes a
-subset of $U$ and returns a value in some domain $X$.  Compared to
-Algorithm~\ref{alg:sac} two different inputs are needed. First, a function must
-be defined that that checks whether a given subset $d$ of $U$ is suitable for
-computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
-where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
-checking whether there are sufficient records available, or whether certain
-variables have a low enough fraction of missing values. Second, we need a
-\emph{collapsing scheme}
+possibly using values external to the part ---conditional to restrictions
+placed on each part. The input of the algorithm consists again of a finite set
+$U$ and an aggregation function $\phi$ that takes a subset of $U$ and returns a
+value in some domain $X$.  Compared to Algorithm~\ref{alg:sac} two other inputs
+are needed. First, a function must be defined that that checks whether a given
+subset $d$ of $U$ is suitable for computing $\phi(d)$.  We will denote this
+function $\beta: 2^U\to \mathbb{B}$, where
+$\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are checking
+whether there are sufficient records available, or whether certain variables
+have a low enough fraction of missing values. Second, we need a
+\emph{collapsing scheme} $C$, defined as sequence of $n+1$ mappings
 \begin{equation}
-U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
+C\equiv U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
 \label{eq:collapsingsequence}
@@
 \begin{algorithm}[H]
-\caption{Split-Apply-Combine with Collapsing Groups}
+\caption{Split-Apply-Combine with Collapsing Groups: $\textsc{SACCG}(U,\phi,\beta,C)$}
 \label{alg:saccg}
@@
 \Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
-      and a collapsing sequence $U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots \xonto{f_n}  A_n$.}
+      and a collapsing sequence $C\equiv U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots \xonto{f_n}  A_n$.}
 
@@
 Algorithm~\ref{alg:saccg} is achieved by setting $\beta:d\mapsto\texttt{True}$.
-It is not hard to see that in this case Algorithm~\ref{alg:saccg} reduces to
-Algorithm~\ref{alg:sac}. The worst case is achieved by setting $\beta: d\mapsto
+In this case Algorithm~\ref{alg:saccg} reduces to Algorithm~\ref{alg:sac}. In
+other words, we have
+%
+\begin{displaymath}
+\textsc{SACCG}(U,\phi,\texttt{True},C) = \textsc{SAC}(U,\phi,f).
+\end{displaymath}
+%
+To see this, observe that in Algorithm~\ref{alg:saccg} the condition in
+line~\ref{line:while} is always \code{False} and the condition in
+line~\ref{line:R} is always \code{True} in this case.
+
+The worst case is achieved by setting $\beta: d\mapsto
 \texttt{False}$. In that case the while loop in Line~\ref{line:while} is
@@
 The analyses above leaves open the question of how the runtime depends on
-application of the pullback $f^*$. First note that there is no difference in
-applying the $f^*$ or $f^*\circ F_i^*\circ F_i$: in practice a nested function
-call is not necessary as the collapsing scheme is fully available as input. The
-pullback performs a lookup of records based on matching one or more attributes
-with a set of attribute values. The time complexity of such operations are
-typically reduced by proper preparation of the dataset. For example, in
-databases one may define indices that speed up certain often-used lookup
-operations (that then may rely on those indices). In the implementation of the
-\code{accumulate()} package the pullback is (currently) implemented using
-standard join operations as implemented by \proglang{R}'s standard
-\code{merge()} function.  This implies that the time complexity of
-\code{accumulate()} and \code{cumulate()} should be multiplied with the time
-dependence of \code{merge()}.
+application of the pullbacks in both algorithms. First note that there needs to
+be no difference in applying $f^*$ or $f^*\circ F_i^*\circ F_i$: in practice a
+collapsing scheme can be represented in tabular form just like $f$.  The
+pullback then comes down to a lookup of records based on matching one or more
+attributes with a set of attribute values. The time complexity of such
+operations are typically reduced by proper preparation of the dataset. For
+example, databases can be prepared to speed up certain often-used lookup
+operations. In the implementation of the \code{accumulate} package the
+pullback is implemented using standard join operations as
+implemented by \proglang{R}'s standard \code{merge()} function.
 
@@
 convenient interfaces for computing grouped aggregates where the grouping is
-dynamically determined based on a user-defined conditions and a user-defined
-collapsing scheme. We demonstrated that these interfaces can be used in
-situations including multiple grouping variables and complex collapsing
-schemes, as well as situations where multiple variables need to be aggregated
-over. Moreover, the package supports the collection of complex aggregates, such
-as model outputs and offers several service functions that aim to facilitate
-the definition of the collapsing conditions.
+dynamically determined based on user-defined conditions and a user-defined
+group collapsing scheme. We demonstrated that these interfaces can be used in
+several situations, including those where multiple grouping variables are used
+, where complex collapsing schemes are applied, as well as situations where
+multiple variables need to be aggregated over. Conditions on groups of records
+are principally represented by a function, but the package includes several
+convenience functions for defining conditions on record groups. It also
+interfaces with the \pkg{validate} \proglang{R} package to facilitate cases
+where mutliple test conditions must be met. Moreover, the package supports the
+collection of complex aggregates, such as model outputs and offers several
+service functions that aim to facilitate the definition of the collapsing
+conditions.
 
-Finally, the pseudocode underlying the package's main functions has been
-formally analyzed and it is shown that aggregation with collapsing groups can
-be interpreted as a generalization of standard grouped aggregation in a very
-precise sense. We hope that this stimulates broader implementation of this
-algorithm in software offering grouped aggregation.
+The pseudocode underlying the package's main functions has been formally
+analyzed and it is shown that aggregation with collapsing groups can be
+interpreted as a precise generalization of standard grouped aggregation.  We
+hope that this stimulates broader implementation of this algorithm in software
+offering grouped aggregation.
 

### 2023-09-22 paper/setup.R
added paper stuff

@@ -1,0 +1,2 @@
+# to setup your code before running article.R, use on the R prompt
+install.packages("accumulate_0.9.0.tar.gz")

### 2023-09-22 pkg/R/helpers.R
fixed example

@@ -114,7 +114,7 @@ frac_complete <- function(r, vars=TRUE){
 #'
 #' @examples
 #'
-#' if (requireNamespace("validate", quitetly=TRUE)){
+#' if (requireNamespace("validate", quietly=TRUE)){
 #'  v <- validator(height >= 0, weight >= 0)
 #'  f <- from_validator(v)
 #'  f(women)  # TRUE (all heights and weights are nonnegative)

### 2023-09-22 pkg/R/helpers.R
several doc improvements

@@ -1,6 +1,6 @@
 
 
-#' Demand minimal number of functons
+#' Demand minimal number of records 
 #' 
 #' @param n Minimal number of records in a group.
 #'
@@ -90,10 +90,11 @@ frac_complete <- function(r, vars=TRUE){
   function(d) mean(complete_cases(d[vars])) >= r
 }
 
-#' Use a validate::validator object to test a group
+#' Use a validate::validator object to define a test
 #'
-#' Returns \code{TRUE} when the data passes all checks defined in the
-#' \code{validator} object, otherwise \code{FALSE}.
+#' Create a test function that accepts a data.frame, and returns \code{TRUE}
+#' when the data passes all checks defined in the \code{validator} object, and
+#' otherwise \code{FALSE}.
 #'
 #' @param v \code{[validator]} a validator object from the
 #'        \code{validate} package.
@@ -133,7 +134,7 @@ from_validator <- function(v,...){
 
 
 
-#' Derive collapsing scheme from hierarchical classification
+#' Derive collapsing scheme from a hierarchical classification
 #'
 #' Derive a collapsing scheme where group labels collapse to their
 #' parents in the hierarchy.
@@ -170,13 +171,13 @@ csh_from_digits <- function(x, levels=max(nchar(x))-1){
   as.data.frame(A)[1:(levels+1)]
 }
 
-#' Test your testing function for common edge cases
+#' Check your testing function against common edge cases
 #'
 #' Writing a testing function that works on any subset of records of a
 #' dataframe can be quite subtle. This function tries the testing function on a
 #' number of common (edge) cases that are easily overlooked.  It is \emph{not}
 #' a unit test: a smoke test will not tell you whether your output is correct.
-#' It only checks the output data type (mustbe \code{TRUE} or \code{FALSE} and
+#' It only checks the output data type (must be \code{TRUE} or \code{FALSE} and
 #' reports if errors, warnings, or messages occur.
 #'
 #' @param dat an example dataset. For example the full dataset

### 2023-09-22 pkg/R/utils.R

@@ -6,7 +6,10 @@ last <- function(x) x[length(x)]
 
 complete_cases <- function(d) !Reduce(`|`, lapply(d, is.na))
 
-#' create a classed list
+#' Create a classed list
+#'
+#' Classed lists are used to pretty-print a list that is stored
+#' in a data frame.
 #'
 #' @param x a list
 #'
@@ -16,16 +19,19 @@ complete_cases <- function(d) !Reduce(`|`, lapply(d, is.na))
 #' object_list(list(lm(speed ~ dist, data=cars)))
 #'
 #' @export
+#' @keywords internal
 object_list <- function(x) structure(x, class=c("object_list","list"))
 
 #' @rdname object_list
 #' @export
+#' @keywords internal
 format.object_list <- function(x,...){
   sapply(x, function(u) sprintf("<%s>",paste(class(u),collapse=",")))
 }
 
 #' @rdname object_list
 #' @export
+#' @keywords internal
 print.object_list <- function(x,...) print(format.object_list(x,...))
 
 #' @rdname object_list

### 2023-09-22 paper/article.Rnw
passes R CMD check --as-cran

@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
-<<preliminaries, echo=FALSE, results=hide>>=
+\SweaveOpts{engine = R, eps = FALSE, keep.source = TRUE}
+<<preliminaries, echo=FALSE, results = hide>>=
 options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
@@
 determined using attributes from out-of-group entities. One example where this
-occurs is in the area of Small Area Estimation (SAE, see \emph{e.g.}
+occurs is in the area of Small Area Estimation (SAE, see \emph{e.g.},
 \citet{rao2015small, molina2015sae}). Here, one wishes to estimate an aggregate
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Dynamic grouping} \label{sect:dynamicgrouping}
+\section{Dynamic Grouping} \label{sect:dynamicgrouping}
 In this section the illustrate the concept of dynamic grouping with a minimal
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5%%%%%%%
-\section{The accumulate R package}\label{sect:accumulate}
+\section[R package accumulate]{\proglang{R Package} \pkg{accumulate} }\label{sect:accumulate}
 Grouped aggregation with a fall-though scenario based on a collapsing scheme
@@
 \code{cumulate()}, the ellipsis is a sequence of comma-separated
-\code{name=expression} pairs in the style of \code{summarise()} from the
+\code{name = expression} pairs in the style of \code{summarise()} from the
 \code{dplyr} package.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The formula interface}\label{sect:formula}
+\subsection{The Formula Interface}\label{sect:formula}
 We will use the example of Section~\ref{sect:dynamicgrouping} to illustrate
@@
 <<>>=
-library(accumulate)
+library("accumulate")
 input <- data.frame(
-     A  = c(1,1,1,2,2,2,3,3,3)
-   , B  = c(11,11,11,12,12,13,21,22,12)
-   , B1 = c(1,1,1,1,1,1,2,2,1)
+     A  = c( 1,  1,  1,  2,  2,  2,  3,  3,  3)
+   , B  = c(11, 11, 11, 12, 12, 13, 21, 22, 12)
+   , B1 = c( 1,  1,  1,  1,  1,  1,  2,  2,  1)
    , Y  = 1:9
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The data frame interface} \label{sect:dfint}
+\subsection{The Data Frame Interface} \label{sect:dfint}
 The data frame interface is somewhat limited because it only allows for a
@@
 input1 <- input
-input1$AB <- paste(input$A, input$B, sep="-")
+input1$AB <- paste(input$A, input$B, sep = "-")
 input1 <- input1[-(1:3)]
@@
 csh <- data.frame(
-  AB  = c("1-11", "2-12","2-13","3-21","3-22","3-12")
-, AB1 = c("1-1" , "2-1" ,"2-1" ,"3-2" ,"3-2" ,"3-1" )
-, A   = c("1"   , "2"   ,"2"   ,"3"   ,"3"   ,"3"   ))
+  AB  = c("1-11", "2-12", "2-13", "3-21", "3-22", "3-12")
+, AB1 = c("1-1" , "2-1" , "2-1" , "3-2" , "3-2" , "3-1" )
+, A   = c("1"   , "2"   , "2"   , "3"   , "3"   , "3"   ))
 csh
@@
 relation. For example, in the first collapsing step the groups defined by
-\code{AB == "2-12"} and \code{AB=="2-13"} both collapse to \code{AB1 == "2-1"}.
+\code{AB == "2-12"} and \code{AB == "2-13"} both collapse to \code{AB1 == "2-1"}.
 In this artificial example the codes do not mean anything, but in realistic
@@
 <<>>=
-accumulate(input1, collapse=csh, test = min_records(3), mean)
-cumulate(input1, collapse=csh, test=min_records(3), muY=mean(Y), muY2 = mean(Y2))
+accumulate(input1, collapse = csh, test = min_records(3), mean)
+cumulate(input1, collapse = csh, test = min_records(3)
+        , muY = mean(Y), muY2 = mean(Y2))
 @
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Specifying tests}
+\subsection{Specifying Tests}
 The \code{test} parameter of \code{cumulate()} and \code{accumulate()} accepts a
@@
 \code{min_records(n)}        & At least $n$ records.\\ 
-\code{min_complete(n,vars)}  & At least $n$ records complete for variables \code{vars}.\\
-\code{frac_complete(r,vars)} & At least $100r$\% complete records for variables \code{vars}.\\
-\code{from_validator(v,...)} & Construct a testing function from a \code{validator} object
+\code{min_complete(n, vars)}  & At least $n$ records complete for variables \code{vars}.\\
+\code{frac_complete(r, vars)} & At least $100r$\% complete records for variables \code{vars}.\\
+\code{from_validator(v, ...)} & Construct a testing function from a \code{validator} object
                                of \proglang{R} package \pkg{validate}.\\
@@
 <<>>=
-library(validate)
+library("validate")
 rules <- validator(nrow(.) >= 3, sum(Y >= 2) >= 3)
@@
 accumulate(input, collapse = A*B ~ A*B1 + B1,
-    test = from_validator(rules), fun=mean)
+    test = from_validator(rules), fun = mean)
 @
@@
 <<>>=
-my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >=3
-accumulate(input, collapse = A*B ~ A*B1 + B1, test=my_test, fun=mean)
+my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >= 3
+accumulate(input, collapse = A*B ~ A*B1 + B1, test = my_test, fun = mean)
 @
@@
 not robust against missing values for $Y$. This can be remedied by passing
-\code{na.rm=TRUE} as parameter to \code{sum()} in the test function.
+\code{na.rm = TRUE} as parameter to \code{sum()} in the test function.
 <<>>=
-my_test1 <- function(d) nrow(d) >= 3 && sum(d$Y>=2, na.rm=TRUE) >=3 
+my_test1 <- function(d) nrow(d) >= 3 && sum(d$Y >= 2, na.rm = TRUE) >= 3 
 smoke_test(input, my_test1)
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Balanced and unbalanced hierarchical classifications}
+\subsection{Balanced and Unbalanced Hierarchical Classifications}
 Hierarchical classifications are abundant in (official) statistics.
@@
 <<>>=
-nace <- c("0111","0112","0113","0121","0121","0122","0123","0124")
-csh_from_digits(nace, levels=2)
+nace <- c("0111", "0112", "0113", "0121"
+         , "0121", "0122", "0123", "0124")
+csh_from_digits(nace, levels = 2)
 @ 
@@
 <<>>=
-nace <- c("0111","0112","0113","0121","0122","0123","01241","01242")
-csh_from_digits(nace,levels=3)
+nace <- c("0111", "0112", "0113", "0121", "0122", "0123", "01241", "01242")
+csh_from_digits(nace, levels = 3)
 @
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extensive example: economic data} \label{sect:example}
+\section{Extensive Example: Economic Data} \label{sect:example}
 In this section we discuss three practical examples using a synthetic
@@
 <<>>=
-data(producers)
+data("producers")
 head(producers)
@@
 
-\subsection{Small area estimation} 
+\subsection{Small Area Estimation} 
 Small area estimation (SAE) is a collection of methods that are aimed at
@@
 <<>>=
-producers$sbi2 <- substr(producers$sbi,1,2)
-producers$sbi1 <- substr(producers$sbi,1,1)
+producers$sbi2 <- substr(producers$sbi, 1, 2)
+producers$sbi1 <- substr(producers$sbi, 1, 1)
 head(producers, 3)
@@
              , collapse = sbi*size ~ sbi + sbi2 + sbi1
-             , test = min_complete(n=10, vars="industrial")
-             , mean_industrial = mean(industrial, na.rm=TRUE))
+             , test = min_complete(n = 10, vars = "industrial")
+             , mean_industrial = mean(industrial, na.rm = TRUE))
 head(a,3)
@@
 
-\subsection{Imputing missing values using SAE and ratio imputation} \label{sect:ratio}
+\subsection{Imputing Missing Values Using SAE and Ratio Imputation} \label{sect:ratio}
 Our goal in this example is to impute missing values for the \code{industrial}
@@
          , collapse = sbi*size ~ sbi + sbi2 + sbi1
-         , test = min_complete(n=10, vars="industrial")
-         , R = mean(industrial, na.rm=TRUE)/mean(total, na.rm=TRUE))
+         , test = min_complete(n = 10, vars = "industrial")
+         , R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
 head(r,3)
@@
 <<>>=
-library(simputation)
+library("simputation")
 dat <- merge(producers, r)
@@
 iNA <- is.na(producers$industrial)
-head(dat[iNA, c("sbi","size","level","R","industrial","total")])
+head(dat[iNA, c("sbi", "size", "level", "R", "industrial", "total")])
 @
-From this, we have all the information to interpret the imputed values. For
+From this, we get all the information to interpret the imputed values. For
 example, we see that in record 664, the ratio was estimated after collapsing
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Random Nearest neighbours imputation with collapsing groups}
+\subsection{Random Nearest Neighbours Imputation with Collapsing Groups}
 Nearest neighbor (NN) imputation is a donor imputation method where the
@@
 <<>>=
-random_element <- function(x) sample(x[!is.na(x)], 1)
+random_element <- function(x) sample(x[ !is.na(x)], 1)
 @
@@
 <<>>=
-producers <- cbind(id = sprintf("ID%03d",seq_len(nrow(producers)))
+producers <- cbind(id = sprintf("ID%03d", seq_len(nrow(producers)))
                   , producers)
@@
            impute_proxy(trade ~ donor_trade)
-cols <- c(1:3,9:10,5)
-head(producers[cols],3)
-head(imputed[c(cols,11)],3)
+cols <- c(1:3, 9:10, 5)
+head(producers[cols], 3)
+head(imputed[c(cols, 11)], 3)
 @
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computing complex aggregates} \label{sect:complex}
+\subsection{Computing Complex Aggregates} \label{sect:complex}
 Until now, the aggregates have been simple (scalar) values.  With the
@@
          , collapse = sbi*size ~ sbi + sbi2 + sbi1
-         , test = min_complete(n=10, vars=c("total","industrial"))
+         , test = min_complete(n = 10, vars = c("total","industrial"))
          , model = lm(total ~ industrial))
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Formal description and algorithms} \label{sect:algorithm}
+\section{Formal Description and Algorithms} \label{sect:algorithm}
 In this section we give a formal description of the algorithm for aggregation
@@
 
-\subsection{Split-apply-combine}
+\subsection{Split-Apply-Combine}
 To analyse a data set group by group we need to specify a data set, a
@@
 \end{equation*}
-where $S$ is a subset of $A$ (See e.g. \citet[Section 1.4]{fong2019invitation}).
+where $S$ is a subset of $A$ (See \emph{e.g.}, \citet[Section 1.4]{fong2019invitation}).
 
@@
 \For{$a\in A$}{
-  $d=f^*(\{a\})$\tcp*{get subset of $U$}
+  $d = f^*(\{a\})$\tcp*{get subset of $U$}
   $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
@@
 <<>>=
-aggregate(iris[1:2], by=iris["Species"], FUN=mean)
+aggregate(iris[1:2], by = iris["Species"], FUN = mean)
 @
@@
 
-\subsection{Split-apply-combine with collapsing groups}
+\subsection{Split-Apply-Combine with Collapsing Groups}
 \label{sect:saccg}
@@
 that are mapped to those labels via the collapsing sequence of
-Equation~\eqref{eq:collapsingsequence}.  With this notation we can define the
+Equation~\ref{eq:collapsingsequence}.  With this notation we can define the
 Algorithm for aggregation with dynamic grouping as follows.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary and conclusion}\label{sect:conclusion}
+\section{Summary and Conclusion}\label{sect:conclusion}
 The \proglang{R} package \code{accumulate} introduced in this paper offers

### 2023-09-22 pkg/R/helpers.R

@@ -116,7 +116,7 @@ frac_complete <- function(r, vars=TRUE){
 #' @examples
 #'
 #' if (requireNamespace("validate", quietly=TRUE)){
-#'  v <- validator(height >= 0, weight >= 0)
+#'  v <- validate::validator(height >= 0, weight >= 0)
 #'  f <- from_validator(v)
 #'  f(women)  # TRUE (all heights and weights are nonnegative)
 #' }

### 2023-09-22 pkg/R/producers.R

@@ -1,5 +1,5 @@
 #' @name producers
-#' @title synthetic data on producers
+#' @title Synthetic data on producers
 #' @description 
 #'   A synthetic dataset listing several sources of turnover
 #'   and other income for producers. The producers are classified

### 2023-09-22 pkg/R/helpers.R
fixing CRAN note

@@ -243,15 +243,12 @@ try_this <- function(d, f, verbose, info){
          )
     , error   = function(e) err <<- append(err, e$message)
   )
-  print(smoke(list(result=out, msg=msg, wrn=wrn, err=err, info=info)),verbose=verbose)
+  print_smoke(list(result=out, msg=msg, wrn=wrn, err=err, info=info), verbose=verbose)
   invisible(length(err) == 0)
 }
 
-smoke <- function(x){
-  structure(x,class="smoke")
-}
 
-print.smoke <- function(x, verbose){
+print_smoke <- function(x, verbose){
   if (isTRUE(x$result)|| isFALSE(x$result)){
     if(verbose) catf("\nTest with %s: OK", x$info)
   } else {

### 2023-09-22 pkg/R/utils.R

@@ -36,6 +36,7 @@ print.object_list <- function(x,...) print(format.object_list(x,...))
 
 #' @rdname object_list
 #' @export
+#' @keywords internal
 `[.object_list` <- function(x,i,j,...,drop=TRUE){
   object_list(unclass(x)[i])
 }

### 2023-10-03 paper/article.Rnw
added spaces around *

@@
 
-cumulate(input, collapse = A*B ~ A*B1 + A
+cumulate(input, collapse = A * B ~ A * B1 + A
         , test = function(d) nrow(d) >= 3, muY = mean(Y) )
@@
 %
-Consider the formula \code{A*B ~ A*B1 + A} in the call to \code{cumulate()}.
-The left-hand-side \code{A*B} is the target output grouping. The
+Consider the formula \code{A * B ~ A * B1 + A} in the call to \code{cumulate()}.
+The left-hand-side \code{A * B} is the target output grouping. The
 right-hand-side is to be interpreted as the collapsing sequence: if an instance
-of \code{A*B} does not pass the test, then collapse to \code{A*B1}, and if that
+of \code{A * B} does not pass the test, then collapse to \code{A * B1}, and if that
 does not pass the test collapse to \code{A}. If this final grouping also does
@@
 input$Y2 <- 11:19
-accumulate(input, collapse = A*B ~ A*B1 + A
+accumulate(input, collapse = A * B ~ A * B1 + A
           , test = min_records(3), fun = mean)
@@
 <<>>=
-accumulate(input[-3], collapse = A*B ~ A
+accumulate(input[-3], collapse = A * B ~ A
           , test = min_records(3), fun = mean)
@@
 @
-We use \code{A*B ~ A*B1 + B1} as collapsing scheme. The function
+We use \code{A * B ~ A * B1 + B1} as collapsing scheme. The function
 \code{from_validator} passes the requirements as a test function to
@@
 <<>>=
-accumulate(input, collapse = A*B ~ A*B1 + B1,
+accumulate(input, collapse = A * B ~ A * B1 + B1,
     test = from_validator(rules), fun = mean)
@@
 my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >= 3
-accumulate(input, collapse = A*B ~ A*B1 + B1, test = my_test, fun = mean)
+accumulate(input, collapse = A * B ~ A * B1 + B1, test = my_test, fun = mean)
 @
@@
 records for which turnover has been measured. The collapsing scheme is given by
-\code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
+\code{sbi * size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
 classification by respectively the first two SBI digits and the first SBI
@@
 a <- cumulate(producers
-             , collapse = sbi*size ~ sbi + sbi2 + sbi1
+             , collapse = sbi * size ~ sbi + sbi2 + sbi1
              , test = min_complete(n = 10, vars = "industrial")
@@
 r <- cumulate(producers
-         , collapse = sbi*size ~ sbi + sbi2 + sbi1
+         , collapse = sbi * size ~ sbi + sbi2 + sbi1
          , test = min_complete(n = 10, vars = "industrial")
@@
 dat <- merge(producers, r)
-dat <- impute_proxy(dat, industrial ~ R*total)
+dat <- impute_proxy(dat, industrial ~ R * total)
 @
@@
 imputations <- cumulate(producers
-                 , collapse = id ~ sbi*size + sbi + sbi2 + sbi1
+                 , collapse = id ~ sbi * size + sbi + sbi2 + sbi1
                  , test = min_complete(5,"trade")
@@
 r <- cumulate(producers
-         , collapse = sbi*size ~ sbi + sbi2 + sbi1
+         , collapse = sbi * size ~ sbi + sbi2 + sbi1
          , test = min_complete(n = 10, vars = c("total","industrial"))

### 2024-03-27 paper/article.Rnw
processed minor comments of Reviewer 1

@@
 objects. The more recent \citet{polars2023} library for \proglang{Python} and
-\proglang{Rust} also implements such features. Similarly, the \pkg{DataFrames}
+\proglang{Rust} also implement such features. Similarly, the \pkg{DataFrames}
 package for \proglang{julia} implements split-apply-combine functionality
@@
 could drop size class and compute the average ratio over all records within a
-certain economic activity. Or, one could choose to course-grain economic
+certain economic activity. Or, one could choose to coarse-grain economic
 activities by collapsing groups with activities that are deemed similar enough.
@@
 $B$ is a hierarchical classification with levels $\{11,12,13,21,22\}$. Variable
-$B_1$ is a course graining of $B$: for each record the value for $B_1$ is the
+$B_1$ is a coarse-graining of $B$: for each record the value for $B_1$ is the
 first digit of $B$. Hence, $B_1$ has levels $\{1,2\}$.
@@
 The \code{test} parameter of \code{cumulate()} and \code{accumulate()} accepts a
-function that takes a subset of the data and must returns a boolean. For common
+function that takes a subset of the data and returns a boolean. For common
 test conditions, including requiring a minimal number of records, or a minimal

### 2024-03-27 paper/article.Rnw
changed verbatim to Code environment

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5%%%%%%%
-\section[R package accumulate]{\proglang{R Package} \pkg{accumulate} }\label{sect:accumulate}
+\section[R Package accumulate]{\proglang{R Package} \pkg{accumulate} }\label{sect:accumulate}
 Grouped aggregation with a fall-though scenario based on a collapsing scheme
@@
 that offer slightly different interfaces.
-\begin{verbatim}
+\begin{Code}
   accumulate(data, collapse, test, fun, ...)
   cumulate(data, collapse, test, ...)
-\end{verbatim}
+\end{Code}
 Here \texttt{data} is a data frame holding data to be aggregated;
@@
 data frame and can schematically be represented as follows. 
-\begin{verbatim}
+\begin{Code}
   [Grouping Variables, Collapse level, Output aggregates]
-\end{verbatim}
+\end{Code}
 The first columns represent the variables that define the output grouping, the
@@
 input <- data.frame(
-     A  = c( 1,  1,  1,  2,  2,  2,  3,  3,  3)
-   , B  = c(11, 11, 11, 12, 12, 13, 21, 22, 12)
-   , B1 = c( 1,  1,  1,  1,  1,  1,  2,  2,  1)
-   , Y  = 1:9
+  A  = c( 1,  1,  1,  2,  2,  2,  3,  3,  3),
+  B  = c(11, 11, 11, 12, 12, 13, 21, 22, 12),
+  B1 = c( 1,  1,  1,  1,  1,  1,  2,  2,  1),
+  Y  = 1:9
 )
@@
 Summarizing, the \class{formula} interface is always of the following form.
-\begin{verbatim}
+\begin{Code}
   Target grouping ~ Alternative1 + Alternative2 + ... + AlternativeN
-\end{verbatim}
+\end{Code}
 
@@
 function (we skip arguments that are not important for the discussion).
-\begin{verbatim}
+\begin{Code}
   aggregate(x, by, FUN)
-\end{verbatim}
+\end{Code}
 Here, \code{x} is a data frame where each row represents an element of $U$. The

### 2024-03-27 paper/article.Rnw
updated indentation to JSS style

@@
 input$Y2 <- 11:19
-accumulate(input, collapse = A * B ~ A * B1 + A
-          , test = min_records(3), fun = mean)
+accumulate(input, collapse = A * B ~ A * B1 + A, 
+  test = min_records(3), fun = mean)
 @
@@
 <<>>=
-accumulate(input[-3], collapse = A * B ~ A
-          , test = min_records(3), fun = mean)
+accumulate(input[-3], collapse = A * B ~ A,
+  test = min_records(3), fun = mean)
 @
@@
 csh <- data.frame(
-  AB  = c("1-11", "2-12", "2-13", "3-21", "3-22", "3-12")
-, AB1 = c("1-1" , "2-1" , "2-1" , "3-2" , "3-2" , "3-1" )
-, A   = c("1"   , "2"   , "2"   , "3"   , "3"   , "3"   ))
+  AB  = c("1-11", "2-12", "2-13", "3-21", "3-22", "3-12"),
+  AB1 = c("1-1" , "2-1" , "2-1" , "3-2" , "3-2" , "3-1" ),
+  A   = c("1"   , "2"   , "2"   , "3"   , "3"   , "3"   ))
 csh
@@
 accumulate(input1, collapse = csh, test = min_records(3), mean)
-cumulate(input1, collapse = csh, test = min_records(3)
-        , muY = mean(Y), muY2 = mean(Y2))
+cumulate(input1, collapse = csh, test = min_records(3),
+  muY = mean(Y), muY2 = mean(Y2))
 @
@@
 my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >= 3
-accumulate(input, collapse = A * B ~ A * B1 + B1, test = my_test, fun = mean)
+accumulate(input, collapse = A * B ~ A * B1 + B1,
+  test = my_test, fun = mean)
 @
@@
 <<>>=
-nace <- c("0111", "0112", "0113", "0121"
-         , "0121", "0122", "0123", "0124")
+nace <- c("0111", "0112", "0113", "0121", "0121", "0122", "0123", "0124")
 csh_from_digits(nace, levels = 2)
@@
 <<>>=
-a <- cumulate(producers
-             , collapse = sbi * size ~ sbi + sbi2 + sbi1
-             , test = min_complete(n = 10, vars = "industrial")
-             , mean_industrial = mean(industrial, na.rm = TRUE))
+a <- cumulate(producers,
+  collapse = sbi * size ~ sbi + sbi2 + sbi1,
+  test = min_complete(n = 10, vars = "industrial"),
+  mean_industrial = mean(industrial, na.rm = TRUE))
 head(a,3)
@@
 <<>>=
-r <- cumulate(producers
-         , collapse = sbi * size ~ sbi + sbi2 + sbi1
-         , test = min_complete(n = 10, vars = "industrial")
-         , R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
+r <- cumulate(producers,
+  collapse = sbi * size ~ sbi + sbi2 + sbi1,
+  test = min_complete(n = 10, vars = "industrial"),
+  R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
 head(r,3)
@@
 <<>>=
-r <- cumulate(producers
-         , collapse = sbi * size ~ sbi + sbi2 + sbi1
-         , test = min_complete(n = 10, vars = c("total","industrial"))
-         , model = lm(total ~ industrial))
+r <- cumulate(producers,
+  collapse = sbi * size ~ sbi + sbi2 + sbi1,
+  test = min_complete(n = 10, vars = c("total","industrial")),
+  model = lm(total ~ industrial))
 head(r,3)

### 2024-03-27 paper/article.Rnw
tables on page 2 are now in a figure environment

@@
 \usepackage{tikz}
-
+\usepackage{float}
 % shortcut for surjection arrow, and surjection version of xrightarrow
@@
 \section{Dynamic Grouping} \label{sect:dynamicgrouping}
-In this section the illustrate the concept of dynamic grouping with a minimal
+In this Section the illustrate the concept of dynamic grouping with a minimal
 worked example. This example is not very realistic, but it is constructed
@@
 
-The following tables illustrate the idea. The left table represents the data
-set to be aggregated by $A$ and $B$.  The table on the right represents the
-output. Colors indicate which data was used. 
+The tables in Figure~\ref{fig:example} illustrate the idea. The left table
+represents the data set to be aggregated by $A$ and $B$.  The table on the
+right represents the output. Colors indicate which data was used. 
 %
-\begin{center}
+\begin{figure}[H]
+\centering
 \begin{tabular}{cccl}
@@
 \end{tabular}\hspace{1cm}\begin{tabular}{ccl}
-\multicolumn{3}{l}{Output}\\
+\multicolumn{3}{l}{Output aggregates}\\
 \hline
@@
 \end{tabular}
-\label{eq:example}
-\end{center}
+\label{fig:example}
+\caption{Input data (left) with grouping variables $A$, $B$ and $B_1$, and
+means of $Y$ per $A\times B$ after dynamic grouping (right).}
+\end{figure}
 %
@@
 \section{Extensive Example: Economic Data} \label{sect:example}
-In this section we discuss three practical examples using a synthetic
+In this Section we discuss three practical examples using a synthetic
 dataset included with the package.
@@
 \section{Formal Description and Algorithms} \label{sect:algorithm}
-In this section we give a formal description of the algorithm for aggregation
+In this Section we give a formal description of the algorithm for aggregation
 with dynamic grouping.  We start by giving an algorithm for ordinary

### 2024-03-27 paper/article.Rnw
Equation numbering, and abbreviations according to JSS style

@@
 \begin{tabular}{cccl}
-\multicolumn{4}{l}{Input data}\\
+\multicolumn{4}{l}{Input Data}\\
 \hline
@@
 \end{tabular}\hspace{1cm}\begin{tabular}{ccl}
-\multicolumn{3}{l}{Output aggregates}\\
+\multicolumn{3}{l}{Output Aggregates}\\
 \hline
-$A\times B$ & \code{level} & $\mu_Y$\\
+$A\times B$ & \code{Level} & $\mu_Y$\\
 \hline
@@
 They represent a classification of entities into non-overlapping, nested
-groupings. Examples include the International Standard Industrial Classification
-of Economic Activities (ISIC, \citet{un2022isic}), the related 
-Statistical Classification of Economic Activities in Europe (NACE, \citet{eu2006nace})
-and the Euroean Skills, Competences, Qualifications and Occupations classification
+groupings. Examples include the international standard industrial classification
+of economic activities (ISIC, \citet{un2022isic}), the related 
+statistical classification of economic activities in europe (NACE, \citet{eu2006nace})
+and the Euroean skills, competences, qualifications and occupations classification
 (ESCO, \citet{eu2022esco}).

### 2024-03-27 paper/article.Rnw
fixed formatting of programming languages in a few occasions

@@
 on missing data\footnote{https://cran.r-project.org/web/views/MissingData.html}
-currently lists 203 \pkg{R} packages that support some form of estimating
+currently lists 203 \proglang{R} packages that support some form of estimating
 missing data.  The \pkg{simputation} package \citep{loo2022simputation} seems
@@
 split-apply-combine aggregation where groups can be collapsed dynamically in
-the form of \pkg{R} package \pkg{accumulate} \citep{loo2022accumulate}.
+the form of \proglang{R} package \pkg{accumulate} \citep{loo2022accumulate}.
 

### 2024-05-17 paper/jss5097.Rnw
.

--- 2024-03-27 paper/article.Rnw
@@
 \author{Mark P.J. van der Loo~\orcidlink{0000-0002-9807-4686}\\
-        Statistics Netherlands and University of Leiden}
+        Statistics Netherlands and Leiden University}
 
@@
 %% - Should be comma-separated and in sentence case.
-\Keywords{data analysis, \proglang{R}}
+\Keywords{data analysis, estimation, aggregation, \proglang{R}}
 \Plainkeywords{data analysis, R}
@@
   Leiden Institute of Advanced Computer Science (LIACS)\\
-  University of Leiden\\
+  Leiden University\\
   P.O. Box 9512\\
@@
 \newpage
-\bibliography{refs}
+\bibliography{jss5097}
 
